---
title: "Biophysical models of green anole body temperatures"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

## Overview

We hypothesize that the green anole's poleward range edge is set not by the cold air temperatures in the canopy, but by the thermal conditions within the leaf litter where lizards overwinter. To test this, we first need to know the body temperatures a green anole would experience in different microclimates across southeastern North America. In this vignette, we:

1. Define the green anole's range using curated iNaturalist observations.

2. Obtain spatially explicit data on forest cover, elevation, and minimum and maximum temperatures across southeastern North America.

3. Use biophysical models to transform these macroclimatic data into the minimum annual body (i.e., operative) temperatures a green anole would experience in the canopy and at different depths in the leaf litter, and validate these estimates with data from our operative temperature models (OTMs).

We then use these estimates to evaluate which microclimate best predicts the green anole's range in the [mechanistic envelope models](envelope-models.qmd) vignette.

```{r r-setup, warning = FALSE, message = FALSE}
#| code-summary: "R setup"

# packages
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(TrenchR)
library(kableExtra)
library(viridis)
library(ggpubr)
library(geosphere)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)
```

## Defining the green anole's range

We downloaded all research-quality iNaturalist observations of the green anole (*Anolis carolinensis*). Below we present the structure of the raw data, which is stored in the `data` folder:

```{r present-structure-inat-obs-data}
#| code-fold: false

# load occurrence data
caro_inat_data <- readRDS("data/caro_inat_data.rds")
as_tibble(caro_inat_data)
str(caro_inat_data)
```

We use the occurence data to define the green anole's range. To do this, we first curate the original dataset by:

1. Rounding coordinates to one decimal place.

2. Removing observations with low positional accuracy (positional uncertainty > 1 km).

3. Filtering to the native range of the green anole defined by the following bounding box: longitude between -110 and -70 W, latitude between 22.5 and 37.5 N.

4. Adding up all observations within every 0.1 x 0.1 degree grid cell and keeping only cells with at least 3 observations. The goal of this step is to eliminate stand-alone observations caused by potential errors or rare occurrences.

5. Removing outliers based on distance to other observations. Any observation is considered an outlier if there are fewer than 3 observations within 50 km.

Lastly, we use the curated occurrence data to manually define the green anole's range using the `mapedit` package.

```{r define-caro-range, eval = FALSE}
#| code-summary: "Curate iNat data and define green anole range"

## Note: This code chunk is not evaluated as it requires manual interaction
## to define the green anole's range using the `mapedit` package. The resulting
## range points are saved as an RDS file for later use.

# load packages
library(mapview)
library(mapedit)

# round coordinates and remove observations with low positional accuracy
caro_occ_data <- caro_inat_data |>
  mutate(
    lat = round(latitude, 1), lon = round(longitude, 1),
  ) |>
  filter(
    !is.na(positional_accuracy) & positional_accuracy <= 1000
  ) |>
  group_by(lon, lat) |>
  summarise(n = n(), .groups = "drop") |>
  filter(n > 2) |>
  select(lon, lat)

# filter data to native range of green anole
caro_occ_data <- caro_occ_data |>
  filter(
    lon > -110 & lon < -70, lat > 22.5 & lat < 37.5
  )

# remove outliers based on distance to other observations
caro_occ_data$is_outlier <- apply(
  caro_occ_data[, c("lon", "lat")], 1, function(x) {
    distances <- geosphere::distm(
      x,
      caro_occ_data[, c("lon", "lat")],
      fun = geosphere::distHaversine
    )
    # if there are fewer than 3 observations within 50 km, mark as outlier
    if (length(distances) > 0 && sum(distances < 50000) < 3) {
      1
    } else {
      0
    }
  }
)

# save curated occurrence data
saveRDS(caro_occ_data, "data/caro_occ_data.rds")

# transform into a spatial object
caro_occ_sf <- st_as_sf(
  caro_occ_data[caro_occ_data$is_outlier == 0, ],
  coords = c("lon", "lat"),
  crs = 4326
)

# manually draw map around range
mapviewOptions(fgb = FALSE)
caro_range_sf <- editMap(mapview(caro_occ_sf))

# transform into a polygon object
caro_range_sf <- st_as_sf(caro_range_sf$finished)

# filter data to only include points within the continental US
caro_range_sf <- st_intersection(
  caro_range_sf,
  ne_countries(scale = "medium", returnclass = "sf") |>
    filter(admin %in% c("United States of America", "Mexico")) |>
    st_crop(xmin = -125, xmax = -65, ymin = 20, ymax = 50)
)

# change the spatial resolution of the polygon to 0.04166667 degrees (~5 km)
caro_range_sf <- st_transform(caro_range_sf, crs = 4326)

# combine into a single (multi)polygon suitable for masking other layers
caro_range_poly <- st_as_sf(caro_range_sf) |>
  st_union() |>
  st_make_valid() |>
  st_as_sf()

# save caro range
saveRDS(caro_range_poly, "data/caro_range.rds")
```

Below we plot the range in relationship to the curated occurance data indicating which observations we consider to be outliers. Both the curated occurance data (`caro_occ_data.rds`) and the defined range (`caro_range.rds`, saved as a `MULTIPOLYGON` object) are saved in the `data` folder.

```{r plot-range}
#| code-summary: "Plot the native range and occurrence data of the green anole"
#| fig-align: "center"
#| fig-width: 5

# load datasets
caro_range <- readRDS("data/caro_range.rds")
caro_occ_data <- readRDS("data/caro_occ_data.rds")

# plot
ggplot() +
  geom_polygon(
    data = map_data("world"),
    aes(x = long, y = lat, group = group),
    fill = "lightgray", color = "darkgray"
  ) +
  geom_polygon(
    data = map_data("state"),
    aes(x = long, y = lat, group = group),
    fill = NA, color = "gray"
  ) +
  geom_sf(
    data = caro_range,
    fill = "forestgreen", color = NA, alpha = 0.25
  ) +
  geom_point(
    data = caro_occ_data,
    aes(x = lon, y = lat, color = as.factor(is_outlier)),
    size = 1, alpha = 0.75
  ) +
  scale_color_manual(
    values = c("darkgreen", "darkred"),
    labels = c("In Range", "Outlier")
  ) +
  coord_sf(xlim = c(-101, -75), ylim = c(24, 37.5)) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(-100, -75, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(25, 37.5, by = 2.5),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank(),
    plot.margin = margin(r = 10, t = -20, b = -20, l = 10)
  ) +
  guides(color = guide_legend(override.aes = list(size = 3)))
```

## Obtaining and processing ecological data

Next, we obtain the layers of ecological data we will need for the biophysical models. These include:

1. **Forest cover**: A raster layer representing the proportion of land covered by forests in each spatial cell. We obtained this data using the [`landcover`](https://rspatial.github.io/geodata/reference/landcover.html) function of the [`geodata`](https://rspatial.github.io/geodata/index.html) package and combined the outputs of running this function setting `var` to `"trees"`, `"mangroves"`, and `"wetlands"`. This data is, by default, offered at a 30 arc-second resolution. As part of our processing, we aggregated it to a 2.5 arc-minute resolution to match the resolution of the climate data.

2. **Elevation**: A raster layer representing the elevation of each spatial cell. We obtained this data from the [WorldClim](https://www.worldclim.org/data/worldclim21.html) database at a 2.5 arc-minute resolution.

3. **Minimum and maximum temperatures**: Raster layers representing the minimum and maximum temperatures for each spatial cell. We obtained this data from the [WorldClim](https://www.worldclim.org/data/worldclim21.html) database at a 2.5 arc-minute resolution.

We `masked` all ecological data to the spatial extent of the green anole's range polygon + 2 degrees in all directions to have a representative sample of areas where the species does and does not occur. Further, we `masked` the rasters to only land areas within the continental United States and Mexico, the two countries were green anoles are considerd to be native. 

Lastly, we add the species presence data to the ecological layers and save the resulting dataset (`eco_df`) as an RDS file for later use. The file is available in the `data` folder.

```{r process-ecological-data, eval = FALSE}
#| code-summary: "Obtaining and processing ecological data"

## Note: This code chunk is not evaluated as it requires large raster files
## that are not included in the repository. The output of this code chunk is
## saved as an RDS file for later use.

# define land polygon
land_edge <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(admin %in% c("United States of America", "Mexico")) |>
  st_transform(crs = 4326)

# crop the land polygon to the area surrounding the green anole range
range_extent <- terra::ext(
  st_bbox(caro_range_poly)["xmin"] - 2,
  st_bbox(caro_range_poly)["xmax"] + 2,
  st_bbox(caro_range_poly)["ymin"] - 2,
  st_bbox(caro_range_poly)["ymax"] + 2
)

# crop land edge
land_edge <- land_edge |>
  terra::vect() |>
  terra::crop(range_extent) |>
  st_as_sf()

# read and process vegetation rasters
trees <- terra::rast("raw_data/trees.tif")
mangroves <- terra::rast("raw_data/mangroves.tif")
wetlands <- terra::rast("raw_data/wetlands.tif")
veg <- trees + mangroves + wetlands
veg <- terra::mask(veg, land_edge)
veg <- terra::crop(veg, range_extent)
# 30'' to 2.5'
veg <- terra::aggregate(veg, fact = 5, fun = "mean", na.rm = TRUE)

# read and process elevation raster
elev <- terra::rast("raw_data/elev.tif")
elev <- terra::mask(elev, land_edge)
elev <- terra::crop(elev, range_extent)

# read and process tmin and tmax rasters
tmin <- terra::rast("raw_data/tmin.tif")
tmin <- terra::mask(tmin, land_edge)
tmin <- terra::crop(tmin, range_extent)
tmax <- terra::rast("raw_data/tmax.tif")
tmax <- terra::mask(tmax, land_edge)
tmax <- terra::crop(tmax, range_extent)

# match resultion of all raster
elev <- terra::resample(elev, veg, method = "bilinear")
tmin <- terra::resample(tmin, veg, method = "bilinear")
tmax <- terra::resample(tmax, veg, method = "bilinear")

# create a raster for green anole range
range_rast <- terra::rasterize(
  terra::vect(caro_range), veg, field = 1, background = 0
)
range_rast <- terra::mask(range_rast, land_edge)
range_rast <- terra::resample(range_rast, veg, method = "bilinear")

# combine all rasters into a single SpatRaster object
eco_rast <- c(veg, elev, tmin, tmax, range_rast)

# transform into data frame
eco_df <- as.data.frame(eco_rast, xy = TRUE, na.rm = TRUE)
colnames(eco_df) <- c(
  "lon", "lat", "veg", "elev", "tmin", "tmax", "presence"
)
eco_df$presence <- ifelse(eco_df$presence > 0, 1, 0)

# save final ecological data
saveRDS(eco_df, "data/eco_df.rds")

```

Below we plot the ecological data for the green anole's range, including forest cover, minimum and maximum temperatures, elevation, and species presence.

```{r plot-eco-data}
#| code-summary: "Plotting ecological data for the green anole's range"
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10

eco_df <- readRDS("data/eco_df.rds")

# define common theme and axis labels for all plots
common_theme <- theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank()
  )
x_axis_common <- scale_x_continuous(
  breaks = seq(-100, -75, by = 5),
  labels = function(x) paste0(abs(x), "°W"),
  expand = c(0, 0)
)
y_axis_common <- scale_y_continuous(
  breaks = seq(25, 37.5, by = 2.5),
  labels = function(x) paste0(x, "°N"),
  expand = c(0, 0)
)

# generate all plots
veg_plot <- ggplot(eco_df, aes(x = lon, y = lat, fill = veg)) +
  geom_raster() +
  scale_fill_gradient(
    low = "lightyellow", high = "darkgreen", na.value = NA
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 37.5)) +
  common_theme + x_axis_common + y_axis_common +
  labs(fill = "Forest\nCover")
tmin_plot <- ggplot(eco_df, aes(x = lon, y = lat, fill = tmin)) +
  geom_raster() +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 37.5)) +
  common_theme + x_axis_common + y_axis_common +
  labs(fill = "Minimum\nTemp. (°C)")
tmax_plot <- ggplot(eco_df, aes(x = lon, y = lat, fill = tmax)) +
  geom_raster() +
  scale_fill_viridis(
    option = "plasma", na.value = NA
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 37.5)) +
  common_theme + x_axis_common + y_axis_common +
  labs(fill = "Maximum\nTemp. (°C)")
elev_plot <- ggplot(eco_df, aes(x = lon, y = lat, fill = elev)) +
  geom_raster() +
  scale_fill_viridis() +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 37.5)) +
  common_theme + x_axis_common + y_axis_common +
  labs(fill = "Elevation\n(m)")

# combine all plots into a single figure
ggarrange(
  veg_plot, tmin_plot, tmax_plot, elev_plot,
  ncol = 2, nrow = 2,
  align = "hv"
)

```

## Biophysical models

### Functions & parameters

#### **Solar radiation profile**

We estimated direct hourly solar radiation ($R_{solar}$) as a function of geographic coordinates, seasonal timing and local topography. We first estimated the position of the sun throughout the year cycle (i.e., its ecliptic longitude $\lambda$) as a function of day of the year ($DOY$) and the Earth’s orbital eccentricity ($e = 0.0167$) as:
$$
\lambda = \frac{2\pi}{365}\cdot(DOY - 80) + 2 \cdot e \cdot (\sin(\frac{2\pi}{365}\cdot DOY) - \sin(\frac{2\pi}{365}\cdot 80))
$$

Above, 80 is chosen to represent the day of the year when the vernal equinox occurs (typically March 21st). Using the ecliptic longitude, we then calculated the solar declination angle ($\delta$), which dictates day length and solar intensity as.

$$
\delta = \arcsin(0.3879672 \cdot \sin(\lambda))
$$

In the expression above, $0.3879672$ is chosen to represent the Earth’s approximate average axial tilt throughout a year (arcsin(0.3879672) = 22.82$^\circ$). From this we calculate the sine of the solar altitude angle at a given hour of the day ($a(h)$) as a function of the latitude ($lat$) and the difference between the hour of the day ($h$) and the time of solar noon ($S_n$).

$$
a(h) = \sin(lat)\cdot \sin(\delta) + \cos(lat)\cdot\cos(\delta) \cdot \cos\left(0.2618 \cdot (h - S_n)\right)
$$

Where $0.2618 \text{ rad h}^{-1}$ represents the Earth’s rotational speed ($15^\circ h^{-1}$). To obtain the time of solar noon at a given location in a given day of the year we used the `solar_noon` function from the `R` package `TrenchR` (Buckley et al. 2023).
To account for the effects of elevation when computing the degree to which the atmosphere absorbs incoming solar radiation, we modeled barometric pressure ($p_a$) using an exponential attitudinal decay function.

$$
p_a = 101.3 \cdot e^{-Elev /8200}
$$

Combining $\text{sin}(a)$ and $p_a$ we then compute the path length of solar radiation through the atmosphere as:

$$
m_a(h) = \frac{p_a}{101.3 \cdot (\frac{1}{a_h})}
$$

Finally, hourly direct solar radiation ($R_{solar}$, in $\text{W m}^{-2}$) is modeled by combining $\text{sin}(\alpha)$, $m_a$, the extraterrestrial solar constant ($E_{sc} = 1.36 \text{kW m}^-2$) with the atmospheric transmittance coefficient ($\tau = 0.83$).

$$
R_{solar}(h) = E_{sc} \cdot \tau^{m_a(h)} \cdot a(h) \cdot 1000
$$

We implement the function in `R` as follows:

```{r define-solar-radiation-profile-function}
#| code-fold: false

# define constants
ecc <- 0.0167 # Earth's orbital eccentricity
spo <- 1.36 # extraterrestrial solar constant in kW m^-2
atc <- 0.83 # atmospheric transmittance coefficient

# define function
get_solar_rad_profile <- function(lat, elev, doy, solar_noon) {

  hours <- 0:23

  # calculate solar longitude and declination angle
  elon <- (2 * pi / 365) * (doy - 80) + 2 *
    ecc * (sin(2 * pi / 365 * doy) - sin(2 * pi / 365 * 80))
  dec_ang <- asin(0.3879672 * sin(elon))

  # calculate atmospheric pressure at the given elevation
  p_a <- 101.3 * exp(-elev / 8200)

  # calculate solar altitude angle
  geo_lat <- lat * (pi / 180)
  sin_alt_ang <- sin(geo_lat) * sin(dec_ang) +
    cos(geo_lat) * cos(dec_ang) *
      outer(solar_noon, hours, function(s, h) cos(0.2618 * (h - s)))

  # calculate air mass and solar radiation
  m_a <- p_a / (101.3 * (1 / sin_alt_ang))
  solar_rad <- spo * atc ^ m_a * sin_alt_ang
  solar_rad[solar_rad < 0] <- 0
  drop(solar_rad * 1000)

}

```

#### **Air temperature profile**

We estimated hourly air temperature ($T_{air}$) by interpolating between the daily maximum ($T_{max}$) and minimum ($T_{min}$) temperatures using a sinusoidal weighting function. We first defined the angular frequency of the diurnal temperature cycle ($\omega$) as:

$$
\omega = \frac{\pi}{12}
$$

reflecting a full cycle over the 24 hours of the day. From this we computed an hourly weighting coefficient ($\gamma(h)$), which determines the relative contribution of $T_{max}$ and $T_{min}$ to air temperature at a given hour of the day ($h$):

$$
\gamma(h) = 0.44 - 0.46\cdot\sin(0.9 + \omega \cdot h) + 0.11\cdot\sin(0.9 + 2 \cdot \omega \cdot h)
$$

Finally, hourly air temperature ($T_{air}(h)$, in $^\circ C$) was calculated by weighting $T_{max}$ and $T_{min}$ by $\gamma(h)$ and $1 - \gamma(h)$, respectively:

$$
T_{air}(h) = T_{max} \cdot \gamma(h) + T_{min} \cdot (1 - \gamma(h))
$$

We implement the function in `R` as follows:

```{r define-air-temperature-profile-function}
#| code-fold: false
get_air_temp_profile <- function(tmax, tmin) {
  w <- pi / 12
  gamma <- 0.44 - 0.46 * sin(0.9 + w * (0:23)) + 0.11 * sin(0.9 + 2 * w * (0:23)) # nolint
  drop(outer(tmax, gamma) + outer(tmin, 1 - gamma))
}
```

#### **Correct air temperature profile using vegetation cover**

We corrected the hourly air temperature profile ($T_{air}$) to account for the buffering effect of vegetation cover on incoming solar radiation. We assumed that forest canopy intercepts a portion of direct solar radiation ($R_{solar}$) before it reaches the ground, cooling the air in proportion to both the amount of solar radiation available to be intercepted and the fraction of forest cover ($C_{forest}$, ranging from 0 to 1) at a given location. We modeled this using a canopy cooling coefficient ($\kappa = 0.007 \text{°C}\cdot(\text{W m}^{-2})^{-1}$), such that the corrected hourly air temperature ($T_{air}^{\prime}(h)$) is:

$$
T_{air}^{\prime}(h) = T_{air}(h) - \kappa \cdot R_{solar}(h) \cdot C_{forest}
$$

Under peak solar radiation (800 - 1000 W m<sup>-2</sup>) and a fully closed canopy, this coefficient yields a peak daytime cooling buffer of 5.6 - 7 $^\circ$C. This range is consistent with empirical understory microclimate logger networks in temperate and subtropical forests (De Frenne et al., 2013, 2019; Zellweger et al., 2020) and standard surface-layer energy balance assumptions under suppressed convective mixing (Campbell & Norman, 1998).

We implement the function in `R` as follows:

```{r define-correct-air-temp-profile-function}
#| code-fold: false

# define parameters
cc <- 0.007 # canopy cooling coefficient (W m^-2)

# define function
correct_air_temp_profile <- function(
  air_temp_profile, solar_rad_profile, forest_cover
) {
  is_vec <- is.vector(air_temp_profile)
  if (is_vec) {
    air_temp_profile  <- matrix(air_temp_profile,  nrow = 1)
    solar_rad_profile <- matrix(solar_rad_profile, nrow = 1)
  }
  result <- air_temp_profile - 
    sweep(cc * solar_rad_profile, 1, forest_cover, "*")
  if (is_vec) drop(result) else result
}
```

`air_temp_profile` and `solar_rad_profile` may each represent either a single day (a vector of 24 hourly values) or multiple locations/days at once (a matrix with one row per location/day and one column per hour). The function first coerces vector inputs into single-row matrices so that the remainder of the computation can be written generically for both cases, then `drop()`s the result back to a vector at the end if needed. `forest_cover` varies by location (i.e., by row) but must be applied uniformly across all 24 hourly columns of `solar_rad_profile`. Because R recycles values column-by-column, multiplying the `solar_rad_profile` matrix directly by the `forest_cover` vector would misalign values whenever there is more than one row. `sweep()` avoids this by explicitly multiplying along `MARGIN = 1` (rows), ensuring that each location's forest cover value is applied only to its own row of hourly solar radiation values.

#### **Surface temperature profile**

We estimated the hourly surface temperature profile ($T_{surf}$) by adjusting hourly corrected air temperature ($T_{air}^{\prime}$) upward to reflect the additional heating that exposed ground surfaces experience from direct solar radiation ($R_{solar}$), relative to the surrounding air. We assumed that this surface heating effect is proportional to the fraction of ground not shaded by forest cover ($C_{veg}$, ranging from 0 to 1), such that fully covered surfaces experience no additional heating while fully exposed surfaces experience the maximum effect. We modeled this using a surface radiative heat flux coefficient ($\sigma = 0.02 \text{°C}\cdot(\text{W m}^{-2})^{-1}$), such that hourly surface temperature ($T_{surf}(h)$) is:

$$
T_{surf}(h) = T_{air}^{\prime}(h) + \sigma \cdot R_{solar}(h) \cdot (1 - C_{veg})
$$

The surface radiative heat flux coefficient ($\sigma = 0.02 \text{ °C}\cdot(\text{W m}^{-2})^{-1}$) reflects the thermal response of dry organic leaf litter under low boundary-layer convective mixing. Under maximum unshaded solar radiation (800 - 1000 W m<sup>-2</sup>), this yields a daytime surface temperature excess of 16 - 20 $^\circ$C above ambient air, consistent with standard microclimate energy balance formulations (Campbell & Norman, 1998) and empirical forest floor microclimate measurements (De Frenne et al., 2019; Maclean et al., 2019).

We implement the function in `R` as follows:

```{r define-surface-temperature-profile-section}
#| code-fold: false

# define parameters
srh <- 0.02 # surface radiative heat flux coefficient

# surface temperature profile
get_surf_temp_profile <- function(
  air_temp_profile, solar_rad_profile, veg_cover
) {
  is_vec <- is.vector(air_temp_profile)
  if (is_vec) {
    air_temp_profile  <- matrix(air_temp_profile,  nrow = 1)
    solar_rad_profile <- matrix(solar_rad_profile, nrow = 1)
  }
  result <- air_temp_profile + 
    sweep(srh * solar_rad_profile, 1, 1 - veg_cover, "*")
  if (is_vec) drop(result) else result
}

```

#### **Above- and Below-ground (leaf-litter) temperature profile**

To model the vertical temperature profile we defined separate functions for above and below the surface. Starting with above the surface, we assumed a smooth transition between the surface and the corrected air temperature measured at a reference height of 1.8 m (the reference height assumed by WorldClim). Following this assumption we first defined a dimensionless vertical profile function $\phi(z)$ that would capture the relative influence of the surface temperature with height ($z$), as: 

$$
\phi(z) = \frac{\ln\left(\frac{z + z_0}{z_0}\right)}{\ln\left(\frac{z_{ref} + z_0}{z_0}\right)}
$$

Above we assume $z_0 = 0.01$ m. This provides a small roughness-length offset added to avoid the singularity of the logarithm at the surface ($z = 0$). Then we, estimated the temperature at any height above the surface by applying this profile function as:

$$
T(z,h) = T_{surf}(h) - \left[T_{surf}(h) - T_{air}^{\prime}(h)\right] \cdot \phi(z)
$$

Below the surface, we modeled the propagation of temperature from the surface using the analytical solution to the one-dimensional heat conduction equation defined in Campbell and Norman 1998. To achieve this, we first calculated the thermal diffusivity of leaf litter ($\kappa$) from its thermal conductivity ($k$), bulk density ($\rho$), and specific heat capacity ($c_p$):

$$
\kappa = \frac{k}{\rho \cdot c_p}
$$

We parameterized $k$, $\rho$, and $c_p$ using literature values representative of dry, loosely-packed organic leaf litter  rather than mineral soil, since litter's high porosity ($\approx$ 80-90%) traps insulating pockets of still air that give it much lower thermal conductivity and bulk density than mineral soil. We set:

- $k = 0.05 \text{ W m}^{-1}\text{K}^{-1}$, within the $0.04\text{-}0.06 \text{ W m}^{-1}\text{K}^{-1}$ range reported for dry organic litter layers (Campbell and Norman 1998; Oke 1987).

- $\rho = 150 \text{ kg m}^{-3}$, representing a slightly settled field-layer density within the $60\text{--}180 \text{ kg m}^{-3}$ range measured for surficial $O_i$ litter (Sato et al. 2004; Gaudinski et al. 2000).

- $c_p = 2500 \text{ J kg}^{-1}\text{K}^{-1}$, the standard value used for the dry solid organic fraction of soils (De Vries 1963; Bonan 2019).

Given the angular frequency of the diurnal temperature cycle ($\omega = 2\pi/24 \text{ rad h}^{-1}$), we then calculated the damping depth ($D$), i.e. the depth at which the amplitude of the surface temperature wave is reduced by a factor of $e^{-1}$:

$$
D = \sqrt{\frac{2\kappa}{\omega}}
$$

At a given depth, temperature fluctuations are both damped in amplitude and delayed in phase relative to the surface. To account for this, we calculated the damping ratio ($A(z)$) and phase lag ($\Delta t(z)$, in hours) as:

$$
A(z) = e^{-|z|/D}
$$

$$
\Delta t(z) = \frac{|z|}{D} \cdot \frac{24}{2\pi}
$$

Finally, temperature at depth $z$ and hour $h$ ($T(z,h)$) was calculated by damping the phase-shifted surface temperature ($T_{surf}(h - \Delta t(z))$, obtained by linearly interpolating the surface temperature profile) around the mean daily surface temperature ($\bar{T}_{surf}$):

$$
T(z,h) = \bar{T}_{surf} + \left[T_{surf}(h - \Delta t(z)) - \bar{T}_{surf}\right] \cdot A(z)
$$

We implement the function in `R` as follows:

```{r define-below-ground-leaf-litter-profile-section}
#| code-fold: false

# heights and depths of interests
z_list <- zs <- c(seq(-0.2, -0.01, by = 0.01), 0, 1.8, 2)

# parameters
k <- 0.05 # thermal conductivity of leaf litter (W m^-1 K^-1)
rho <- 150 # bulk density of leaf litter (kg m^-3)
cp <- 2500 # specific heat capacity of leaf litter (J kg^-1 K^-1)
ref_height <- 1.8 # reference height (m)

# derived parameters
diff_sec <- k / (rho * cp) # thermal diffusivity of leaf litter (m^2 s^-1)
diff_hour <- diff_sec * 3600 # thermal diffusivity of leaf litter (m^2 h^-1)
omega <- 2 * pi / 24 # angular frequency of diurnal temperature cycle (rad h^-1)
damping_depth <- sqrt((2 * diff_hour) / omega) # damping depth (m)

# function to calculate temperature profile at different heights and depths
get_z_temp_profile <- function(z_list, surf_temp_profile, air_temp_profile) {

  # check if the input profiles are vectors or matrices
  is_vec <- is.vector(surf_temp_profile)
  if (is_vec) {
    surf_temp_profile <- matrix(surf_temp_profile, nrow = 1)
    air_temp_profile  <- matrix(air_temp_profile,  nrow = 1)
  }

  # define extended hours for interpolation
  hours <- 0:23
  surf_temp_mean <- rowMeans(surf_temp_profile)
  ext_hours <- c(hours - 24, hours, hours + 24)

  # calculate temperature profiles at different heights and depths
  z_temp_profiles <- lapply(z_list, function(z) {

    if (z == 0) {
      surf_temp_profile

    } else if (z == ref_height) {

      air_temp_profile

    } else if (z > 0) {

      a <- log((z + 0.01) / 0.01) / log((ref_height + 0.01) / 0.01)
      surf_temp_profile - (surf_temp_profile - air_temp_profile) * a

    } else {

      damp_ratio <- exp(-abs(z) / damping_depth)
      phase_lag <- (abs(z) / damping_depth) * (24 / (2 * pi))
      xout <- hours - phase_lag
      idx <- findInterval(xout, ext_hours)
      w <- (xout - ext_hours[idx]) / (ext_hours[idx + 1] - ext_hours[idx])
      ext_surf <- cbind(surf_temp_profile, surf_temp_profile, surf_temp_profile)
      shifted_temp <- sweep(ext_surf[, idx], 2, 1 - w, "*") +
        sweep(ext_surf[, idx + 1], 2, w, "*")
      surf_temp_mean + (shifted_temp - surf_temp_mean) * damp_ratio

    }
  })

  # combine the temperature profiles into a matrix
  names(z_temp_profiles) <- z_list
  if (is_vec) z_temp_profiles <- lapply(z_temp_profiles, drop)
  z_temp_profiles

}

```

The function above relies on several programming techniques to efficiently compute temperature at many heights/depths, locations and hours at once. `lapply()` iterates over `z_list`, evaluating a separate branch of the `if`/`else if`/`else` logic for each height/depth `z`, and returns the results as a list, avoiding an explicit `for` loop while keeping one profile (of potentially many rows/locations, by 24 columns/hours) per element. Because `get_z_temp_profile()` is a closure, the branches inside `lapply()` can freely access objects defined outside the function itself (e.g., `damping_depth`, `ref_height`) as well as arguments of the function itself (e.g., `z_list`) without having to pass them in explicitly. As with the earlier profile-correction functions, vector inputs are first coerced into single-row matrices so that the same matrix-based code handles both a single location/day and many at once, and `drop()` reverses this at the end when appropriate.

The belowground branch (`z < 0`) additionally needs to look up surface temperature at fractional, phase-shifted hours (e.g., `hour - phase_lag`) that can fall before hour `0` or after hour `23`. To do this without treating the daily cycle as artificially bounded, `hours` is tiled into `ext_hours`, a length-72 sequence spanning the day before, the day itself, and the day after (`hours - 24`, `hours`, `hours + 24`), and `surf_temp_profile` is likewise tiled column-wise into `ext_surf` via `cbind()`. `findInterval()` then locates, for each shifted hour in `xout`, the index of the nearest surrounding pair of columns in `ext_hours`, and linear interpolation weights (`w`) are computed from the relative distance to each. `sweep()` is used again here, this time along `MARGIN = 2` (columns), to apply a different interpolation weight to each of the 24 target hours while still allowing multiple rows (locations/days) to be processed simultaneously.

#### **Operative temperature (body temperature) transformation**

We estimated the minimum operative temperature ($T_e$) of a lizard (i.e., the steady-state body temperature reached at thermal equilibrium) across a range of above-ground and below-ground (leaf-litter) microhabitats. Because the coldest temperatures of a given day generally occur at night, this allowed us to assume that shortwave solar radiation was absent. Further, if the lizard is fully enclosed within the leaf litter, its view factor to the open sky drops to zero, all surrounding surfaces and boundary air are at ambient temperature, and radiative cooling to the sky is eliminated making $T_e = T(z)$. Instead, if the lizard is aboveground, we assumed that $T_e$  equilibrium, $T_e$ is a conductance-weighted average of minimum ambient temperature in a given microclimate ($T(z)$), ground temperature ($T_{ground}$), and an equivalent radiant temperature ($T_{rad}$):

$$
T_e = \frac{h_c \cdot T(z) + h_k \cdot T_{ground} + h_r \cdot T_{rad}}{h_c + h_k + h_r}
$$

Where $h_c$, $h_k$, and $h_r$ are, respectively, the convective, conductive, and radiative heat transfer coefficients. To solve for $T_e$ when exposed, we first scaled body mass ($m$, in kg) to a characteristic length ($L$) assuming standard reptile allometry:

$$
L = 0.02 \cdot m^{1/3}
$$

We then calculated the convective heat transfer coefficient ($h_c$) as the greater of forced convection (driven by wind speed, $u$) and free convection (driven by buoyancy):

$$
h_{c,forced} = \frac{10.3 \cdot u^{0.6}}{L^{0.4}} \qquad h_{c,free} = 3.5 \cdot L^{-0.25} \qquad h_c = \max(h_{c,forced},\ h_{c,free})
$$

We set wind speed to $u = 0.1 \text{ m s}^{-1}$, representing the near-still boundary layer typical of sheltered forest understories at night, which minimizes forced convection and thereby amplifies the relative impact of radiative cooling. We treated conduction to the ground as a fixed empirical coefficient, $h_k = 2.5 \text{ W m}^{-2}\text{K}^{-1}$, representing a small reptile resting against a solid substrate to maximize heat exchange.

Radiative heat transfer was linearized around the Stefan-Boltzmann constant ($\sigma = 5.670374\times10^{-8} \text{ W m}^{-2}\text{K}^{-4}$) and lizard skin emissivity ($\epsilon_{lizard} = 0.95$, reflecting near-blackbody behavior of biological tissue in the thermal infrared):

$$
h_r = 4 \cdot \epsilon_{lizard} \cdot \sigma \cdot T_{air,K}^3
$$

The equivalent radiant temperature ($T_{rad}$) integrates thermal radiation from the night sky and the ground, weighted by their respective view factors ($F_{sky}$ and $F_{ground}$, each set to $0.5$ assuming a flat, horizontal perching posture typical of a lizard when sleeping):

$$
T_{sky,K} = \epsilon_{sky}^{0.25} \cdot T_{air,K} \qquad T_{rad,K} = \left(F_{sky} \cdot T_{sky,K}^4 + F_{ground} \cdot T_{ground,K}^4\right)^{0.25}
$$

Where sky emissivity ($\epsilon_{sky} = 0.75$) represents a clear night sky, whose lack of cloud cover lets substantial longwave radiation escape to space, dropping the effective sky temperature well below ambient air temperature. 

While it likely that some of the assumptions we take here are simplifications of real-world conditions, they provide a useful framework for estimating the operative temperature of a small reptile in a forest understory at night.

We implement the function in `R` as follows:

```{r define-tb-function}
#| code-fold: false

# parameters
sigma <- 5.670374e-8  # Stefan-Boltzmann constant (W m^-2 K^-4)
lizard_emissivity <- 0.95 # Longwave emissivity of reptile skin
sky_emissivity <- 0.75 # Longwave emissivity of the night sky
wind_speed <- 0.1 # Wind speed in m/s
f_sky <- 0.5 # Sky view factor above ground (50%)
f_ground <- 0.5 # Ground view factor above ground (50%)
convection_ground <- 2.5 # Conductive/convective transfer from ground

# define function to calculate operative (body) temperature
get_body_temp <- function(ambient_t, mass, in_litter) {

  # calculate derived values
  ambient_t_k <- ground_t_k <- ambient_t + 273.15 # C to K

  # characteristic length of lizard (m))
  l <- 0.02 * (mass / 1000)^(1 / 3)

  # forced convection (W m^-2 K^-1)
  forced_convection <- 10.3 * (wind_speed^0.6) / (l^0.4)

  # baseline free convection (W m^-2 K^-1)
  free_convection <- 3.5 * (1 / l)^0.25

  # total convection (W m^-2 K^-1)
  total_convection <- pmax(forced_convection, free_convection)

  # radiative transfer coefficient (W m^-2 K^-1)
  rad_heat_transfer <- 4 * lizard_emissivity * sigma * (ambient_t_k^3)

   # night sky radiative temperature
  sky_t_k <- (sky_emissivity)^0.25 * ambient_t_k

  # equivalent radiant temperature (°C)
  radiant_t <- (f_sky * (sky_t_k^4) + f_ground * (ground_t_k^4))^0.25 - 273.15

  # operative temperature above ground
  a <- (
    total_convection * ambient_t +
      convection_ground * ambient_t +
      rad_heat_transfer * radiant_t
  )
  b <- total_convection + convection_ground + rad_heat_transfer
  op_t <- a / b

  # return above or below ground based on `in_litter` parameter
  ifelse(in_litter, ambient_t, op_t)

}

```

### Run biophysical models

Below we implement all the biophyscial modeling functions described above to estimate the minimum annual operative temperature of a green anole across 23 microclimates (20 leaf-litter depths from 1 to 20 cm, the surface, and heights of 1.8 and 2 m) in a typical year for every 2.5' latitude and longitude spatial cell. Because minimum annual temperatures in the northern hemisphere typically occur in January, we use January 15 (`doy = 15`) as the representative day for our calculations. 

```{r run-biophys-models, eval = FALSE}
#| code-summary: "Run biophysical models"

## Note: This code chunk is not evaluated as it takes a long time to run.
## The output is saved in the `data/micro_df.rds` file.

eco_df$solar_noon <- mapply(
  TrenchR::solar_noon,
  lon = eco_df$lon,
  doy = 15
)
air_temp_prof <- get_air_temp_profile(
  tmax = eco_df$tmax,
  tmin = eco_df$tmin
)
solar_rad_prof <- get_solar_rad_profile(
  lat = eco_df$lat,
  elev = eco_df$elev,
  doy = 15,
  solar_noon = eco_df$solar_noon
)
air_temp_prof_corrected <- correct_air_temp_profile(
  air_temp_profile = air_temp_prof,
  solar_rad_profile = solar_rad_prof,
  forest_cover = eco_df$veg
)
surf_temp_prof <- get_surf_temp_profile(
  air_temp_profile = air_temp_prof_corrected,
  solar_rad_profile = solar_rad_prof,
  veg_cover = eco_df$veg
)
zs_temp_profiles <- get_z_temp_profile(
  z_list = zs,
  surf_temp_profile = surf_temp_prof,
  air_temp_profile = air_temp_prof_corrected
)
# summarize zs temperature profiles into a data frame
micro_df <- Map(
  function(z_temp, z_val) {
    data.frame(
      lon = eco_df$lon,
      lat = eco_df$lat,
      elev = eco_df$elev,
      veg = eco_df$veg,
      presence = eco_df$presence,
      z  = z_val,
      min_temp = do.call(pmin, as.data.frame(z_temp)),
      hours_below_0C   = rowSums(z_temp < 0),
      hours_below_6.8C = rowSums(z_temp < 6.8)
    )
  },
  zs_temp_profiles, zs
) |>
  bind_rows() |>
  group_by(lon, lat, elev, veg, presence, z) |>
  summarise(
    min_annual_temp   = min(min_temp),
    hours_below_0C    = sum(hours_below_0C),
    hours_below_ctmin = sum(hours_below_6.8C),
    .groups = "drop"
  )
micro_df$min_annual_body_temp <- get_body_temp(
  in_litter = ifelse(micro_df$z < 0, TRUE, FALSE),
  ambient_t = micro_df$min_annual_temp,
  mass = 3.1 # typical mass of a green anole (g)
)

# save the microclimate data frame
saveRDS(micro_df, "data/micro_df.rds")

```

### Validation

We compare the minimum annual operative temperatures estimated from our biophysical models (`micro_df`) with the observed minimum temperatures from the operative temperature models (`otms`). This validation helps assess the accuracy of our microclimate and biophysical modeling approach.

```{r validate-biophys-model-results}
#| code-summary: "Validate biophysical model results"
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5
#| out-width: "66%"

micro_df <- readRDS("data/micro_df.rds")
otms <- readRDS("data/otms-data.rds")

# prepare OTM data for validation
otms_val <- otms |>
  filter(doy > 10 & doy < 20) |>
  group_by(location, doy, id, microhabitat) |>
  summarise(
    min_temp = min(temp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(location, id, microhabitat) |>
  summarise(
    mean_min_temp = mean(min_temp, na.rm = TRUE),
    .groups = "drop"
  )

# define locations of interest
locations <- tibble(
  location = c("MIA", "PSL", "ORL", "GNV", "ATL"),
  lon = c(-80.1918, -80.3256, -81.3792, -82.3248, -84.3880),
  lat = c(25.7617, 27.305, 28.5383, 29.6516, 33.7490)
)

# filter microclimate temps to only include locations only
# within 0.1 degrees of the defined locations
micro_df_loc <- micro_df |>
  filter(as.character(z) %in% c("-0.08", "1.8")) |>
  filter(
    purrr::pmap_lgl(
      list(lon, lat),
      ~ any(abs(.x - locations$lon) < 0.1 & abs(.y - locations$lat) < 0.1)
    )
  ) |>
  mutate(
    location = purrr::pmap_chr(
      list(lon, lat),
      ~ locations$location[
        which.min(abs(.x - locations$lon) + abs(.y - locations$lat))
      ]
    )
  ) |>
  group_by(location, z) |>
  summarise(
    mean_min_annual_temp = mean(min_annual_body_temp, na.rm = TRUE),
    mean_lat = mean(lat),
    .groups = "drop"
  ) |>
  arrange(location, z) |>
  mutate(
    microhabitat = ifelse(z < 0, "leaflitter", "branch")
  )

# create a combined data frame for plotting
validation_df <- micro_df_loc |>
  left_join(otms_val, by = c("location", "microhabitat"))

# plot validation of microclimate model against operative temperature models
validation_df |>
  ggplot(
    aes(
      x = mean_min_annual_temp,
      y = mean_min_temp,
      color = microhabitat,
      fill = microhabitat
    )
  ) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", alpha = 0.25, linetype = "solid") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_manual(
    values = c("leaflitter" = "darkorange", "branch" = "forestgreen"),
    labels = c("leaflitter" = "Leaf Litter", "branch" = "Canopy")
  ) +
  scale_fill_manual(
    values = c("leaflitter" = "darkorange", "branch" = "forestgreen"),
    labels = c("leaflitter" = "Leaf Litter", "branch" = "Canopy")
  ) +
  labs(
    x = "Prediction (°C)",
    y = "Observation (°C)",
    color = "Operative temperature in:",
    fill = "Operative temperature in:"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.25),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    legend.position = "inside",
    legend.position.inside = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.key.spacing.y = unit(0.5, "lines")
  )

```

Below we quantify the agreement between predicted and observed operative temperatures in each microhabitat. We report the bias (mean of predicted minus observed temperatures; positive values indicate the model runs warm), the mean absolute error (MAE) and the root mean square error (RMSE) across individual OTMs, as well as the RMSE expressed as a percentage of the range of observed temperatures (normalized RMSE). We use the normalized RMSE rather than the error relative to each observed temperature because the latter becomes unstable for temperatures close to 0 °C. Because Atlanta contributes about half of all OTMs, we also report the RMSE across site means (i.e., averaging all OTMs per site and microhabitat first), which weights each site equally.

```{r validate-biophys-model-rmse}
#| code-summary: "Calculate bias, MAE and RMSE of biophysical model predictions"

# calculate prediction errors for each OTM
validation_err <- validation_df |>
  mutate(
    microhabitat = ifelse(
      microhabitat == "leaflitter", "Leaf Litter", "Canopy"
    ),
    error = mean_min_annual_temp - mean_min_temp
  )

# calculate RMSE across site means
validation_site_rmse <- validation_err |>
  group_by(microhabitat, location) |>
  summarise(error = mean(error), .groups = "drop") |>
  group_by(microhabitat) |>
  summarise(rmse_site = sqrt(mean(error^2)), .groups = "drop")

# summarise errors by microhabitat
validation_metrics <- validation_err |>
  group_by(microhabitat) |>
  summarise(
    n_sites = n_distinct(location),
    n_otms = n(),
    bias = mean(error),
    mae = mean(abs(error)),
    rmse = sqrt(mean(error^2)),
    nrmse = rmse / diff(range(mean_min_temp)) * 100,
    .groups = "drop"
  ) |>
  left_join(validation_site_rmse, by = "microhabitat")

# table validation metrics
validation_metrics |>
  mutate(across(c(bias, mae, rmse, rmse_site), ~ round(.x, 2))) |>
  mutate(nrmse = round(nrmse, 1)) |>
  relocate(nrmse, .after = rmse) |>
  kbl(
    col.names = c(
      "Microhabitat", "Sites", "OTMs", "Bias (°C)", "MAE (°C)",
      "RMSE (°C)", "RMSE (% of observed range)", "RMSE, site means (°C)"
    ),
    align = "lccccccc"
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
