---
title: "Biophysical Modeling"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

```{r r-setup, warning = FALSE, message = FALSE}
# packages
library(tidyverse)
library(terra)
library(sf)
library(geosphere)
library(rnaturalearth)
library(mapview)
library(mapedit)
library(kableExtra)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)

```

## Overview

## Green anole range

```{r define-caro-range, eval = FALSE}
## Note: This code chunk is not evaluated as it requires manual interaction
## to define the green anole's range using the `mapedit` package. The resulting 
## range points are saved as an RDS file for later use.

caro_occ_data <- read.csv("raw_data/anolis_carolinensis_inat_obs.csv")

# round coordinates and remove observations with low positional accuracy
caro_occ_data <- caro_occ_data |>
  mutate(
    lat = round(latitude, 1), lon = round(longitude, 1),
  ) |>
  filter(
    !is.na(positional_accuracy) & positional_accuracy <= 1000
  ) |>
  group_by(lon, lat) |>
  summarise(n = n(), .groups = "drop") |>
  filter(n > 2) |>
  select(lon, lat)

# filter data to native range of green anole
caro_occ_data <- caro_occ_data |>
  filter(
    lon > -110 & lon < -70, lat > 22.5 & lat < 37.5
  )

# remove outliers based on distance to other observations
caro_occ_data$is_outlier <- apply(
  caro_occ_data[, c("lon", "lat")], 1, function(x) {
    distances <- geosphere::distm(
      x,
      caro_occ_data[, c("lon", "lat")],
      fun = distHaversine
    )
    # if there are fewer than 3 observations within 50 km, mark as outlier
    if (length(distances) > 0 && sum(distances < 50000) < 3) {
      1
    } else {
      0
    }
  }
)

# save processed occurrence data
saveRDS(caro_occ_data, "data/caro_occ_data.rds")

# transform into a spatial object
caro_occ_sf <- st_as_sf(
  caro_occ_data[caro_occ_data$is_outlier == 0, ],
  coords = c("lon", "lat"),
  crs = 4326
)

# manually draw map around range
mapviewOptions(fgb = FALSE)
caro_range_sf <- editMap(mapview(caro_occ_sf))

# transform into a polygon object
caro_range_sf <- st_as_sf(caro_range_sf$finished)

# filter data to only include points within the continental US
caro_range_sf <- st_intersection(
  caro_range_sf,
  ne_countries(scale = "medium", returnclass = "sf") |>
    filter(admin == "United States of America") |>
    st_crop(xmin = -125, xmax = -65, ymin = 20, ymax = 50)
)

# generate a grid of points within the range polygon
grid <- st_make_grid(caro_range_sf, cellsize = 0.1, square = TRUE)
caro_range_points <- grid[st_intersects(grid, caro_range_sf, sparse = FALSE)]
caro_range_points <- as.data.frame(st_coordinates(caro_range_points))
caro_range_points <- caro_range_points |>
  rename(lon = X, lat = Y) |>
  mutate(
    lon = round(lon, 1), lat = round(lat, 1)
  ) |>
  distinct() |>
  select(lon, lat)

# save caro range points
saveRDS(caro_range_points, "data/caro_range_points.rds")

```

```{r plot-range}
#| fig-align: "center"
#| fig-width: 6

caro_range_points <- readRDS("data/caro_range_points.rds")
caro_occ_data <- readRDS("data/caro_occ_data.rds")

ggplot() +
  geom_polygon(
    data = map_data("world"),
    aes(x = long, y = lat, group = group),
    fill = "lightgray", color = "darkgray"
  ) +
  geom_polygon(
    data = map_data("state"),
    aes(x = long, y = lat, group = group),
    fill = NA, color = "gray"
  ) +
  geom_tile(
    data = caro_range_points,
    aes(x = lon, y = lat),
    fill = "forestgreen", alpha = 0.25
  ) +
  geom_point(
    data = caro_occ_data,
    aes(x = lon, y = lat, color = as.factor(is_outlier)),
    size = 1, alpha = 0.75
  ) +
  scale_color_manual(
    values = c("darkgreen", "darkred"),
    labels = c("In Range", "Outlier")
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 37.5)) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(-100, -75, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(25, 37.5, by = 2.5),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank()
  )
```

## Ecological data

```{r get-ecological-data, eval = FALSE}
## Note: This code chunk is not evaluated as it requires downloading large
## datasets. The resulting ecological data is saved as an RDS file
## for later use.

# get polygon for the continental US
continental_us_polygon <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(admin == "United States of America") |>
  st_crop(
    xmin = min(caro_range_points$lon) - 2,
    xmax = max(caro_range_points$lon) + 2,
    ymin = min(caro_range_points$lat) - 2,
    ymax = max(caro_range_points$lat) + 2
  ) |>
  st_transform(crs = 4326)

# download 30s rasters for tree cover, wetland and mangrove
tree_cover_rast <- geodata::landcover(
  var = "trees", path = "raw_data/landuse"
)
wetland_rast <- geodata::landcover(
  var = "wetland", path = "raw_data/landuse"
)
mangrove_rast <- geodata::landcover(
  var = "mangroves", path = "raw_data/landuse"
)
vegetation_rast <- tree_cover_rast + wetland_rast + mangrove_rast

# crop and summarize vegetation rasters
vegetation_rast_us <- terra::crop(vegetation_rast, continental_us_polygon)
vegetation_rast_us <- terra::aggregate(vegetation_rast_us, fact = 12, fun = "mean") # nolint

# rdownload and process minimum temperature data
tmin_rast <- geodata::worldclim_global(
  var = "tmin", res = 0.5, path = "raw_data/wc2.1_30s_tmin"
)
tmin_files <- list.files("raw_data/wc2.1_30s_tmin", full.names = TRUE)
tmin_rast_list <- list()
for (i in seq_along(tmin_files)) {
  rast <- terra::rast(tmin_files[i])
  rast_us <- terra::crop(rast, continental_us_polygon)
  rast_us <- terra::aggregate(rast_us, fact = 12, fun = mean)
  tmin_rast_list <- append(tmin_rast_list, rast_us)
}

# read and process maximum temperature data
tmax_rast <- geodata::worldclim_global(
  var = "tmax", res = 0.5, path = "raw_data/wc2.1_30s_tmax"
)
tmax_files <- list.files("raw_data/wc2.1_30s_tmax", full.names = TRUE)
tmax_rast_list <- list()
for (i in seq_along(tmax_files)) {
  rast <- terra::rast(tmax_files[i])
  rast_us <- terra::crop(rast, continental_us_polygon)
  rast_us <- terra::aggregate(rast_us, fact = 12, fun = mean)
  tmax_rast_list <- append(tmax_rast_list, rast_us)
}

# combine all rasters into a single data frame
eco_rast_list <- c(vegetation_rast_us, tmin_rast_list, tmax_rast_list)
eco_df <- as.data.frame(eco_rast_list, xy = TRUE, na.rm = TRUE)
colnames(eco_df) <- c("lon", "lat", "veg", paste0("tmin_", 1:12), paste0("tmax_", 1:12)) # nolint
eco_df$lon <- round(eco_df$lon, 1)
eco_df$lat <- round(eco_df$lat, 1)

# get elevation raster
elevation <- geodata::elevation_30s(
  lon = eco_df$lon,
  lat = eco_df$lat,
  path = "raw_data/elevation",
  country = "USA"
)
elevation <- terra::crop(elevation, continental_us_polygon)
elevation_df <- as.data.frame(elevation, xy = TRUE, na.rm = TRUE) |>
  setNames(c("lon", "lat", "elevation")) |>
  mutate(lat = round(lat, 1), lon = round(lon, 1)) |>
  group_by(lon, lat) |>
  summarise(elevation = mean(elevation, na.rm = TRUE), .groups = "drop")

# merge data
elevation_df <- merge(
  elevation_df,
  caro_range_points |> mutate(present = 1),
  by = c("lon", "lat"),
  all.x = TRUE
)
elevation_df$present[is.na(elevation_df$present)] <- 0

# combine all rasters into a single data frame
eco_df <- merge(eco_df, elevation_df, by = c("lon", "lat"), all.x = TRUE)
eco_df <- eco_df |> filter(!is.na(elevation))

# pivot longer for temperature data
eco_df <- eco_df |>
  pivot_longer(
    cols = c(starts_with("tmin_"), starts_with("tmax_")),
    names_to = c("temp_type", "month"),
    names_pattern = "(tmin|tmax)_(\\d+)",
    values_to = "temp"
  ) |>
  mutate(month = as.integer(month)) |>
  pivot_wider(
    names_from = temp_type,
    values_from = temp
  ) |>
  unnest(cols = c(tmin, tmax))

# merge with caro range points to indicate presence/absence
eco_df <- merge(
  eco_df,
  caro_range_points |> mutate(present = 1),
  by = c("lon", "lat"),
  all.x = TRUE
)
eco_df$present[is.na(eco_df$present)] <- 0

# save final ecological data
saveRDS(eco_df, "data/eco_data.rds")
```

:::{.panel-tabset}

## Physiological & Behavioral Thresholds

In the plot below, we visualize how the green anole's range is distributed across different temperature thresholds imposed by the physiology and behavior of the species. The thresholds are defined as follows:

- **Hard Threshold**: Minimum annual temperature below 0°C, which is lethal for the species.
- **Soft Threshold**: Minimum annual temperature between 0°C and 5.35°C, which is the lowest ever empirically estimated CT<sub>min</sub>.
- **Behavioral Threshold**: Minimum annual temperature between 5.35°C and 6.6°C, which is the temperature at which there is < 10% of observing green anoles perching at night based on our surveys.

```{r plot-temp-thresholds, warning = FALSE, message = FALSE}
#| fig-align: "center"
#| fig-width: 6

eco_data <- readRDS("data/eco_data.rds")

# define temp thresholds
thresh1 <- 0
thresh2 <- 5.35
thresh3 <- 6.6

eco_data |>
  group_by(lon, lat, present) |>
  summarise(
    min_temp = min(tmin, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    buckets = case_when(
      min_temp < thresh1 ~ "hard_thresh",
      min_temp >= thresh1 & min_temp < thresh2 ~ "soft_thresh",
      min_temp >= thresh2 & min_temp < thresh3 ~ "behavioral_thresh",
      TRUE ~ "rest_of_range"
    )
  ) |>
  mutate(
    buckets = factor(
      buckets,
      levels = c(
        "hard_thresh",
        "soft_thresh",
        "behavioral_thresh",
        "rest_of_range"
      )
    )
  ) |>
  ggplot(aes(x = lon, y = lat, fill = buckets)) +
  geom_tile() +
  geom_contour(
    aes(z = present),
    color = "white", linewidth = 1
  ) +
  scale_fill_manual(
    values = c(
      "hard_thresh" = "darkblue",
      "soft_thresh" = "steelblue",
      "behavioral_thresh" = "skyblue",
      "rest_of_range" = "gray"
    ),
    name = "Temperature\nThresholds",
    labels = c(
      "Hard Threshold (< 0°C)",
      "Soft Threshold (0-5.35°C)",
      "Behavioral Threshold (5.35-6.6°C)",
      "Rest of Range (> 6.6°C)"
    ),
    guide = guide_legend(
      ncol = 2
    )
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(25, 37.5)) +
  scale_x_continuous(
    breaks = seq(-100, -75, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(25, 37.5, by = 2.5),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank(),
  )
```

## Vegetation Cover

In the plot below, we consider any area to be vegetated if it has tree cover, wetland, or mangrove cover. The vegetation cover is expressed as a percentage of the total area.

```{r plot-vegetation-cover}
#| fig-align: "center"
#| fig-width: 6

eco_data |>
  select(lon, lat, veg) |>
  distinct() |>
  ggplot(aes(x = lon, y = lat, fill = veg)) +
  geom_tile() +
  scale_fill_gradient(
    low = "yellow",
    high = "darkgreen",
    name = "Vegetation\nCover"
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(25, 37.5)) +
  scale_x_continuous(
    breaks = seq(-100, -75, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(25, 37.5, by = 2.5),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank(),
    legend.title = element_blank()
  ) +
  labs(
    title = "Vegetation Cover (%)",
  )

```

## Minimum Annual Temperature

```{r plot-min-temp, warning = FALSE, message = FALSE}
#| fig-align: "center"
#| fig-width: 6

eco_data |>
  group_by(lon, lat) |>
  summarise(min_temp = min(tmin, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = lon, y = lat, fill = min_temp)) +
  geom_tile() +
  geom_contour(
    aes(z = min_temp),
    color = "gray80",
    linewidth = 0.2,
    alpha = 0.25
  ) +
  scale_fill_gradient(
    low = "darkblue",
    high = "lightblue",
    name = "Minimum\nAnnual Temp (°C)"
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(25, 37.5)) +
  scale_x_continuous(
    breaks = seq(-100, -75, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous( 
    breaks = seq(25, 37.5, by = 2.5),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank(),
    legend.title = element_blank()
  ) +
  labs(
    title = "Minimum Annual Temperature (°C)",
  )
```

:::

## The green anole's cold-range overfill

A key question is the extent to which the green anole's range is overfilled on the cold side, meaning what % of its range experiences a minimum annual temperature below it's CT<sub>min</sub>. The plot below shows the minimum annual temperature across all latitudes encompassing the green anole's range, along with empirically estimated CT<sub>min</sub> values from two studies.

```{r range-overfill, warning = FALSE, message = FALSE}
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5
#| out-width: "70%"

# generate CTmin data from studies
ctmin_data_cs <- tibble(
  lat = c(25.9, 28.8, 30.266, 32.75, 33.47),
  ctmin = c(10, 9.65, 8.45, 8.32, 7.8),
  study = "Campbell-Staton et al., 2017"
)
ctmin_data_husak <- tibble(
  lat = c(26.51, 29.9, 32.77, 35.44),
  ctmin = 2^c(2.7, 2.85, 2.56, 2.42),
  study = "Husak et al., 2026"
)
ctmin_data <- rbind(ctmin_data_cs, ctmin_data_husak)

eco_data |>
  filter(present == 1) |>
  group_by(lon, lat) |>
  summarise(
    tmin = min(tmin, na.rm = TRUE),
  ) |>
  group_by(lat) |>
  summarise(
    mean_tmin = mean(tmin, na.rm = TRUE),
    min_tmin = min(tmin, na.rm = TRUE),
    max_tmin = max(tmin, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = lat)) +
  geom_ribbon(
    aes(ymin = min_tmin, ymax = max_tmin),
    fill = "lightblue", alpha = 0.5
  ) +
  geom_line(
    aes(y = mean_tmin),
    color = "darkblue",
    size = 1
  ) +
  geom_point(
    data = ctmin_data,
    aes(x = lat, y = ctmin, color = study),
    size = 3
  ) +
  geom_smooth(
    data = ctmin_data,
    aes(x = lat, y = ctmin, color = study),
    method = "lm",
    se = FALSE,
    fullrange = TRUE,
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  scale_color_manual(
    values = c("darkorange", "purple"),
    name = expression(paste("Empirical CT"[min]," from:"))
  ) +
  scale_x_continuous(
    expand = c(0, 0)
  ) +
  labs(
    x = "Latitude (°N)",
    y = expression(paste("Minimum Annual Temperature (°C)"))
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.75, 0.9),
    legend.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.grid = element_blank()
  ) +
  coord_cartesian(xlim = c(25, 37.5))
```

Based on this pattern, we can assess that, even in the most generous scenario, the green anole's range is overfilelled on the cold side by > 87% of its range based on the lowest empirically estimated CT<sub>min</sub> value. Further, > 25% of its range falls below 0°C. The table below summarizes the proportion of the green anole's range that falls below the lowest, average, and maximum empirically estimated CT<sub>min</sub> values.

```{r table-proportions-below-ctmin}
# table proportions of range falling below lowest, average and maximum CTmin
eco_data |>
  filter(present == 1) |>
  group_by(lon, lat) |>
  summarise(
    tmin = min(tmin, na.rm = TRUE),
  ) |>
  mutate(
    below_lowest_ctmin = tmin < min(ctmin_data$ctmin, na.rm = TRUE),
    below_avg_ctmin = tmin < mean(ctmin_data$ctmin, na.rm = TRUE),
    below_max_ctmin = tmin < max(ctmin_data$ctmin, na.rm = TRUE),
    below_0 = tmin < 0
  ) |>
  ungroup() |>
  summarise(
    prop_below_lowest_ctmin = mean(below_lowest_ctmin, na.rm = TRUE),
    prop_below_avg_ctmin = mean(below_avg_ctmin, na.rm = TRUE),
    prop_below_max_ctmin = mean(below_max_ctmin, na.rm = TRUE),
    prop_below_0 = mean(below_0, na.rm = TRUE),
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "ctmin_category",
    values_to = "proportion"
  ) |>
  mutate(
    ctmin_value = case_when(
      ctmin_category == "prop_below_lowest_ctmin" ~ round(min(ctmin_data$ctmin, na.rm = TRUE), 2),
      ctmin_category == "prop_below_avg_ctmin" ~ round(mean(ctmin_data$ctmin, na.rm = TRUE), 2),
      ctmin_category == "prop_below_max_ctmin" ~ round(max(ctmin_data$ctmin, na.rm = TRUE), 2),
      ctmin_category == "prop_below_0" ~ 0
    ),
    ctmin_category = case_when(
      ctmin_category == "prop_below_lowest_ctmin" ~ paste("Lowest CTmin - ", ctmin_value, "°C"),
      ctmin_category == "prop_below_avg_ctmin" ~ paste("Average CTmin - ", ctmin_value, "°C"),
      ctmin_category == "prop_below_max_ctmin" ~ paste("Maximum CTmin - ", ctmin_value, "°C"),
      ctmin_category == "prop_below_0" ~ paste("0°C (Freezing Threshold)")
    )
  ) |>
  select(-ctmin_value) |>
  rename(
    "Threshold" = ctmin_category,
    "Proportion of Range Below" = proportion
  ) |>
  kable() |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

Below it is also expressed in degrees latitude overfilled.

```{r table-overfilling-degrees-lat}
eco_data |>
  filter(present == 1) |>
  group_by(lon, lat) |>
  summarise(
    tmin = min(tmin, na.rm = TRUE),
  ) |>
  mutate(
    below_lowest_ctmin = tmin < min(ctmin_data$ctmin, na.rm = TRUE),
    below_avg_ctmin = tmin < mean(ctmin_data$ctmin, na.rm = TRUE),
    below_highest_ctmin = tmin < max(ctmin_data$ctmin, na.rm = TRUE),
    below_0 = tmin < 0
  ) |>
  ungroup() |>
  pivot_longer(
    cols = c(below_lowest_ctmin, below_avg_ctmin, below_highest_ctmin, below_0),
    names_to = "category",
    values_to = "below_threshold"
  ) |>
  filter(below_threshold) |>
  group_by(category) |>
  filter(lat == min(lat, na.rm = TRUE)) |>
  slice(1) |>
  ungroup() |>
  mutate(
    degrees_overfill = max(caro_range_points$lat, na.rm = TRUE) - lat
  ) |>
  mutate(
    category = case_when(
      category == "below_lowest_ctmin" ~ paste("Lowest CTmin - ", round(min(ctmin_data$ctmin, na.rm = TRUE), 2), "°C"),
      category == "below_avg_ctmin" ~ paste("Average CTmin - ", round(mean(ctmin_data$ctmin, na.rm = TRUE), 2), "°C"),
      category == "below_highest_ctmin" ~ paste("Maximum CTmin - ", round(max(ctmin_data$ctmin, na.rm = TRUE), 2), "°C"),
      category == "below_0" ~ paste("0°C (Freezing Threshold)")
    )
  ) |>
  select(category, degrees_overfill) |>
  rename(
    "Threshold" = category,
    "Degrees Overfill" = degrees_overfill
  ) |>
  kable() |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )

```

This degree of range overfilling falls 

## Biophysical Modeling

```{r define-params}

# define physical constants
k <- 0.05 # thermal conductivity of leaf litter (W m^-1 K^-1)
rho <- 150 # bulk density of leaf litter (kg m^-3)
cp <- 2500 # specific heat capacity of leaf litter (J kg^-1 K^-1)
cc <- 0.007 # canopy cooling coefficient (W m^-2)
srh <- 0.02 # surface radiative heat flux coefficient (C / W m^-2 K^-1)
ecc <- 0.0167 # earth orbit eccentricity
spo <- 1.36 # extraterrestial solar radiation (kW m^-2)
atc <- 0.83 # atmospheric transmissivity coefficient (unitless)

# calculate derived physical constants
diff_sec <- k / (rho * cp) # thermal diffusivity of leaf litter (m^2 s^-1)
diff_hour <- diff_sec * 3600 # thermal diffusivity of leaf litter (m^2 h^-1)
omega <- 2 * pi / 24 # angular frequency of diurnal temperature cycle (rad h^-1)
damping_depth <- sqrt((2 * diff_hour) / omega) # damping depth (m)

# define list of heights and depths of interest
zs <- c(seq(-0.2, -0.01, by = 0.01), 0, 1.8, 2.2, 3)

```

```{r estimate-microclimate, eval = FALSE}
## Note: This code chunk is not evaluated as it requires significant computation time. 
## The resulting microclimate data is saved as an RDS file for later use.

eco_df <- eco_data

# define function to calculate solar radiation profile
get_solar_rad_profile <- function(lat, elev, doy, solar_noon) {
  hours <- 0:23
  elon <- (2 * pi / 365) * (doy - 80) + 2 * ecc * (sin(2 * pi / 365 * doy) - sin(2 * pi / 365 * 80)) # nolint
  dec_ang <- asin(0.3879672 * sin(elon))
  p_a <- 101.3 * exp(-elev / 8200)
  geo_lat <- lat * (pi / 180)
  sin_alt_ang <- sin(geo_lat) * sin(dec_ang) + cos(geo_lat) * cos(dec_ang) * outer(solar_noon, hours, function(s, h) cos(0.2618 * (h - s))) # nolint
  m_a <- p_a / (101.3 * (1 / sin_alt_ang))
  solar_rad <- spo * atc ^ m_a * sin_alt_ang
  solar_rad[solar_rad < 0] <- 0
  drop(solar_rad * 1000)
}

# define function to calculate air temperature profile
get_air_temp_profile <- function(tmax, tmin) {
  w <- pi / 12
  gamma <- 0.44 - 0.46 * sin(0.9 + w * (0:23)) + 0.11 * sin(0.9 + 2 * w * (0:23)) # nolint
  drop(outer(tmax, gamma) + outer(tmin, 1 - gamma))
}

# define function to correct air temperature profile from canopy cooling
correct_air_temp_profile <- function(
    air_temp_profile, cc = 0.007, solar_rad_profile, veg_cover) {
  is_vec <- is.vector(air_temp_profile)
  if (is_vec) {
    air_temp_profile  <- matrix(air_temp_profile,  nrow = 1)
    solar_rad_profile <- matrix(solar_rad_profile, nrow = 1)
  }
  result <- air_temp_profile - sweep(cc * solar_rad_profile, 1, veg_cover, "*")
  if (is_vec) drop(result) else result
}

# define function to calculate surface temperature profile
get_surf_temp_profile <- function(
    air_temp_profile, solar_rad_profile, srh = 0.02, veg_cover) {
  is_vec <- is.vector(air_temp_profile)
  if (is_vec) {
    air_temp_profile  <- matrix(air_temp_profile,  nrow = 1)
    solar_rad_profile <- matrix(solar_rad_profile, nrow = 1)
  }
  result <- air_temp_profile + sweep(srh * solar_rad_profile, 1, 1 - veg_cover, "*") # nolint
  if (is_vec) drop(result) else result
}

get_z_temp_profile <- function(
  z_list, surf_temp_profile, air_temp_profile, damping_depth, ref_height = 1.8
) {

  # check if the input profiles are vectors or matrices
  is_vec <- is.vector(surf_temp_profile)
  if (is_vec) {
    surf_temp_profile <- matrix(surf_temp_profile, nrow = 1)
    air_temp_profile  <- matrix(air_temp_profile,  nrow = 1)
  }

  # define hours and extended hours for interpolation
  hours <- 0:23
  surf_temp_mean <- rowMeans(surf_temp_profile)
  extended_hours <- c(hours - 24, hours, hours + 24)

  # calculate temperature profiles at different heights and depths
  z_temp_profiles <- lapply(z_list, function(z) {
    if (z == 0) {
      surf_temp_profile
    } else if (z == ref_height) {
      air_temp_profile
    } else if (z > 0) {
      a <- log((z + 0.01) / 0.01) / log((ref_height + 0.01) / 0.01)
      surf_temp_profile - (surf_temp_profile - air_temp_profile) * a
    } else {
      damp_ratio <- exp(-abs(z) / damping_depth)
      phase_lag <- (abs(z) / damping_depth) * (24 / (2 * pi))
      xout <- hours - phase_lag
      idx <- findInterval(xout, extended_hours)
      w <- (xout - extended_hours[idx]) / (extended_hours[idx + 1] - extended_hours[idx]) # nolint
      ext_surf <- cbind(surf_temp_profile, surf_temp_profile, surf_temp_profile)
      shifted_temp <- sweep(ext_surf[, idx],     2, 1 - w, "*") + sweep(ext_surf[, idx + 1], 2, w, "*") # nolint
      surf_temp_mean + (shifted_temp - surf_temp_mean) * damp_ratio
    }
  })

  # combine the temperature profiles into a matrix
  names(z_temp_profiles) <- z_list
  if (is_vec) z_temp_profiles <- lapply(z_temp_profiles, drop)
  z_temp_profiles

}

# get solar noon time for each point in eco_df
eco_df$solar_noon <- mapply(
  TrenchR::solar_noon,
  lon = eco_df$lon,
  doy = eco_df$month * 30 - 15
)

# calculate air temperature profile for every point in eco_df
air_temp_prof <- get_air_temp_profile(
  tmax = eco_df$tmax,
  tmin = eco_df$tmin
)

# calculate solar radiation profile for every point in eco_df
solar_rad_prof <- get_solar_rad_profile(
  lat = eco_df$lat,
  elev = eco_df$elevation,
  doy = eco_df$month * 30 - 15,
  solar_noon = eco_df$solar_noon
)

# calculate corrected air temperature profile for every point in eco_df
air_temp_prof_corrected <- correct_air_temp_profile(
  air_temp_profile = air_temp_prof,
  solar_rad_profile = solar_rad_prof,
  veg_cover = eco_df$veg
)

# calculate surface temperature profile for every point in eco_df
surf_temp_prof <- get_surf_temp_profile(
  air_temp_profile = air_temp_prof_corrected,
  solar_rad_profile = solar_rad_prof,
  veg_cover = eco_df$veg
)

# calculate temperature profiles at different heights and depths for every point in eco_df # nolint
zs_temp_profiles <- get_z_temp_profile(
  z_list = zs,
  surf_temp_profile = surf_temp_prof,
  air_temp_profile = air_temp_prof_corrected,
  damping_depth = damping_depth
)

# summarize zs temperature profiles into a data frame
microclimate_temps <- Map(
  function(z_temp, z_val) {
    data.frame(
      lon = eco_df$lon,
      lat = eco_df$lat,
      elev = eco_df$elevation,
      veg = eco_df$veg,
      presence = eco_df$present,
      z                = z_val,
      min_temp         # calculate temperature profiles at different heights and depths for every point in eco_df # nolint
zs_temp_profiles <- get_z_temp_profile(
  z_list = zs,
  surf_temp_profile = surf_temp_prof,
  air_temp_profile = air_temp_prof_corrected,
  damping_depth = damping_depth
)= do.call(pmin, as.data.frame(z_temp)),
      hours_below_0C   = rowSums(z_temp < 0),
      hours_below_6.8C = rowSums(z_temp < 6.8)
    )
  },
  zs_temp_profiles, zs
) |>
  bind_rows() |>
  group_by(lon, lat, elev, veg, presence, z) |>
  summarise(
    min_annual_temp   = min(min_temp),
    hours_below_0C    = sum(hours_below_0C),
    hours_below_ctmin = sum(hours_below_6.8C),
    .groups = "drop"
  )

# save zs_temp_profiles
saveRDS(microclimate_temps, "data/microclimate_temps.rds")

```

```{r plot-microclimate-by-lat}

microclimate_temps <- readRDS("data/microclimate_temps.rds")

summarized_microclimates <- microclimate_temps |>
  filter(presence == 1) |>
  group_by(lat, z) |>
  summarise(
    mean_min_temp = mean(min_annual_temp, na.rm = TRUE),
    max_min_temp = max(min_annual_temp, na.rm = TRUE),
    min_min_temp = min(min_annual_temp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(as.character(z) %in% c("-0.05", "2.2")) |>
  mutate(
    z = factor(z, levels = c("-0.05", "2.2"), labels = c("Leaf Litter", "Canopy"))
  )

ggplot() +
  geom_ribbon(
    data = summarized_microclimates,
    aes(
      x = lat,
      ymin = min_min_temp,
      ymax = max_min_temp,
    ),
    fill = "lightblue", alpha = 0.5
  ) +
  geom_line(
    data = summarized_microclimates,
    aes(
      x = lat,
      y = mean_min_temp
    ),
    linewidth = 1, color = "darkblue"
  ) +
  geom_point(
    data = ctmin_data,
    aes(x = lat, y = ctmin, color = study),
    size = 3
  ) +
  geom_smooth(
    data = ctmin_data,
    aes(x = lat, y = ctmin, color = study),
    method = "lm",
    se = FALSE,
    fullrange = TRUE,
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "red"
  ) +
  scale_color_manual(
    values = c("darkorange", "purple"),
    name = expression(paste("Empirical CT"[min]," from:"))
  ) +
  scale_x_continuous(
    expand = c(0, 0)
  ) +
  labs(
    x = "Latitude (°N)",
    y = expression(paste("Minimum Annual Temperature (°C)"))
  ) +
  theme_minimal() +
  theme(
    legend.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  facet_wrap(~z)

```