---
title: "Leaf litter temperatures predict the green anole's poleward range limit"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

## Overview

We hypothesize that the green anole's poleward range edge is set not by the cold air temperatures in the canopy, but by the thermal conditions within the leaf litter where lizards overwinter. In this vignette, we use mechanistic envelope models to evaluate which microclimate and forest cover threshold best predict the green anole's range, both across its entire range and near its poleward range edge. We use the green anole's range and the minimum annual body temperatures in the canopy and at different depths in the leaf litter that we estimated in the [biophysical models](biophysical-models.qmd) vignette.

```{r r-setup, warning = FALSE, message = FALSE}
#| code-summary: "R setup"

# packages
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(TrenchR)
library(kableExtra)
library(viridis)
library(ggpubr)
library(geosphere)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)
```

```{r load-biophysical-model-results}
#| code-summary: "Load biophysical model results"

# minimum annual body temperatures in each microclimate and forest cover, and
# presence of the green anole, for each spatial cell (see
# biophysical-models.qmd)
micro_df <- readRDS("data/micro_df.rds")
```

## Mechanistic envelope models

Envelope models compare the capacity of a set of threshold values across environmental variables to predict a species' range. These models assign a binary suitability score to each spatial cell based on where the values of the environmental variables fall relative to the thresholds of interest. By comparing the inferred suitability map against the observed range cell-by-cell, we can obtain the True Skill Statistic (TSS), which evaluates model performance as a function of both sensitivity and specificity (i.e., its ability to simultaneously predict presence and absence):

$$
TSS = \frac{TP}{TP + FN} + \frac{TN}{TN + FP} - 1
$$

where $TP$, $FN$, $TN$ and $FP$ are the number of true positives, false negatives, true negatives and false positives, respectively. TSS ≤ 0 indicates a predictive ability no better than random, while a value of 1 indicates a perfect match. Thus, if a set of thresholds A yields a greater TSS than a set B, we can infer that the species' range is more likely determined by A than by B.

Traditionally, envelope models rely on macroclimatic layers (e.g., air temperatures) and statistically identified thresholds. Instead, our mechanistic approach uses the minimum annual body temperatures estimated with our biophysical models for each microclimate and sets the threshold based on the green anole's physiology. Specifically, we deem a cell unsuitable if the minimum annual body temperature in a given microclimate falls below 0 °C, which we assume is the lethal cold threshold for the green anole. Because the green anole is arboreal during its active season, we also include forest cover as a second variable, but we do not set its threshold *a priori*. Instead, we evaluate every forest cover threshold between 0 and 100% at 5% intervals. The combination of microclimate and forest cover threshold that yields the highest TSS is the most probable determinant of the green anole's range.

What limits a species' range at its poleward edge might differ from what limits it within its core. For this reason, we repeat the analysis across four spatial extents defined by the minimum annual temperatures in the canopy (1.8 m):

1. **Entire range**: all spatial cells for which we obtained estimates from our biophysical models.

2. **Behavioral threshold**: only cells where canopy temperatures fall below the retreating temperature we inferred from our surveys (T<sub>ret</sub> = 8.17 °C, at the average photoperiod and canopy temperature of our surveys; see [Green anoles retreat to the leaf litter to avoid extreme cold](leaf-litter-retreat.qmd)).

3. **Soft physiological threshold**: only cells where canopy temperatures fall below the lowest population-average CT<sub>min</sub> of the species (6.8 °C).

4. **Hard physiological threshold**: only cells where canopy temperatures fall below 0 °C.

Below we classify every cell according to these thresholds and plot them.

```{r define-thresholds}
#| code-summary: "Define thresholds for mechanistic envelope models"
#| fig-align: "center"
#| fig-width: 7
#| out-width: "66%"

# thresholds (°C): Tret from the temperature x photoperiod model in
# leaf-litter-retreat.qmd and the lowest population-average CTmin across the
# range (live-beyond-ctmin.qmd)
tret <- 8.17
ctmin_min <- 6.8

# estimate thresholds
thresh <- micro_df |>
  filter(z == 1.8) |>
  mutate(
    thresh = case_when(
      min_annual_temp < 0 ~ "below_zero",
      min_annual_temp < ctmin_min ~ "below_ctmin",
      min_annual_temp < tret ~ "below_tret",
      TRUE ~ "rest_of_range"
    )
  ) |>
  select(lon, lat, elev, veg, presence, thresh)

# join micro_df with thresholds
micro_df <- micro_df |>
  left_join(thresh, by = c("lon", "lat", "elev", "veg", "presence"))

# range edge: boundaries between adjacent land cells where the green anole is
# present and absent. Unlike a contour, this does not trace the coastline. The
# range polygon also stops slightly short of the coast in many places, leaving
# a narrow fringe of "absent" coastal cells; we ignore boundaries with absent
# cells within `coast_buffer` cells (~4 km each) of the coast so that the line
# only shows the inland range edge
range_edge <- function(cells, res = 1 / 24, coast_buffer = 3,
                       min_neighbors = 10) {
  grid <- cells |>
    mutate(
      i = round((lon - min(lon)) / res),
      j = round((lat - min(lat)) / res),
      key = paste(i, j)
    ) |>
    select(i, j, key, lon, lat, presence)
  shift_key <- function(df, di, dj) paste(df$i + di, df$j + dj)
  steps <- list(c(1, 0), c(-1, 0), c(0, 1), c(0, -1))

  # cells within coast_buffer cells of the coast (i.e., of a missing cell)
  near_coast <- !Reduce(
    `&`, lapply(steps, \(k) shift_key(grid, k[1], k[2]) %in% grid$key)
  )
  for (b in seq_len(coast_buffer - 1)) {
    near_keys <- grid$key[near_coast]
    near_coast <- near_coast | Reduce(
      `|`, lapply(steps, \(k) shift_key(grid, k[1], k[2]) %in% near_keys)
    )
  }
  grid$near_coast <- near_coast

  # boundaries with the eastern and northern neighbor of each cell
  neighbor <- function(di, dj) {
    grid |>
      mutate(i = i - di, j = j - dj) |>
      select(i, j, presence_n = presence, near_coast_n = near_coast)
  }
  inland_boundary <- function(df) {
    df |>
      filter(presence != presence_n) |>
      # skip boundaries where the absent cell is in the coastal fringe
      filter(!ifelse(presence == 0, near_coast, near_coast_n))
  }
  east <- grid |>
    inner_join(neighbor(1, 0), by = c("i", "j")) |>
    inland_boundary() |>
    transmute(
      x = lon + res / 2, xend = lon + res / 2,
      y = lat - res / 2, yend = lat + res / 2
    )
  north <- grid |>
    inner_join(neighbor(0, 1), by = c("i", "j")) |>
    inland_boundary() |>
    transmute(
      x = lon - res / 2, xend = lon + res / 2,
      y = lat + res / 2, yend = lat + res / 2
    )
  edges <- bind_rows(east, north)

  # drop isolated specks (e.g., small islands): keep segments with at least
  # `min_neighbors` other segments within 0.25 degrees, which is always true
  # along the continuous range edge
  mid_x <- (edges$x + edges$xend) / 2
  mid_y <- (edges$y + edges$yend) / 2
  n_close <- vapply(seq_len(nrow(edges)), \(k) {
    sum((mid_x - mid_x[k])^2 + (mid_y - mid_y[k])^2 < 0.25^2) - 1
  }, numeric(1))
  edges[n_close >= min_neighbors, ]
}
range_edge_segments <- range_edge(thresh)

# plot thresholds
thresh_plot <- thresh |>
  ggplot(aes(x = lon, y = lat, fill = thresh)) +
  geom_raster() +
  geom_segment(
    data = range_edge_segments,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "white", linewidth = 0.5, lineend = "square",
    inherit.aes = FALSE
  ) +
  scale_x_continuous(
    breaks = seq(-100, -75, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(25, 35, by = 2.5),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c(
      "below_zero" = "darkblue",
      "below_ctmin" = "steelblue",
      "below_tret" = "lightblue",
      "rest_of_range" = "gray80"
    ),
    breaks = c("below_zero", "below_ctmin", "below_tret"),
    labels = c(
      "below_zero" = "< 0°C",
      "below_tret" = expression("< " * T[ret]),
      "below_ctmin" = expression("< " * CT[min])
    )
  ) +
  coord_cartesian(xlim = c(-101, -75), ylim = c(24, 37.5), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.border = element_rect(fill = NA),
    axis.ticks = element_line(),
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_text(size = 10),
    legend.position = "inside",
    legend.position.inside = c(0.5, 0.06),
    legend.justification = c(0.5, 0),
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.key.spacing.x = unit(0.5, "lines")
  )
```

Next, we run the mechanistic envelope models for every combination of spatial extent, microclimate and forest cover threshold, and store the resulting TSS values in a grid (`tss_grid`).

```{r run-memo, eval = FALSE}
#| code-summary: "Run mechanistic envelope models"
#| code-fold: false

## Note: This code chunk is not evaluated as it takes a long time to run.
## The output is saved in the `data/tss_grid.rds` file.

# define TSS grid
thresh_names <- unique(micro_df$thresh)
veg <- seq(0, 1, by = 0.05)
zs <- unique(micro_df$z)
tss_grid <- expand.grid(
  thresh = thresh_names,
  veg = veg,
  z = zs
)
tss_grid$tss <- NA

# loop through
for (i in seq_len(nrow(tss_grid))) {

  # filter data of interest
  micro_df_sub <- micro_df |> filter(z == tss_grid$z[i])
  if (tss_grid$thresh[i] == "below_zero") {
    micro_df_sub <- micro_df_sub |> filter(thresh == "below_zero")
  } else if (tss_grid$thresh[i] == "below_ctmin") {
    micro_df_sub <- micro_df_sub |> 
      filter(thresh %in% c("below_zero", "below_ctmin"))
  } else if (tss_grid$thresh[i] == "below_tret") {
    micro_df_sub <- micro_df_sub |>
      filter(thresh %in% c("below_zero", "below_ctmin", "below_tret"))
  }

  tss_value <- micro_df_sub |>
    mutate(
      pred = min_annual_body_temp > 0 & veg > tss_grid$veg[i]
    ) |>
    summarise(
      tp = sum(pred & presence),
      fp = sum(pred & !presence),
      fn = sum(!pred & presence),
      tn = sum(!pred & !presence)
    ) |>
    mutate(tss = (tp / (tp + fn)) + (tn / (tn + fp)) - 1) |>
    pull(tss)
  tss_grid$tss[i] <- tss_value

  # track progress
  if (i %% 100 == 0) {
    message(paste("Processed", i, "of", nrow(tss_grid), "rows"))
  }

}

# save tss grid
saveRDS(tss_grid, "data/tss_grid.rds")

```

Lastly, we identify the combination of microclimate and forest cover threshold that maximizes TSS for each spatial extent (Table S8 in the manuscript) and plot the results (see the [manuscript Figure 4](#manuscript-figure-4) section at the end of this vignette for the final version of this figure). Panel A shows the areas below each threshold. Panel B shows the TSS values of every combination of microclimate (y-axis) and forest cover threshold (x-axis) for each spatial extent, with red circles marking the combination that maximizes TSS.

Across all spatial extents, leaf-litter microclimates predict the green anole's range better than the canopy, with depths of 3-4 cm performing best. This effect is strongest in areas where canopy temperatures fall below 0 °C, where the poleward range edge is located: here, canopy microclimates have no predictive capacity (TSS = 0), whereas leaf-litter microclimates yield the highest predictive accuracy of all conditions tested.

```{r plot-memo-results}
#| code-summary: "Plot mechanistic envelope model results"
#| fig-align: "center"
#| fig-height: 11.5
#| fig-width: 6


# load TSS grid
tss_grid <- readRDS("data/tss_grid.rds")

# get maximum TSS values for each threshold
tss_grid_max <- tss_grid |>
  group_by(thresh) |>
  filter(tss == max(tss, na.rm = TRUE)) |>
  ungroup()
tss_grid_max

# refactor threshold names for plotting
tss_grid$thresh <- factor(
  tss_grid$thresh,
  levels = c("below_zero", "below_ctmin", "below_tret", "rest_of_range"),
)

# plot TSS grid
tss_grid_plot <- tss_grid |>
  filter(z < 2) |>
  mutate(tile_height = ifelse(z > 0, 0.2, 0.01)) |>
  mutate(z_mod = ifelse(z > 0, 0.1, z)) |>
  as_tibble() |>
  arrange(desc(tile_height)) |>
  ggplot(aes(x = veg, fill = tss)) +
  geom_tile(aes(y = z_mod, height = tile_height)) +
  geom_point(
    aes(y = z),
    data = tss_grid_max,
    color = "red",
    size = 3,
    shape = 1
  ) +
  geom_hline(
    yintercept = 0, color = "white", linetype = "solid"
  ) +
  scale_fill_viridis_b(
    name = "TSS",
    limits = c(
      min(tss_grid$tss, na.rm = TRUE), max(tss_grid$tss, na.rm = TRUE)
    ),
    breaks = seq(
      min(tss_grid$tss, na.rm = TRUE), max(tss_grid$tss, na.rm = TRUE),
      by = 0.05
    ),
    # only show labels for every other break
    labels = function(x) ifelse(seq_along(x) %% 2 == 1, x, ""),
    guide = guide_colorsteps(
      barheight = unit(8, "cm"),
      barwidth = unit(0.5, "cm")
    )
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.2),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = c(0.2, 0, -0.05, -0.1, -0.2),
    labels = c("1.8", "0", "-0.05", "-0.1", "-0.2")
  ) +
  facet_wrap(
    ~thresh,
    ncol = 2,
    nrow = 2,
    axes = "all",
    labeller = labeller(
      thresh = as_labeller(
        c(
          below_zero = "bold(atop('< Hard Physiological', 'Threshold (' * 0 * degree * C * ')'))", # nolint
          below_ctmin = "bold(atop('< Soft Physiological', 'Threshold (' * CT[min] * ')'))", # nolint
          below_tret = "bold(atop('< Behavior', 'Threshold (' * T[ret] * ')'))", # nolint
          rest_of_range = "bold(atop('Entire', 'range'))"
        ),
        label_parsed
      )
    )
  ) +
  labs(
    x = "Forest Cover",
    y = "Distance from surface above or below leaf litter (m)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 12),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    axis.title = element_text(size = 14, lineheight = 1.25),
    legend.title = element_text(size = 14, hjust = 0.5, vjust = 1),
    axis.text.y = element_text(size = 10)
  )

# align the map with the TSS panels: the map spans the width of the two panel
# columns (~4.4 in), and its height (in inches) keeps the map's aspect ratio
# (as in coord_quickmap) plus space for the axis labels
map_aspect <- (37.5 - 24) / ((-75 - -101) * cos(mean(c(24, 37.5)) * pi / 180))
map_height <- 4.4 * map_aspect + 0.6
cowplot::plot_grid(
  thresh_plot + theme(plot.margin = margin(t = 15, r = 5, b = 5, l = 5)),
  tss_grid_plot,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(map_height, 11.5 - map_height),
  labels = c("A", "B"),
  label_size = 16
)
```

## Contribution of each variable

To better understand what drives the predictive performance of the mechanistic envelope models, we evaluate the predictive capacity of each variable on its own.

```{r run-memo-extra, eval = FALSE}
#| code-summary: "Run single-variable envelope models"
#| code-fold: false

## Note: This code chunk is not evaluated as it takes a long time to run.
## The output is saved in the `data/tss_single.rds` file.

# spatial extents (nested sets of threshold classes)
extents <- list(
  below_zero = "below_zero",
  below_ctmin = c("below_zero", "below_ctmin"),
  below_tret = c("below_zero", "below_ctmin", "below_tret"),
  rest_of_range = c("below_zero", "below_ctmin", "below_tret", "rest_of_range")
)
veg_thresholds <- seq(0, 1, by = 0.05)

# TSS from predicted and observed presences
calc_tss <- function(pred, presence) {
  tp <- sum(pred & presence)
  fn <- sum(!pred & presence)
  tn <- sum(!pred & !presence)
  fp <- sum(pred & !presence)
  (tp / (tp + fn)) + (tn / (tn + fp)) - 1
}

# presence as logical
micro_df <- micro_df |>
  mutate(presence = presence == 1)

# single-variable models across each spatial extent
# forest cover alone (forest cover does not depend on the microclimate, so we
# use one row per cell)
cells <- micro_df |> filter(z == 1.8)
tss_forest <- expand_grid(thresh = names(extents), veg = veg_thresholds) |>
  mutate(
    tss = map2_dbl(thresh, veg, \(e, v) {
      d <- cells[cells$thresh %in% extents[[e]], ]
      calc_tss(d$veg > v, d$presence)
    }),
    model = "forest"
  )

# temperature alone
tss_temp <- expand_grid(
  thresh = names(extents),
  z = sort(unique(micro_df$z))
) |>
  mutate(
    tss = map2_dbl(thresh, z, \(e, zz) {
      d <- micro_df[micro_df$z == zz & micro_df$thresh %in% extents[[e]], ]
      calc_tss(d$min_annual_body_temp > 0, d$presence)
    }),
    model = "temperature"
  )

saveRDS(bind_rows(tss_forest, tss_temp), "data/tss_single.rds")
```

### Temperature and forest cover alone

We evaluate the predictive capacity of each variable on its own across the four spatial extents. For forest cover alone, we deem a cell suitable if its forest cover exceeds a given threshold (0–100% at 5% intervals), regardless of temperature. For temperature alone, we deem a cell suitable if the minimum annual body temperature in a given microclimate is above 0 °C, regardless of forest cover. Below we table the highest TSS achieved by forest cover alone, by temperatures in the canopy alone, by temperatures in the best-performing microclimate alone and by the combination of temperature and forest cover (i.e., the full models above).

```{r table-memo-single}
#| code-summary: "Table TSS of single-variable and combined envelope models"

# load results
tss_single <- readRDS("data/tss_single.rds")

# label microclimates
label_microclimate <- function(z) {
  case_when(
    z > 0 ~ paste0("Canopy (", z, " m)"),
    z == 0 ~ "Leaf-litter surface",
    TRUE ~ paste0(round(-z * 100), " cm deep in leaf litter")
  )
}
extent_labels <- c(
  below_zero = "Canopy T<sub>min</sub> < 0 °C",
  below_ctmin = "Canopy T<sub>min</sub> < CT<sub>min</sub>",
  below_tret = "Canopy T<sub>min</sub> < T<sub>ret</sub>",
  rest_of_range = "Entire area"
)

# best model of each type for each spatial extent
best_forest <- tss_single |>
  filter(model == "forest") |>
  group_by(thresh) |>
  slice_max(tss, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(thresh, forest_tss = tss, forest_veg = veg)
best_canopy <- tss_single |>
  filter(model == "temperature", z == 1.8) |>
  transmute(thresh, canopy_tss = tss)
best_temp <- tss_single |>
  filter(model == "temperature") |>
  group_by(thresh) |>
  slice_max(tss, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(thresh, temp_tss = tss, temp_z = z)
best_full <- tss_grid |>
  mutate(thresh = as.character(thresh)) |>
  group_by(thresh) |>
  slice_max(tss, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(thresh, full_tss = tss, full_z = z, full_veg = veg)

memo_single <- best_forest |>
  left_join(best_canopy, by = "thresh") |>
  left_join(best_temp, by = "thresh") |>
  left_join(best_full, by = "thresh") |>
  ungroup() |>
  mutate(thresh = factor(thresh, levels = names(extent_labels))) |>
  arrange(thresh)

table_memo_single <- memo_single |>
  mutate(
    thresh = extent_labels[as.character(thresh)],
    forest = sprintf("%.3f (> %.0f%%)", forest_tss, forest_veg * 100),
    canopy = sprintf("%.3f", canopy_tss),
    temp = sprintf("%.3f (%s)", temp_tss, label_microclimate(temp_z)),
    full = sprintf(
      "%.3f (%s; > %.0f%%)",
      full_tss, label_microclimate(full_z), full_veg * 100
    )
  ) |>
  select(thresh, forest, canopy, temp, full) |>
  kbl(
    col.names = c(
      "Spatial extent", "Forest cover", "Canopy (1.8 m)",
      "Best microclimate", "Best microclimate + forest cover"
    ),
    escape = FALSE,
    align = "lcccc"
  ) |>
  kable_styling(full_width = FALSE, position = "center") |>
  add_header_above(
    c(" " = 1,
      "Forest cover alone" = 1,
      "Temperature alone" = 2,
      "Temperature and forest cover" = 1
    )
  ) |>
  footnote(
    general = paste(
      "Values are the highest TSS achieved by each type of model, with the",
      "threshold or microclimate that achieved it in parentheses. In areas",
      "where canopy temperatures fall below 0 °C, canopy temperatures",
      "predict all cells to be unsuitable by definition (TSS = 0)."
    ),
    general_title = ""
  )
table_memo_single
```

Forest cover alone (TSS = `{r} paste(sprintf("%.2f", range(memo_single$forest_tss)), collapse = "–")`) and canopy temperatures alone (TSS = `{r} paste(sprintf("%.2f", range(memo_single$canopy_tss[memo_single$thresh != "below_zero"])), collapse = "–")` outside of areas below 0 °C) have a limited capacity to predict the green anole's range. Leaf-litter temperatures alone perform substantially better (TSS = `{r} paste(sprintf("%.2f", range(memo_single$temp_tss)), collapse = "–")`, at depths of `{r} paste(sort(unique(round(-memo_single$temp_z * 100))), collapse = "–")` cm), and combining them with forest cover yields the highest predictive capacity (TSS = `{r} paste(sprintf("%.2f", range(memo_single$full_tss)), collapse = "–")`). Thus, both leaf-litter temperatures and forest cover contribute to predicting the green anole's range, and leaf-litter temperatures outperform canopy temperatures even when forest cover is not considered.

### Minimum leaf-litter depth required in each cell

The envelope models above identify the microclimate and forest cover threshold that best predict the green anole's range across a spatial extent, but an optimal leaf-litter depth or forest cover threshold cannot be defined for a single cell: TSS requires both presences and absences, and, for a single cell where the green anole is present, every microclimate that keeps lizards above 0 °C predicts its presence equally well. What we can calculate for each cell independently, using only its own biophysical model estimates, is the minimum leaf-litter depth at which a green anole's minimum annual body temperature stays above a given threshold. Minimum annual body temperatures never decrease with depth in any cell, so this minimum depth is well defined. Below we map the minimum depth required to stay above 0 °C (A), above CT<sub>min</sub> (B) and above T<sub>ret</sub> (C) in each cell, where "canopy" indicates that body temperatures in the canopy already stay above the threshold (i.e., no leaf litter is needed) and "> 20 cm" that none of the leaf-litter depths we modeled is warm enough. The black line shows the edge of the green anole's range.

```{r map-required-depth}
#| code-summary: "Map the minimum leaf-litter depth required in each cell"
#| fig-align: "center"
#| fig-width: 15
#| fig-height: 5

# one row per cell with the presence of the green anole, and the range edge
# (range_edge() is defined above, when plotting the spatial extents)
map_cells <- micro_df |>
  filter(z == 1.8) |>
  select(lon, lat, presence)
map_range <- range_edge(map_cells)

# minimum depth at which minimum annual body temperatures stay above a
# threshold in each cell
required_depth <- function(threshold) {
  micro_df |>
    group_by(lon, lat) |>
    summarise(
      canopy = any(z == 1.8 & min_annual_body_temp > threshold),
      # rounded to avoid floating-point errors (e.g., 4.000000000000001 cm)
      depth = round(suppressWarnings(
        min(-z[z <= 0 & min_annual_body_temp > threshold]) * 100
      )),
      .groups = "drop"
    ) |>
    mutate(
      depth_class = case_when(
        canopy ~ "Canopy",
        is.infinite(depth) ~ "> 20 cm",
        depth <= 4 ~ paste(round(depth), "cm"),
        depth <= 6 ~ "5–6 cm",
        depth <= 10 ~ "7–10 cm",
        TRUE ~ "11–20 cm"
      )
    )
}
required_levels <- c(
  "Canopy", paste(1:4, "cm"), "5–6 cm", "7–10 cm", "11–20 cm", "> 20 cm"
)
required_colors <- setNames(
  c("forestgreen", hcl.colors(8, "Oranges", rev = TRUE)[2:8], "gray25"),
  required_levels
)
required_maps <- bind_rows(
  required_depth(0) |> mutate(threshold = "zero"),
  required_depth(ctmin_min) |> mutate(threshold = "ctmin"),
  required_depth(tret) |> mutate(threshold = "tret")
) |>
  mutate(
    depth_class = factor(depth_class, levels = required_levels),
    threshold = factor(threshold, levels = c("zero", "ctmin", "tret"))
  )

# share of cells inside and outside the range requiring > 4 cm of leaf litter
# to stay above 0 °C
required_summary <- required_maps |>
  filter(threshold == "zero") |>
  left_join(map_cells, by = c("lon", "lat")) |>
  group_by(presence) |>
  summarise(
    deep = mean(!canopy & (is.infinite(depth) | depth > 4)),
    .groups = "drop"
  )

# shared map elements: full study area, with the range edge on top
required_map <- function(df, fill_var) {
  ggplot() +
    geom_raster(data = df, aes(x = lon, y = lat, fill = .data[[fill_var]])) +
    geom_segment(
      data = map_range,
      aes(x = x, y = y, xend = xend, yend = yend),
      color = "white", linewidth = 1.2, lineend = "square"
    ) +
    geom_segment(
      data = map_range,
      aes(x = x, y = y, xend = xend, yend = yend),
      color = "black", linewidth = 0.4, lineend = "square"
    ) +
    scale_x_continuous(
      breaks = seq(-100, -80, by = 10),
      labels = function(x) paste0(abs(x), "°W"),
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      breaks = seq(25, 37.5, by = 5),
      labels = function(x) paste0(x, "°N"),
      expand = c(0, 0)
    ) +
    coord_quickmap(xlim = c(-101.8, -75), ylim = c(24, 39)) +
    theme_minimal() +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
      panel.grid = element_blank(),
      axis.title = element_blank(),
      axis.text = element_text(size = 9),
      axis.ticks = element_line(color = "black", linewidth = 0.25),
      plot.title = element_text(size = 13, hjust = 0.5),
      legend.position = "bottom",
      legend.title = element_text(size = 10, vjust = 0.8),
      legend.text = element_text(size = 9)
    )
}

required_zero_map <- required_maps |>
  filter(threshold == "zero") |>
  required_map("depth_class") +
  scale_fill_manual(name = NULL, values = required_colors, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(title = "Minimum depth to stay above 0 °C")
required_ctmin_map <- required_maps |>
  filter(threshold == "ctmin") |>
  required_map("depth_class") +
  scale_fill_manual(name = NULL, values = required_colors, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(title = expression("Minimum depth to stay above " * CT[min]))
required_tret_map <- required_maps |>
  filter(threshold == "tret") |>
  required_map("depth_class") +
  scale_fill_manual(name = NULL, values = required_colors, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(title = expression("Minimum depth to stay above " * T[ret]))
ggarrange(
  required_zero_map,
  required_ctmin_map,
  required_tret_map,
  nrow = 1,
  align = "h",
  labels = c("A", "B", "C"),
  common.legend = TRUE,
  legend = "bottom"
)
```

Within the green anole's range, `{r} round(100 * required_summary$deep[required_summary$presence == 1])`% of cells require more than 4 cm of leaf litter to keep body temperatures above 0 °C, compared with `{r} round(100 * required_summary$deep[required_summary$presence == 0])`% of cells outside the range. Thus, the poleward range edge broadly coincides with the transition from areas where a few centimeters of leaf litter are enough to avoid freezing to areas where deeper litter (or no available microclimate) would be required. In contrast, staying above CT<sub>min</sub>, and even more so above T<sub>ret</sub>, requires deep leaf litter, or is not possible in any of the microclimates we modeled, across most of the range, consistent with green anoles tolerating temperatures below their CT<sub>min</sub> but not freezing.

Lastly, we summarise these maps by latitude. For each threshold, we calculate the average minimum leaf-litter depth required across all cells at each latitude (excluding latitudes with 100 cells or fewer). Cells where none of the modeled depths is warm enough require at least 20 cm of leaf litter, so we assign them a depth of 20 cm and the resulting averages are lower bounds of the depth required. The dark gray band at the bottom of the plot marks depths beyond 20 cm; curves approaching it indicate that most cells at that latitude require more than 20 cm of leaf litter. We then compare these averages with the latitude of the green anole's northern range limit (the northernmost cell where the green anole is present in each 1° longitudinal band). Lines show a running mean of the average depth over ± 0.25° of latitude, and shaded areas a running mean of its standard deviation across cells at each latitude. The orange band marks depths of 3–4 cm. The dashed line shows the average northern range limit and the light gray area its standard deviation across longitudinal bands.

```{r plot-required-depth-by-latitude}
#| code-summary: "Plot the average minimum leaf-litter depth required by latitude"
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5
#| out.width: "75%"

# northern range limit: northernmost cell where the green anole is present in
# each 1-degree longitudinal band
max_lat_range <- micro_df |>
  select(lon, lat, presence) |>
  distinct() |>
  filter(presence == 1) |>
  mutate(lon_round = round(lon)) |>
  group_by(lon_round) |>
  filter(lat == max(lat)) |>
  ungroup()
north_range_limit <- max_lat_range |>
  summarise(
    mean = mean(lat),
    sd = sd(lat)
  )

# average minimum depth required by latitude; cells where none of the modeled
# depths is warm enough require at least 20 cm, so we set them to 20 cm and
# the averages are lower bounds
max_depth <- 20
avg_depth_by_lat <- required_maps |>
  mutate(depth = ifelse(is.finite(depth), depth, max_depth)) |>
  group_by(threshold, lat) |>
  summarise(
    n_obs = n(),
    avg_depth = mean(depth, na.rm = TRUE),
    sd_depth = sd(depth, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_obs > 100)

# running mean of the averages and standard deviations over +/- 0.25 degrees
# of latitude (unlike a loess smoother, it cannot overshoot the 0-20 cm range
# of depths)
avg_depth_smooth <- avg_depth_by_lat |>
  group_by(threshold) |>
  mutate(
    smooth_depth = map_dbl(lat, \(l) mean(avg_depth[abs(lat - l) <= 0.25])),
    smooth_sd = map_dbl(lat, \(l) mean(sd_depth[abs(lat - l) <= 0.25])),
    lower = pmax(smooth_depth - smooth_sd, 0),
    upper = pmin(smooth_depth + smooth_sd, max_depth)
  ) |>
  ungroup()

# colors and labels of each threshold (as in Fig. 4)
threshold_colors <- c(zero = "darkblue", ctmin = "steelblue", tret = "lightblue")
threshold_labels <- c(
  zero = expression(0 * degree * C),
  ctmin = expression(CT[min]),
  tret = expression(T[ret])
)

depth_lat_plot <- avg_depth_smooth |>
  ggplot(aes(x = lat, color = threshold, fill = threshold)) +
  # depths at which body temperatures stay above 0 °C at the northern range
  # limit
  annotate(
    "rect",
    xmin = -Inf, xmax = Inf, ymin = 3, ymax = 4,
    fill = "darkorange", alpha = 0.25
  ) +
  # boundary between above-ground (canopy) and below-leaf-litter microclimates
  geom_hline(yintercept = 0.5, color = "gray60", linewidth = 0.3) +
  annotate(
    "rect",
    xmin = north_range_limit$mean - north_range_limit$sd,
    xmax = north_range_limit$mean + north_range_limit$sd,
    ymin = -Inf, ymax = Inf,
    fill = "gray85", alpha = 0.6
  ) +
  # depths beyond those modeled
  annotate(
    "rect",
    xmin = -Inf, xmax = Inf, ymin = max_depth, ymax = max_depth + 1,
    fill = "gray40"
  ) +
  annotate(
    "text",
    x = -Inf, y = max_depth + 0.5, label = "> 20 cm",
    hjust = -0.1, color = "white", size = 3.5, fontface = "bold"
  ) +
  geom_vline(
    xintercept = north_range_limit$mean,
    linetype = "dashed", color = "gray30"
  ) +
  annotate(
    "text",
    x = north_range_limit$mean, y = 14,
    label = "Average northern range limit",
    angle = 90, hjust = 0, vjust = -0.5, size = 3.5, color = "gray30"
  ) +
  geom_ribbon(aes(ymin = lower, ymax = upper), color = NA, alpha = 0.2) +
  geom_line(aes(y = smooth_depth), linewidth = 1) +
  # callout: depth at which body temperatures stay above 0 °C at the northern
  # range limit
  annotate(
    "curve",
    x = north_range_limit$mean + 1.6, y = -1.3,
    xend = north_range_limit$mean + 0.08, yend = 3.4,
    curvature = 0.3, linewidth = 0.4,
    arrow = arrow(length = unit(0.18, "cm"), type = "closed")
  ) +
  annotate(
    "label",
    x = Inf, y = -1.8,
    label = "Lizard body temperature remains\n> 0 °C at 3–4 cm depth",
    hjust = 1, vjust = 0.5, size = 3.5, lineheight = 0.9,
    fill = "white", linewidth = 0
  ) +
  scale_color_manual(
    name = "Body temperature\nabove",
    values = threshold_colors,
    labels = threshold_labels
  ) +
  scale_fill_manual(
    name = "Body temperature\nabove",
    values = threshold_colors,
    labels = threshold_labels
  ) +
  scale_x_continuous(
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  ) +
  # extra room above 0 cm for the callout
  scale_y_reverse(
    breaks = seq(0, 20, by = 2),
    labels = function(x) ifelse(x == 0, "Canopy", x),
    limits = c(max_depth + 1, -3),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.02, 0.1),
    legend.justification = c(0, 0),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 12)
  ) +
  labs(
    x = "Latitude",
    y = "Average minimum depth in leaf litter (cm)"
  )
depth_lat_plot
```

## Manuscript Figure 4

Finally, we combine the plots above into Figure 4 of the manuscript. (A) Areas where minimum canopy temperatures fall below each threshold. (B) Average minimum leaf-litter depth at which green anoles can keep body temperatures above each threshold, by latitude. (C) Minimum leaf-litter depth at which green anoles can keep body temperatures above T<sub>ret</sub>, CT<sub>min</sub> and 0 °C in each cell. (D) TSS of every combination of microclimate and forest cover threshold in areas where canopy temperatures fall below T<sub>ret</sub>, CT<sub>min</sub> and 0 °C, with red circles marking the combination that maximizes TSS.

```{r plot-manuscript-figure-4}
#| code-summary: "Plot manuscript Figure 4"
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 8.83
#| out.width: "100%"
#| message: false

# Figure 4 is built at its final print size (PNAS two-column width: 17.8 cm /
# 7 in), so font sizes below are the sizes in print (PNAS minimum: 6 pt)
fig4_width <- 7
fig4_left_share <- 0.46
pt <- 1 / ggplot2::.pt # converts points to ggplot2 text size units (mm)
fig4_text <- theme(
  axis.title = element_text(size = 8),
  axis.text = element_text(size = 7),
  legend.title = element_text(size = 7),
  legend.text = element_text(size = 7),
  strip.text = element_text(size = 8),
  plot.title = element_text(size = 8)
)

# rescale text and line layers of an existing plot to print size
fig4_rescale_layers <- function(p, text_size = 6.5 * pt, line_scale = 0.6) {
  p$layers <- lapply(p$layers, \(l) {
    if (inherits(l$geom, c("GeomText", "GeomLabel"))) {
      l$aes_params$size <- text_size
    }
    if (!is.null(l$aes_params$linewidth)) {
      l$aes_params$linewidth <- l$aes_params$linewidth * line_scale
    }
    l
  })
  p
}

# thresholds shown in panels C and D, from top to bottom
fig4_thresholds <- c("tret", "ctmin", "zero")
fig4_extents <- c(tret = "below_tret", ctmin = "below_ctmin", zero = "below_zero")

# all maps share the same extent, cropped around the green anole's range; we
# use coord_cartesian (instead of coord_quickmap) so that maps fill the width
# of the column, and set the height of each row below to keep the maps'
# aspect ratio
fig4_xlim <- c(-100.2, -75.3)
fig4_ylim <- c(24.3, 37.5)
fig4_coord <- coord_cartesian(xlim = fig4_xlim, ylim = fig4_ylim, expand = FALSE)
fig4_map_scales <- list(
  scale_x_continuous(
    breaks = seq(-95, -80, by = 5),
    labels = function(x) paste0(abs(x), "°W"),
    expand = c(0, 0)
  ),
  scale_y_continuous(
    breaks = seq(26, 36, by = 4),
    labels = function(x) paste0(x, "°N"),
    expand = c(0, 0)
  )
)
fig4_map_theme <- fig4_text +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_line(linewidth = 0.25),
    axis.ticks.length = unit(1.5, "pt"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    plot.title = element_blank(),
    plot.margin = margin(t = 2, r = 4, b = 2, l = 2)
  )

# A: areas below each threshold
fig4_a <- fig4_rescale_layers(thresh_plot, line_scale = 0.5) +
  fig4_map_scales +
  fig4_coord +
  fig4_map_theme +
  theme(
    legend.position.inside = c(0.4, 0.04),
    legend.title = element_blank(),
    legend.text = element_text(size = 7),
    legend.key.size = unit(7, "pt"),
    legend.key.spacing.x = unit(4, "pt"),
    plot.margin = margin(t = 12, r = 4, b = 2, l = 2)
  )

# B: average minimum depth by latitude
fig4_b <- fig4_rescale_layers(depth_lat_plot)
# move the "Average northern range limit" label lower so it clears the 3-4 cm
# band at print size
for (k in seq_along(fig4_b$layers)) {
  l <- fig4_b$layers[[k]]
  if (inherits(l$geom, "GeomText") && isTRUE(l$aes_params$angle == 90)) {
    l$data$y <- 19
    fig4_b$layers[[k]] <- l
  }
}
fig4_b <- fig4_b +
  fig4_text +
  theme(
    axis.line = element_line(linewidth = 0.3),
    axis.ticks = element_line(linewidth = 0.3),
    legend.key.size = unit(8, "pt"),
    legend.position.inside = c(0.02, 0.08),
    plot.margin = margin(t = 12, r = 4, b = 2, l = 6)
  ) +
  labs(y = "Average minimum depth\nin leaf litter (cm)") +
  # label the surface as 0 (instead of "Canopy") to save space; the line just
  # below it still separates above- and below-ground values
  scale_y_reverse(
    breaks = seq(0, 20, by = 2),
    limits = c(max_depth + 1, -3),
    expand = c(0, 0)
  )

# C: minimum depth maps (Tret at the top, 0 °C at the bottom), labeled inside
# the Gulf of Mexico; only the bottom map shows longitude labels
fig4_c_labels <- c(
  tret = "'Body temperature' > T[ret]",
  ctmin = "'Body temperature' > CT[min]",
  zero = "'Body temperature' > 0 * degree * C"
)
fig4_c_maps <- list(
  tret = required_tret_map,
  ctmin = required_ctmin_map,
  zero = required_zero_map
)[fig4_thresholds] |>
  imap(\(p, th) {
    fig4_rescale_layers(p, line_scale = 0.5) +
      annotate(
        "text",
        x = -89.5, y = 27, label = fig4_c_labels[[th]], parse = TRUE,
        size = 9.5 * pt
      ) +
      fig4_map_scales +
      fig4_coord +
      fig4_map_theme +
      theme(legend.position = "none")
  })
fig4_c_legend <- cowplot::get_plot_component(
  required_zero_map +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
    theme(
      legend.text = element_text(size = 7),
      legend.key.size = unit(7, "pt"),
      legend.key.spacing.x = unit(3, "pt")
    ),
  "guide-box-bottom",
  return_all = TRUE
)

# D: TSS surfaces (Tret at the top, 0 °C at the bottom; entire range excluded)
fig4_d_data <- tss_grid |>
  filter(thresh %in% fig4_extents, z < 2) |>
  mutate(
    thresh = factor(as.character(thresh), levels = fig4_extents),
    # canopy shown as a thin strip above the leaf litter
    tile_height = ifelse(z > 0, 0.054, 0.01),
    z_mod = ifelse(z > 0, 0.035, z)
  ) |>
  as_tibble() |>
  arrange(desc(tile_height))
fig4_d_max <- tss_grid_max |>
  filter(thresh %in% fig4_extents) |>
  mutate(thresh = factor(as.character(thresh), levels = fig4_extents))
fig4_d <- fig4_d_data |>
  ggplot(aes(x = veg, fill = tss)) +
  geom_tile(aes(y = z_mod, height = tile_height)) +
  geom_point(
    aes(y = z),
    data = fig4_d_max,
    color = "red", size = 1.8, shape = 1, stroke = 0.6
  ) +
  geom_hline(yintercept = 0.0065, color = "white", linewidth = 0.8) +
  annotate(
    "text",
    x = 0.02, y = 0.035, label = "Canopy",
    hjust = 0, color = "white", size = 7 * pt, fontface = "bold"
  ) +
  scale_fill_viridis_b(
    name = "TSS",
    limits = c(
      min(tss_grid$tss, na.rm = TRUE), max(tss_grid$tss, na.rm = TRUE)
    ),
    breaks = seq(
      min(tss_grid$tss, na.rm = TRUE), max(tss_grid$tss, na.rm = TRUE),
      by = 0.05
    ),
    labels = function(x) ifelse(seq_along(x) %% 2 == 1, x, ""),
    guide = guide_colorsteps(
      barwidth = unit(4.5, "cm"),
      barheight = unit(0.25, "cm"),
      title.vjust = 0.9
    )
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = function(x) paste0(x * 100, "%"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = c(0, -0.05, -0.1, -0.2),
    labels = c("0", "5", "10", "20")
  ) +
  facet_wrap(
    ~thresh,
    ncol = 1,
    axes = "all_y",
    labeller = labeller(
      thresh = as_labeller(
        c(
          below_zero = "'Canopy ' * T[min] < 0 * degree * C",
          below_ctmin = "'Canopy ' * T[min] < CT[min]",
          below_tret = "'Canopy ' * T[min] < T[ret]"
        ),
        label_parsed
      )
    )
  ) +
  labs(
    x = "Forest cover threshold",
    y = "Depth in leaf litter (cm)"
  ) +
  theme_minimal() +
  fig4_text +
  theme(
    legend.position = "bottom",
    legend.margin = margin(t = 0),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.grid = element_blank(),
    panel.spacing.y = unit(4, "pt"),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    strip.text = element_text(size = 9.5, margin = margin(b = 2, t = 2)),
    plot.margin = margin(t = 2, r = 10, b = 2, l = 6)
  )

# row heights (in inches): map panels are the width of the left column minus
# the latitude labels, and their height keeps the maps' aspect ratio (as in
# coord_quickmap)
fig4_map_aspect <- diff(fig4_ylim) /
  (diff(fig4_xlim) * cos(mean(fig4_ylim) * pi / 180))
fig4_map_height <- (fig4_width * fig4_left_share - 0.4) * fig4_map_aspect
fig4_gap <- 0.12 # vertical space between maps
fig4_heights <- c(
  a = fig4_map_height + 0.4, # map + longitude labels + top margin
  c = fig4_map_height + 0.25 + fig4_gap, # map + longitude labels + gap
  legend = 0.35
)
fig4_c_maps <- map(
  fig4_c_maps,
  \(p) p + theme(plot.margin = margin(t = fig4_gap * 72, r = 4, b = 2, l = 2))
)

# left column: A and the three C maps, with panel edges aligned
fig4_left <- cowplot::plot_grid(
  fig4_a, fig4_c_maps[[1]], fig4_c_maps[[2]], fig4_c_maps[[3]],
  fig4_c_legend,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = fig4_heights[c("a", "c", "c", "c", "legend")],
  labels = c("A", "C", "", "", ""),
  label_size = 10,
  label_x = 0, label_y = c(1, 1, 1, 1, 1),
  hjust = -0.2
)
fig4_total_height <- sum(fig4_heights[c("a", "c", "c", "c", "legend")])

# right column: B taller than panel A, and D aligned with B
fig4_b_height <- fig4_heights[["a"]] + 0.6
fig4_right <- cowplot::plot_grid(
  fig4_b, fig4_d,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(fig4_b_height, fig4_total_height - fig4_b_height),
  labels = c("B", "D"),
  label_size = 10,
  label_x = 0,
  hjust = -0.2
)

fig4 <- cowplot::plot_grid(
  fig4_left, fig4_right,
  nrow = 1,
  rel_widths = c(fig4_left_share, 1 - fig4_left_share)
)
fig4
```

Figure 4 is drawn at its final print size (`{r} fig4_width` × `{r} round(fig4_total_height, 1)` in).
