---
title: "Behavioral buffering facilitates the poleward range expansion of a tropical lizard beyond its physiological limits"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

## Vignettes and analysis

- [**Green anoles live beyond their physiological cold tolerance**](live-beyond-ctmin.qmd): 

- [**Green anoles retreat to the leaf litter to avoid extreme cold**](leaf-litter-retreat.qmd):

- [**Leaf litter temperatures predict the green anole's poleward range limit**](leaf-litter-temps-range.qmd): 

## Datasets

The following datasets are used in this analysis and included in the `data` folder:

- `caro_inat_data.rds` - Raw iNaturalist occurrence data for the green anole (*Anolis carolinensis*).

- `caro_occ_data.rds` - Curated occurrence data for the green anole (*Anolis carolinensis*), with indicators for outlier locations.

- `caro_range.rds` - Polygon defining the native range of the green anole (*Anolis carolinensis*).

```{r r-setup, warning = FALSE, message = FALSE}
# packages
library(tidyverse)
library(kableExtra)
library(viridis)
library(ggpubr)
library(ggtext)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)
```

```{r plot-maps}
#| fig-align: "center"
#| fig-height: 7.5

# load data
micro_df <- readRDS("data/micro_df.rds")

# plot temperature maps
micro_df |>
  filter(as.character(z) %in% c("-0.04", "1.8")) |>
  mutate(
    z = factor(
      z,
      levels = c("1.8", "-0.04"),
      labels = c(
        "Lizard Body Temperature in <span style='color: forestgreen;'><b>Canopy</b></span>",
        "Lizard Body Temperature in <span style='color: darkorange;'><b>Leaf Litter</b></span>"
      )
    )
  ) |>
  ggplot(aes(x = lon, y = lat)) +
  geom_tile(aes(fill = min_annual_body_temp)) +
  geom_tile(
    data = micro_df |>
      select(lon, lat, presence) |>
      distinct() |> 
      filter(presence == 1),
    aes(x = lon, y = lat),
    fill = "forestgreen",
    alpha = 0.2
  ) +
  geom_contour(
    aes(x = lon, y = lat, z = presence, color = "Range"),
    linewidth = 0.5
  ) +
  geom_contour(
    aes(x = lon, y = lat, z = min_annual_body_temp, color = "0 °C"),
    breaks = c(0),
    linewidth = 1
  ) +
  geom_contour(
    aes(x = lon, y = lat, z = min_annual_body_temp, color = "CTmin"),
    breaks = c(7),
    linewidth = 1
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      "0 °C" = "steelblue4",
      "CTmin" = "skyblue",
      "Range" = "forestgreen"
    ),
    breaks = c("0 °C", "CTmin", "Range"),
    labels = expression(bold("0 °C"), bold(CT[min]), bold("Range"))
  ) +
  scale_fill_viridis(
    option = "magma",
    guide = guide_colorbar(
      title.position = "top",
      label.position = "left",
      barheight = unit(5, "cm")
    ),
    breaks = seq(-10, 25, 10),
    labels = paste0(seq(-10, 25, 10), " °C"),
  ) +
  scale_x_continuous(
    breaks = seq(-100, -70, 5),
    labels = paste0(seq(100, 70, -5), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(0, 45, 2.5),
    labels = paste0(seq(0, 45, 2.5), "°N"),
    expand = c(0, 0)
  ) +
  facet_wrap(
    ~ z,
    ncol = 1,
    axes = "all"
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 38)) +
  theme_minimal() +
  guides(
    color = guide_legend(
      label.position = "left",
      override.aes = list(linewidth = 2),
      theme = theme(
        legend.text = element_text(size = 18, face = "bold"),
        legend.key.spacing.y = unit(0.5, "cm")
      )
    )
  ) +
  theme(
    legend.position = "left",
    legend.box = "vertical",
    legend.box.just = "left",
    legend.box.spacing = unit(0.1, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.text = element_text(size = 10),
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.title = element_blank(),
    panel.border = element_rect(fill = NA, color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    strip.text = ggtext::element_markdown(size = 16, hjust = 0.5),
    plot.margin = margin(r = 10, t = -100, b = -100, l = 10)
  )
```

```{r plot-veg-map}
#| fig-align: "center"
#| fig-height: 5

# plot vegetation map
micro_df |>
  filter(z == 0) |>
  ggplot(aes(x = lon, y = lat)) +
  geom_tile(aes(fill = veg)) +
  geom_contour(
    aes(x = lon, y = lat, z = presence),
    color = "black",
    linewidth = 0.5
  ) +
  scale_fill_gradient(
    low = "lightgoldenrod1",
    high = "darkgreen",
    guide = guide_colorbar(
      title.position = "top",
      label.position = "left",
      barheight = unit(5, "cm")
    ),
    breaks = seq(0, 1, 0.2),
    labels = paste0(seq(0, 1, 0.2) * 100, "%"),
  ) +
  scale_x_continuous(
    breaks = seq(-100, -70, 5),
    labels = paste0(seq(100, 70, -5), "°W"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(0, 45, 2.5),
    labels = paste0(seq(0, 45, 2.5), "°N"),
    expand = c(0, 0)
  ) +
  coord_quickmap(xlim = c(-101, -75), ylim = c(24, 38)) +
  theme_minimal() +
  guides(
    color = guide_legend(
      label.position = "left",
      override.aes = list(linewidth = 2),
      theme = theme(
        legend.text = element_text(size = 14),
        legend.key.spacing.y = unit(0.5, "cm")
      )
    )
  ) +
  theme(
    legend.position = "left",
    legend.box = "vertical",
    legend.box.just = "left",
    legend.box.spacing = unit(0.1, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.text = element_text(size = 10),
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.title = element_blank(),
    panel.border = element_rect(fill = NA, color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    strip.text = ggtext::element_markdown(size = 16, hjust = 0.5),
    plot.title = element_text(size = 17, hjust = 0.5),
    plot.margin = margin(r = 10, t = -10, b = -10, l = 10)
  ) +
  labs(
    title = "Tree Cover"
  )
```

```{r plot-example-otm-dynamic}
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 4

# read data
otm_data <- readRDS("data/otms-data.rds")

# plot example OTM dynamic
otm_data |>
  filter(location %in% c("ATL", "MIA")) |>
  mutate(
    location = factor(
      location,
      levels = c("MIA", "ATL"),
      labels = c("Miami (25.7 °N)", "Atlanta (33.7 °N)")
    )
  ) |>
  filter(doy == 19) |>
  mutate(hour = round(mod / 60)) |>
  group_by(location, microhabitat, hour) |>
  summarise(
    mean_temp = mean(temp, na.rm = TRUE),
    sd_temp = sd(temp, na.rm = TRUE),
  ) |>
  ggplot(aes(x = hour, y = mean_temp, color = microhabitat)) +
  annotate(
    "segment",
    x = -Inf,
    xend = Inf,
    y = 0,
    yend = 0,
    linetype = 2,
    color = "steelblue4"
  ) +
  annotate(
    "text",
    x = 18,
    y = 9.5,
    label = expression(bold(CT[min])),
    color = "skyblue",
    hjust = 0,
    vjust = 0,
    size = 5
  ) +
  annotate(
    "segment",
    x = -Inf,
    xend = Inf,
    y = 9,
    yend = 9,
    linetype = 2,
    color = "skyblue"
  ) +
  geom_ribbon(
    aes(
      ymin = mean_temp - sd_temp,
      ymax = mean_temp + sd_temp,
      fill = microhabitat
    ),
    alpha = 0.2
  ) +
  scale_fill_manual(
    values = c("forestgreen", "darkorange"),
    labels = c("Canopy", "Leaf Litter"),
    name = NULL
  ) +
  scale_color_manual(
    values = c("forestgreen", "darkorange"),
    labels = c("Canopy", "Leaf Litter"),
    name = NULL
  ) +
  facet_wrap(~location, axes = "all", ncol = 2, as.table = TRUE) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(0, 24, 6),
    labels = paste0(seq(0, 24, 6), ":00"),
    expand = c(0, 0)
  ) +
  theme(
    panel.border = element_rect(fill = NA, color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    strip.text = element_text(size = 12, hjust = 0.5, face = "bold"),
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.background = element_blank(),
    legend.text = element_text(size = 14),
    panel.grid = element_blank(),
    axis.text = element_text(size = 10),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 14),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(r = 15, l = 10, b = 5)
  ) +
  labs(
    y = "Lizard Body Temperature (°C)"
  )

```