---
title: "Green anoles live beyond their physiological cold tolerance"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

```{r setup, warning = FALSE, message = FALSE}
#| code-summary: "R setup"

# packages
library(tidyverse)
library(NicheMapR)
library(kableExtra)
library(terra)
library(sf)
library(geodata)
library(elevatr)
library(ggpubr)
library(ggh4x)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)
```

## Overview

Ectotherms are often assumed to need physiological adaptation to expand their ranges poleward. If that were the case for the green anole, its cold tolerance should track the colder climates it encounters at higher latitudes. Below we test this assumption in two steps. First, we compare how the critical thermal minimum (CT<sub>min</sub>) of green anoles and minimum annual temperatures change across the species' latitudinal range. Second, we combine CT<sub>min</sub> data with the geographic ranges of 181 lizard species to quantify how much of each species' range falls below its CT<sub>min</sub> and below 0 °C, and assess whether the green anole is an exceptional case.

## CT<sub>min</sub> does not scale with climate

Below we demonstrate how the critical thermal minimum (CT<sub>min</sub>) of green anoles does not scale with climatic cooling. We do this by comparing empirical CT<sub>min</sub> data with the minimum annual temperatures across a latitudinal gradient spanning the range of the green anole. 

We collected the CT<sub>min</sub> data from 7 locations across the range of *A. carolinensis*. To measure CT<sub>min</sub>, we fitted lizards with a cloacal thermocouple probe, secured it to the base of the tail with surgical tape, and placed them in an experimental chamber submerged in an ice-water bath. Once internal body temperature stabilized after handling, we gradually cooled lizards and, starting at ~15 °C, evaluated the righting reflex at 0.2 °C intervals by placing the lizard on its dorsum and applying gentle pressure to the hind limbs. We recorded CT<sub>min</sub> as the core body temperature at which an individual completely lost the capacity to right itself. We also extracted the minimum temperature of the coldest month (BIO6) from WorldClim and obtained the average, minimum and maximum BIO6 values for each 0.1° latitude band within a bounding box defined by the locations where we measured CT<sub>min</sub> (+1° in each cardinal direction; +0.5° to the north).

The figure below (Fig. S2 in the manuscript) shows that while BIO6 decreases with latitude, CT<sub>min</sub> does not follow this trend, indicating that green anoles are living in climates colder than their physiological cold tolerance.

```{r ctmin-does-not-scale-with-climate-process-plot}
#| code-summary: "Process BIO6 data and plot CTmin and BIO6 across latitudes"
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5
#| out-width: "50%"

# read CTmin data
ctmin_gradient <- readRDS("data/ctmin_gradient.rds")

# read global min temperatures dataset
global_min_temp <- readRDS("data/global_min_temp.rds")

# define a search box for the bio6 dataset
search_box <- ext(
  min(ctmin_gradient$lon) - 1, max(ctmin_gradient$lon) + 1,
  min(ctmin_gradient$lat) - 1, max(ctmin_gradient$lat) + 0.5
)

# read in BIO6 data
bio6 <- terra::rast("raw_data/bio_6.tif")

# extract all BIO6 values and coordinates within the search box
bio6_data <- terra::crop(bio6, search_box)
bio6_df <- as.data.frame(bio6_data, xy = TRUE, na.rm = TRUE)
colnames(bio6_df) <- c("lon", "lat", "bio6")
bio6_df <- bio6_df |>
  mutate(lat = round(lat, 1)) |>
  group_by(lat) |>
  summarise(
    mean_bio6 = mean(bio6, na.rm = TRUE),
    max_bio6 = max(bio6, na.rm = TRUE),
    min_bio6 = min(bio6, na.rm = TRUE),
    .groups = "drop"
  )

# plot CTmin and BIO6 data
ggplot() +
  geom_hline(yintercept = 0, color = "steelblue4", linetype = 2) +
  geom_jitter(
    data = ctmin_gradient,
    aes(x = lat, y = ctmin, color = "CTmin"),
    width = 0.1, height = 0,
    size = 3, alpha = 0.5,
    show.legend = FALSE
  ) +
  geom_smooth(
    data = ctmin_gradient,
    aes(x = lat, y = ctmin, color = "CTmin"),
    method = "lm", fill = "forestgreen", alpha = 0.2,
    show.legend = TRUE,
    fullrange = TRUE
  ) +
  geom_line(
    data = bio6_df,
    aes(x = lat, y = mean_bio6, color = "BIO6"),
    linewidth = 1
  ) +
  geom_ribbon(
    data = bio6_df,
    aes(x = lat, ymin = min_bio6, ymax = max_bio6),
    fill = "steelblue", alpha = 0.2, show.legend = FALSE
  ) +
  scale_color_manual(
    name = NULL,
    values = c("CTmin" = "forestgreen", "BIO6" = "steelblue"),
    labels = c(
      "CTmin" = expression(CT[min]),
      "BIO6" = "Min. Temp."
    )
  ) +
  scale_x_continuous(
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(-10, 20, 5),
  ) +
  theme_classic() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.key.spacing.y = unit(0.5, "lines"),
    legend.text = element_text(size = 15),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 14)
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        linetype = c(1, 1),
        linewidth = c(2.5, 2.5),
        fill = c(NA, NA)
      )
    )
  ) +
  labs(
    x = "Latitude (°N)",
    y = "Temperature (°C)"
  )

```

Below, we perform a linear regression to test whether CT<sub>min</sub> and BIO6 scale with latitude at the same rate. We combine both types of temperature values into a single dataset and fit the following model:

$$
T_i \sim Normal(t_i, \sigma)
$$

$$
t_i = \alpha + \beta_L \cdot L_i + \beta_C \cdot C_i + \beta_{L \cdot C} \cdot L_i \cdot C_i
$$

where temperature ($T$) depends on latitude ($L$), an indicator variable ($C$) identifying the nature of the temperature values ($C = 0$ for BIO6 and $C = 1$ for CT<sub>min</sub>), and their interaction. If the interaction term ($\beta_{L \cdot C}$) is positive and significant, CT<sub>min</sub> scales with latitude at a significantly shallower rate than BIO6. The results of this model are presented in Table S2 of the manuscript.

```{r ctmin-does-not-scale-with-climate-stats}
#| code-summary: "Linear model of CTmin and BIO6 across latitudes"

# prepare data
bio6_df_model <- bio6_df |>
  select(lat, mean_bio6) |>
  rename(temp = mean_bio6) |>
  mutate(type = "bio6")
ctmin_model <- ctmin_gradient |>
  select(lat, ctmin) |>
  rename(temp = ctmin) |>
  mutate(type = "ctmin")
combined_df <- bind_rows(bio6_df_model, ctmin_model)

# fit linear model with interaction between latitude and type
lat_temp_model <- lm(temp ~ lat * type, data = combined_df)
lat_temp_fit <- broom::glance(lat_temp_model)

# table model coefficients
broom::tidy(lat_temp_model) |>
  mutate(
    term = c(
      "Intercept (&alpha;)",
      "Latitude (&beta;<sub>L</sub>)",
      "CT<sub>min</sub> (&beta;<sub>C</sub>)",
      "Latitude &times; CT<sub>min</sub> (&beta;<sub>L&middot;C</sub>)"
    ),
    across(c(estimate, std.error, statistic), ~ round(.x, 3)),
    p.value = ifelse(
      p.value < 0.0001, "< 0.0001", as.character(round(p.value, 4))
    )
  ) |>
  kbl(
    col.names = c("Term", "Estimate", "SE", "t", "p"),
    escape = FALSE,
    align = "lcccc"
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  ) |>
  footnote(
    general = paste0(
      "N = ", lat_temp_fit$nobs,
      "; R<sup>2</sup> = ", round(lat_temp_fit$r.squared, 3),
      "; residual SE = ", round(lat_temp_fit$sigma, 3)
    ),
    general_title = "",
    escape = FALSE
  )
```

The interaction between latitude and type is positive ($\beta_{L \cdot C}$ = `{r} round(lat_temp_model$coefficients["lat:typectmin"], 3)`) and significant (p `{r} ifelse(summary(lat_temp_model)$coefficients["lat:typectmin", "Pr(>|t|)"] < 0.0001, "< 0.0001", paste("=", round(summary(lat_temp_model)$coefficients["lat:typectmin", "Pr(>|t|)"], 4)))`), indicating that CT<sub>min</sub> does not decrease with latitude as BIO6 does. This suggests that green anoles are living in climates colder than their physiological cold tolerance.

Below we table the CT<sub>min</sub> estimates (mean and standard deviation) and BIO6 values (mean, minimum and maximum across the 0.1° latitude band) for each location where we measured CT<sub>min</sub> (Table S1 in the manuscript).

```{r table-ctmin-by-latitude}
#| code-summary: "Table CTmin and BIO6 by location"

# get average temps by location
bio6_df_table <- bio6_df |>
  filter(lat %in% round(unique(ctmin_gradient$lat), 1)) |>
  mutate(
    site = case_when(
      lat == 33.7 ~ "Atlanta, GA",
      lat == 35.9 ~ "Knoxville, TN",
      lat == 29.7 ~ "Gainesville, FL",
      lat == 31.5 ~ "Tifton, GA",
      lat == 27.3 ~ "Port Saint Lucie, FL",
      lat == 28.5 ~ "Orlando, FL",
      lat == 25.8 ~ "Miami, FL"
    )
  ) |>
  select(site, mean_bio6, min_bio6, max_bio6)


ctmin_gradient |>
  group_by(site, lat) |>
  summarise(
    mean_ctmin = mean(ctmin, na.rm = TRUE),
    sd_ctmin = sd(ctmin, na.rm = TRUE),
    n = n(),
  ) |>
  mutate(
    site = case_when(
      site == "ATL" ~ "Atlanta, GA",
      site == "KNX" ~ "Knoxville, TN",
      site == "GNV" ~ "Gainesville, FL",
      site == "TIF" ~ "Tifton, GA",
      site == "PSL" ~ "Port Saint Lucie, FL",
      site == "ORL" ~ "Orlando, FL",
      site == "MIA" ~ "Miami, FL"
    )
  ) |>
  ungroup() |>
  left_join(bio6_df_table, by = c("site")) |>
  mutate(
    across(
      c(mean_ctmin, sd_ctmin, mean_bio6, min_bio6, max_bio6),
      ~ round(.x, 2)
    ),
    lat = round(lat, 2)
  ) |>
  select(
    site, lat, n, mean_ctmin, sd_ctmin, mean_bio6, min_bio6, max_bio6
  ) |>
  kbl(
    col.names = c(
      "Site", "Latitude (°N)", "N", "Mean CT<sub>min</sub> (°C)",
      "SD CT<sub>min</sub> (°C)", "Mean BIO6 (°C)", "Min BIO6 (°C)", "Max BIO6 (°C)"
    ),
    escape = FALSE,
    align = "c"
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

## The green anole is an exceptional case of climatic overfilling

### Data processing

We compiled a dataset of CT<sub>min</sub> and elevational limits for 181 lizard species, representing 17 families and 46 genera. We obtained CT<sub>min</sub> estimates from the GlobTherm database and complemented them with two datasets focused specifically on *Anolis* lizards (Bodensteiner et al. 2024; Salazar et al. 2025). We then obtained elevational range limits, substrate use (arboreal, semi-arboreal or other) and the identifiers linking each species to its range map in the GARD database from SquamBase. The script used to compile this dataset (`process-ctmin-lizards-data.R`) is stored in the `raw_data` folder. For *A. carolinensis*, we set CT<sub>min</sub> to 6.8 °C, the lowest population-average CT<sub>min</sub> we measured (Knoxville, TN).

Below we present the structure of the resulting dataset, which is stored in the `data` folder:

```{r load-ctmin-lizards}
#| code-fold: false

ctmin_lizards <- readRDS("data/ctmin_lizards.rds")
glimpse(ctmin_lizards)
```

Below we combine data on the CT<sub>min</sub> and elevational limits of globally distributed lizards to find the total area and proportion of each species' range that is below their CT<sub>min</sub> and below 0 °C. We also calculate the maximum latitude at which each species can occur above their CT<sub>min</sub> and above 0 °C.

For each species, we:

1. Crop and mask global BIO6 and elevation rasters (2.5' resolution, WorldClim) to the species' range polygon from the GARD database.

2. Keep only cells whose elevation falls within the species' elevational limits.

3. Convert each cell to its area in km<sup>2</sup> (`deg_to_area()`), accounting for the fact that cells shrink with latitude because the Earth is an ellipsoid.

4. Sum the area of all cells, of cells where BIO6 < CT<sub>min</sub> and of cells where BIO6 < 0 °C, and divide the latter two by the total area to obtain proportions.

Lastly, we calculate the global mean (± SD) BIO6 for every degree of absolute latitude, which we use as a reference in the plots below.

```{r extract-data, eval = FALSE}
#| code-summary: "Calculate the area and proportion of each species' range below CTmin and 0 °C"

## Note: This code chunk is not evaluated as it takes a long time to run.
## The same code can be run locally to reproduce the results and the output
## is stored in the `data/lizard_climatic_overfilling.rds` file.

# read in global BIO6 tif
bio6 <- terra::rast("raw_data/bio_6.tif")
elev <- terra::rast("raw_data/elev.tif")
ranges <- readRDS("raw_data/ranges_vector.rds")

# filter ranges to only include lizard species with CTmin data
ctmin_lizards_ranges <- ranges[ranges$TaxonID %in% ctmin_lizards$gard_id, ]

# define function to convert lat-lon degree cell to km^2
deg_to_area <- function(lat, delta_lat = 2.5 / 60, delta_lon = 2.5 / 60) {
  
  # unit conversions
  lat_rad <- lat * pi / 180 # convert to radians

  # define constants
  equat_radius <- 6378.1370 # equatorial radius
  polar_radius <- 6356.7523142 # polar radius
  flat_factor <- 1 / 298.257223563 # flattening factor

  # radius curvature in the meridian
  e_sq <- (equat_radius^2 - polar_radius^2) / equat_radius^2
  denominator <- (1 - e_sq * sin(lat_rad)^2)^(1.5)
  m <- (equat_radius * (1 - e_sq)) / denominator

  # set length of 1 degree latitude in km
  km_per_degree_lat <- m * (pi / 180)
  lat_dist <- km_per_degree_lat * delta_lat

  # Calculate distance of 1 degree longitude along the parallel
  dist_km <- (pi / 180) * equat_radius * cos(lat_rad) / sqrt(1 - (2 * flat_factor - flat_factor^2) * sin(lat_rad)^2) # nolint
  lon_dist <- delta_lon * dist_km

  # Return cell area in km^2
  lat_dist * lon_dist
}

# define results list to extract data
results_list <- vector("list", nrow(ctmin_lizards))

for (i in seq_along(ctmin_lizards$gard_id)) {

  # get species range and ctmin
  range <- ctmin_lizards_ranges[ctmin_lizards_ranges$TaxonID == ctmin_lizards$gard_id[i], ] # nolint
  ctmin <- ctmin_lizards$ctmin[i]
  min_elev <- ctmin_lizards$min_elev[i]
  max_elev <- ctmin_lizards$max_elev[i]

  # crop BIO6 and elevation rasters to species range
  bio6_crop <- terra::crop(terra::mask(bio6, range), range)
  elev_crop <- terra::crop(terra::mask(elev, range), range)

  # filter all areas within range and within elevation limits
  valid_cells <- which(
    !is.na(bio6_crop[]) &
      !is.na(elev_crop[]) &
      elev_crop[] >= min_elev &
      elev_crop[] <= max_elev
  )

  # filter valid cells in BIO6 raster
  bio6_crop_filtered <- bio6_crop[valid_cells]
  bio6_crop_filtered <- bio6_crop_filtered$bio_6

  # extract coordinates of valid cells
  coords <- terra::xyFromCell(bio6_crop, valid_cells)[, "y"]
  dist <- deg_to_area(abs(coords))

  # get metrics
  total_area <- sum(dist)
  below_ctmin_area <- sum(dist[bio6_crop_filtered < ctmin])
  below_0_area <- sum(dist[bio6_crop_filtered < 0])
  below_ctmin_prop <- below_ctmin_area / total_area
  below_0_prop <- below_0_area / total_area
  max_lat_above_ctmin <- ifelse(
    below_ctmin_area %in% c(0, 1),
    NA,
    max(abs(coords[bio6_crop_filtered >= ctmin]), na.rm = TRUE)
  )
  max_lat_above_0 <- ifelse(
    below_0_area %in% c(0, 1),
    NA,
    max(abs(coords[bio6_crop_filtered >= 0]), na.rm = TRUE)
  )

  # store results in list
  results_list[[i]] <- data.frame(
    family = ctmin_lizards$family[i],
    genus = ctmin_lizards$genus[i],
    species = ctmin_lizards$species[i],
    ctmin = ctmin,
    substrate = ctmin_lizards$substrate_type[i],
    min_lat = min(abs(terra::ext(range)[3:4])),
    max_lat = max(abs(terra::ext(range)[3:4])),
    total_area = total_area,
    below_ctmin_area = below_ctmin_area,
    below_0_area = below_0_area,
    below_ctmin_prop = below_ctmin_prop,
    below_0_prop = below_0_prop,
    max_lat_above_ctmin = max_lat_above_ctmin,
    max_lat_above_0 = max_lat_above_0
  )

  # clear up memory
  rm(
    range, ctmin, min_elev, max_elev, bio6_crop, elev_crop,
    valid_cells, bio6_crop_filtered, coords, dist,
    total_area, below_ctmin_area, below_0_area, below_ctmin_prop,
    below_0_prop, max_lat_above_ctmin, max_lat_above_0
  )
  terra::tmpFiles(remove = TRUE)
  gc()

  # print progress
  cat("Processed species", i, "of", nrow(ctmin_lizards), "\n")

}

# bind rows for results and store data
lizard_climatic_overfilling <- bind_rows(results_list)

# change any NaN or -Inf values to NA across all columns
lizard_climatic_overfilling <- lizard_climatic_overfilling |>
  mutate(across(
    where(is.numeric),
    ~ ifelse(is.nan(.x) | is.infinite(.x), NA, .x)
  ))

# calculate global bio6 mean, min and max for every absolute lat degree
bio6_global <- as.data.frame(bio6, xy = TRUE, na.rm = TRUE) |>
  mutate(lat = round(abs(y), 0)) |>
  group_by(lat) |>
  summarise(
    mean_bio6 = mean(bio_6, na.rm = TRUE),
    min_bio6 = mean_bio6 - sd(bio_6, na.rm = TRUE),
    max_bio6 = mean_bio6 + sd(bio_6, na.rm = TRUE),
    .groups = "drop"
  )

saveRDS(lizard_climatic_overfilling, "data/lizard_climatic_overfilling.rds")
saveRDS(bio6_global, "data/bio6_global.rds")
```

### Statistical analysis

To assess whether *A. carolinensis* is a unique case, we conducted an outlier analysis. Below we calculate Z-scores (i.e., the number of standard deviations above or below the mean of a group) for the area and proportion of the green anole's range below CT<sub>min</sub> and below 0 °C, relative to three groups: other *Anolis*, exclusively arboreal lizards, and all lizards in our dataset (Table S4 in the manuscript).

```{r z-score-analysis}
#| code-summary: "Z-scores of A. carolinensis relative to other lizards"

# load datasets
lizard_climatic_overfilling <- readRDS("data/lizard_climatic_overfilling.rds")

# define function to get a Z-score for an observation relative to a group
get_z_score <- function(x, group) {
  (x - mean(group, na.rm = TRUE)) / sd(group, na.rm = TRUE)
}

# get Z-scores of A. carolinensis with respect to other groups
caro_z_scores <- expand_grid(
  var = c(
    "below_ctmin_prop",
    "below_0_prop",
    "below_ctmin_area",
    "below_0_area"
  ),
  group = c("anolis", "arboreal", "all")
) |>
  mutate(
    z_score = map2_dbl(var, group, \(v, g) {
      dat <- lizard_climatic_overfilling[[v]]
      comp <- switch(
        g,
        anolis = dat[lizard_climatic_overfilling$genus == "Anolis"],
        arboreal = dat[lizard_climatic_overfilling$substrate == "arboreal"],
        all = dat
      )
      x <- dat[lizard_climatic_overfilling$species == "Anolis carolinensis"]
      get_z_score(x = x, group = comp)
    })
  )

# tidy up for display as a table
caro_z_scores |>
  pivot_wider(names_from = group, values_from = z_score) |>
  mutate(
    across(
      c(anolis, arboreal, all),
      ~ round(.x, 2)
    )
  ) |>
  mutate(
    metric = case_when(
      var == "below_ctmin_prop" ~ "Proportion of range below CT<sub>min</sub>",
      var == "below_0_prop" ~ "Proportion of range below 0 °C",
      var == "below_ctmin_area" ~ "Area of range below CT<sub>min</sub> (km<sup>2</sup>)",
      var == "below_0_area" ~ "Area of range below 0 °C (km<sup>2</sup>)"
    )
  ) |>
  select(metric, anolis, arboreal, all) |>
  kbl(
    col.names = c(
      "Metric", "Z-score (relative to <i>Anolis</i> genus)",
      "Z-score (relative to arboreal lizards)",
      "Z-score (relative to all lizards)"
    ),
    escape = FALSE,
    align = "lccc"
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

We find that *A. carolinensis* has a much greater proportion and area of its range below CT<sub>min</sub> and below 0 °C than other *Anolis* and other arboreal lizards, indicating that it is an exceptional case of climatic overfilling. Relative to all lizards, the green anole's Z-scores are close to 0, as its overfilling resembles that of terrestrial, fossorial or saxicolous species living in colder climates, which can shelter from the cold in underground refuges.

### Plots

First, we plot the latitudinal distribution of CT<sub>min</sub> and BIO6 for all lizard species in our dataset (Fig. 2 in the manuscript). In panel A, each segment represents the latitudinal range of a species, placed at the height of its CT<sub>min</sub>, and the blue line and area represent the global mean (± SD) BIO6 at each latitude. Panels B and C show the proportion and area (in log scale) of each species' range below CT<sub>min</sub> and below 0 °C. We highlight *A. carolinensis* and other *Anolis* species.

```{r plot-lizard-climatic-overfilling-taxonomically}
#| code-summary: "Plot climatic overfilling by taxonomic group"
#| fig-align: "center"
#| fig-width: 9
#| fig-height: 5

# load datasets
bio6_global <- readRDS("data/bio6_global.rds")
caro_ctmin_gradient <- readRDS("data/ctmin_gradient.rds")

# create data for A. carolinensis
caro_ctmin_pred <- tibble(
  lat = seq(22.75, 36.5, by = 0.1),
  ctmin = 6.8,
  substrate = "arboreal"
)

# filter taxonomic data for plotting
lizard_climatic_overfilling <- lizard_climatic_overfilling |>
  mutate(
    group = case_when(
      genus == "Anolis" & species != "Anolis carolinensis" ~ "Anolis Genus",
      species == "Anolis carolinensis" ~ "Anolis carolinensis",
      TRUE ~ "Other Lizards"
    )
  )

# plot lat ~ temp
lat_temps <- ggplot() +
  geom_hline(
    yintercept = 0, color = "darkblue", linewidth = 0.5, alpha = 0.5,
    linetype = "dashed"
  ) +
  geom_ribbon(
    data = bio6_global,
    aes(x = lat, ymin = min_bio6, ymax = max_bio6),
    fill = "steelblue", alpha = 0.2, show.legend = FALSE
  ) +
  geom_line(
    data = bio6_global,
    aes(x = lat, y = mean_bio6, color = "Global BIO6"),
    linewidth = 1
  ) +
  geom_segment(
    data = lizard_climatic_overfilling |> filter(group != "Anolis carolinensis"), # nolint
    aes(
      x = min_lat, xend = max_lat,
      y = ctmin, yend = ctmin,
      color = group
    ),
    linewidth = 1.25, alpha = 0.5
  ) +
  geom_line(
    data = caro_ctmin_pred,
    aes(x = lat, y = ctmin, color = "Anolis carolinensis"),
    linewidth = 1.5
  ) +
  scale_color_manual(
    name = NULL,
    breaks = c(
      "Anolis carolinensis",
      "Anolis Genus",
      "Other Lizards",
      "Global BIO6"
    ),
    values = c(
      "Anolis carolinensis" = "forestgreen",
      "Anolis Genus" = "green3",
      "Other Lizards" = "darkgray",
      "Global BIO6" = "steelblue"
    ),
    labels = c(
      "Anolis carolinensis" = expression(italic("A. carolinensis")),
      "Anolis Genus" = expression("Other " * italic("Anolis")),
      "Other Lizards" = "Other Lizards",
      "Global BIO6" = "Global Min. Temp."
    )
  ) +
  scale_y_continuous(
    breaks = seq(-10, 20, 5),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    breaks = seq(0, 70, 10),
    expand = c(0, 0)
  ) +
  coord_cartesian(
    xlim = c(0, max(lizard_climatic_overfilling$max_lat, na.rm = TRUE) + 1),
    ylim = c(
      -2, max(lizard_climatic_overfilling$ctmin, na.rm = TRUE) + 2
    )
  ) +
  theme_classic() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.background = element_rect(
      fill = "white", color = "gray", linewidth = 0.5
    ),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 12)
  ) +
  labs(
    x = "Absolute Latitude (°)",
    y = "Physiological cold tolerance (°C)"
  )

# plot range proportions/areas
range_data <- lizard_climatic_overfilling |>
  dplyr::select(
    group, below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area + 1, # add 1 to avoid log10(0) issues
    below_0_area = below_0_area + 1,
    group = factor(
      group,
      levels = c("Other Lizards", "Anolis Genus", "Anolis carolinensis")
    )
  ) |>
  arrange(group) |>
  pivot_longer(
    cols = c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area),
    names_to = "metric", values_to = "value"
  ) |>
  mutate(
    area_prop = case_when(
      metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
      metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
    ),
    temp = case_when(
      metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
      metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
    )
  ) |>
  arrange(group)

proportions <- range_data |>
  filter(area_prop == "Proportion") |>
  ggplot(
    aes(x = temp, y = value, color = group, alpha = group, size = group)
  ) +
  geom_jitter(width = 0.1, height = 0, show.legend = FALSE) +
  facet_wrap(
    ~area_prop,
    scales = "free",
    strip.position = "left",
    labeller = as_labeller(
      c(Proportion = "Proportion~of~range"),
      default = label_parsed
    )
  ) +
  scale_y_continuous() +
  scale_color_manual(
    values = c(
      "Anolis carolinensis" = "darkgreen",
      "Anolis Genus" = "palegreen3",
      "Other Lizards" = "gray"
    ),
    name = "Group"
  ) +
  scale_alpha_manual(
    values = c(0.5, 0.75, 1),
    name = "Group"
  ) +
  scale_size_manual(
    values = c(2, 2.5, 4),
  ) +
  scale_x_discrete(
    labels = c("CT[min]" = expression("< " * "CT"[min]),
    "0°C" = "< 0°C")
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90, margin = margin(r = 0)),
    strip.switch.pad.grid = grid::unit(0, "pt"),
    strip.text = element_text(size = 12)
  )

areas <- range_data |>
  filter(area_prop == "Area") |>
  ggplot(
    aes(x = temp, y = value, color = group, alpha = group, size = group)
  ) +
  geom_jitter(width = 0.1, height = 0, show.legend = FALSE) +
  facet_wrap(
    ~area_prop,
    scales = "free",
    strip.position = "left",
    labeller = as_labeller(
      c(Area = "Area~of~range~(km^2)"),
      default = label_parsed
    )
  ) +
  scale_y_continuous(
    trans = "log10", 
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000),
    labels = scales::label_log()
  ) +
  scale_color_manual(
    values = c(
      "Anolis carolinensis" = "darkgreen",
      "Anolis Genus" = "palegreen3",
      "Other Lizards" = "gray"
    ),
    name = "Group"
  ) +
  scale_alpha_manual(
    values = c(0.5, 0.75, 1),
    name = "Group"
  ) +
  scale_size_manual(
    values = c(2, 2.5, 4),
  ) +
  scale_x_discrete(
    labels = c(
      "CT[min]" = expression("< " * "CT"[min]),
      "0°C" = "< 0°C"
    )
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90, margin = margin(r = 0)),
    strip.switch.pad.grid = grid::unit(0, "pt"),
    strip.text = element_text(size = 12)
  )

combined_plot <- ggarrange(
  lat_temps, proportions, areas,
  ncol = 3, nrow = 1,
  widths = c(1.25, 0.5, 0.5),
  labels = c("A", "B", "C"),
  font.label = list(size = 15, face = "bold", color = "black")
)
combined_plot

```

Next we plot by substrate type to see if there are any differences in the latitudinal distribution of CT<sub>min</sub> and BIO6 across different ecological groups of lizards (Fig. S3 in the manuscript). We also highlight the latitudinal distribution of CT<sub>min</sub> for *A. carolinensis* and compare it to the latitudinal distribution of BIO6 across the globe. In panels B and C, points with error bars represent the mean (± SE) of each substrate group, and the large dark green point represents *A. carolinensis*.

Only a few other arboreal species have ranges that extend into areas below their CT<sub>min</sub> (*Liolaemus tenuis*, four *Chamaeleo* and four *Anolis* species), and only three extend into areas below 0 °C (*L. tenuis*, *Chamaeleo hoehnelii* and *C. schubotzi*). In all cases, the area affected is orders of magnitude smaller than for the green anole. Further, although SquamBase classifies *L. tenuis* as exclusively arboreal, this species adopts a saxicolous/terrestrial strategy in many parts of its range.

```{r plot-by-substrate}
#| code-summary: "Plot climatic overfilling by substrate type"
#| fig-align: "center"
#| fig-width: 9
#| fig-height: 5

# get substrate data and colors
substrate_data <- lizard_climatic_overfilling |>
  mutate(
    substrate = replace_na(as.character(substrate), "Unknown"),
    substrate = fct_infreq(substrate)
  )

substrate_levels <- levels(substrate_data$substrate)
substrate_colors <- setNames(
  c("peachpuff3", "khaki3", "springgreen3"), substrate_levels
)

lat_temps_substrate <- ggplot() +
  geom_hline(
    yintercept = 0, color = "darkblue", linewidth = 0.5, alpha = 0.5,
    linetype = "dashed"
  ) +
  geom_ribbon(
    data = bio6_global,
    aes(x = lat, ymin = min_bio6, ymax = max_bio6),
    fill = "steelblue", alpha = 0.2, show.legend = FALSE
  ) +
  geom_line(
    data = bio6_global,
    aes(x = lat, y = mean_bio6, color = "Global BIO6"), linewidth = 1
  ) +
  geom_segment(
    data = substrate_data |>
      filter(species != "Anolis carolinensis"),
    aes(
      x = min_lat, xend = max_lat, y = ctmin, yend = ctmin,
      color = substrate
    ),
    linewidth = 1.25, alpha = 0.5
  ) +
  geom_line(
    data = caro_ctmin_pred,
    aes(x = lat, y = ctmin, color = "Anolis carolinensis"),
    linewidth = 1.5
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      substrate_colors,
      "Anolis carolinensis" = "forestgreen",
      "Global BIO6" = "steelblue"
    ),
    breaks = c(substrate_levels, "Anolis carolinensis", "Global BIO6"),
    labels = c(
      as.list(setNames(substrate_levels, substrate_levels)),
      list(
        "other" = "Not Arboreal Lizards",
        "semi-arboreal" = "Semi-Arboreal Lizards",
        "arboreal" = "Arboreal Lizards",
        "Anolis carolinensis" = expression(italic("A. carolinensis")),
        "Global BIO6" = "Global Min. Temp."
      )
    )
  ) +
  scale_y_continuous(
    breaks = seq(-10, 20, 5),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    breaks = seq(0, 70, 10),
    expand = c(0, 0)
  ) +
  coord_cartesian(
    xlim = c(0, max(substrate_data$max_lat, na.rm = TRUE) + 1),
    ylim = c(-2, max(substrate_data$ctmin, na.rm = TRUE) + 2)
  ) +
  theme_classic() +
  theme(
    legend.position = "inside", legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.background = element_rect(
      fill = "white", color = "gray", linewidth = 0.5
    ),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 12),
    legend.title = element_blank()
  ) +
  labs(
    x = "Absolute Latitude (°)",
    y = "Physiological cold tolerance (°C)"
  )

make_substrate_plot <- function(data, metric_type, y_label) {
  data |>
    filter(area_prop == metric_type) |>
    ggplot(aes(x = temp, y = value, color = substrate)) +
    geom_point(
      position = position_jitterdodge(
        dodge.width = 1,
        jitter.width = 0.2,
        jitter.height = 0
      ),
      show.legend = FALSE,
      alpha = 0.25
    ) +
    stat_summary(position = position_dodge(width = 1)) +
    geom_point(
      data = substrate_data |>
        filter(species == "Anolis carolinensis") |>
        mutate(
          below_ctmin_area = below_ctmin_area + 1,
          below_0_area = below_0_area + 1,
          substrate = factor(substrate, levels = substrate_levels)
        ) |>
        pivot_longer(
          cols = c(below_ctmin_prop, below_0_prop,
            below_ctmin_area, below_0_area),
          names_to = "metric",
          values_to = "value"
        ) |>
        mutate(
          area_prop = case_when(
            metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
            metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
          ),
          temp = case_when(
            metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
            metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
          )
        ) |>
        filter(area_prop == metric_type),
      aes(x = temp, y = value, group = substrate),
      position = position_nudge(x = 1 / 3),
      color = "forestgreen",
      size = 4,
      show.legend = FALSE
    ) +
    geom_vline(xintercept = 1.5, color = "gray70", linewidth = 0.2) +
    scale_color_manual(values = substrate_colors) +
    scale_x_discrete(
      labels = c(
        "CT[min]" = expression("< " * "CT"[min]),
        "0°C" = "< 0°C"
      )
    ) +
    theme_minimal() +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
      axis.ticks = element_line(color = "black", linewidth = 0.25),
      axis.text.x = element_text(size = 12, color = "black"),
      axis.text.y = element_text(color = "black"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 12, color = "black"),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      strip.text = element_blank()
    ) +
    labs(
      y = y_label,
      subtitle = NULL
    )
}

proportions_substrate <- substrate_data |>
  select(
    substrate, below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area + 1,
    below_0_area = below_0_area + 1,
    substrate = factor(substrate, levels = substrate_levels)
  ) |>
  pivot_longer(
    cols = c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    area_prop = case_when(
      metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
      metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
    ),
    temp = case_when(
      metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
      metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
    )
  ) |>
  make_substrate_plot("Proportion", "Proportion of range") +
  scale_y_continuous()

areas_substrate <- substrate_data |>
  select(
    substrate, below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area + 1,
    below_0_area = below_0_area + 1,
    substrate = factor(substrate, levels = substrate_levels)
  ) |>
  pivot_longer(
    cols = c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    area_prop = case_when(
      metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
      metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
    ),
    temp = case_when(
      metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
      metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
    )
  ) |>
  make_substrate_plot("Area", expression("Area of range (km"^2*")")) +
  scale_y_continuous(
    trans = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000),
    labels = scales::label_log()
  )

ggarrange(
  lat_temps_substrate, proportions_substrate, areas_substrate,
  ncol = 3, nrow = 1,
  widths = c(1.25, 0.5, 0.5),
  labels = c("A", "B", "C"),
  align = "hv",
  font.label = list(size = 15, face = "bold", color = "black")
)
```

## Tables

Below we table the "overfilling" results of all 181 lizard species. All areas are expressed in millions of km<sup>2</sup>.

::: {.panel-tabset}

### By family and genus

We table the average "overfilling" results by family and genus.

```{r lizard-climatic-overfilling-table-family-genus}
#| code-summary: "Table overfilling results by family and genus"

lizard_climatic_overfilling |>
  mutate(
    genus = gsub(" .*", "", species),
    species = gsub(".* ", "", species)
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area / 1e6,
    below_0_area = below_0_area / 1e6,
    total_area = total_area / 1e6
  ) |>
  group_by(family, genus) |>
  summarise(
    below_ctmin_prop = mean(below_ctmin_prop, na.rm = TRUE),
    below_0_prop = mean(below_0_prop, na.rm = TRUE),
    below_ctmin_area = mean(below_ctmin_area, na.rm = TRUE),
    below_0_area = mean(below_0_area, na.rm = TRUE),
    total_area = mean(total_area, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # round everything to 3 decimal digits
  mutate(
    across(
      c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area, total_area),
      ~ round(.x, 3)
    )
  ) |>
  arrange(family, genus) |>
  mutate(genus = paste0("<i>", genus, "</i>")) |>
  kbl(
    col.names = c(
      "Family", "Genus",
      "Proportion < CT<sub>min</sub>",
      "Proportion < 0 °C", "Area < CT<sub>min</sub> (10<sup>6</sup> km<sup>2</sup>)",
      "Area < 0 °C (10<sup>6</sup> km<sup>2</sup>)", "Total area (10<sup>6</sup> km<sup>2</sup>)"
    ),
    escape = FALSE
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  ) |>
  scroll_box(height = "500px")
```

### By species

We table the "overfilling" results for every species (Table S3 in the manuscript).

```{r lizard-climatic-overfilling-table}
#| code-summary: "Table overfilling results by species"

lizard_climatic_overfilling |>
  mutate(
    genus = gsub(" .*", "", species),
    species = gsub(".* ", "", species)
  ) |>
  mutate(
    below_ctmin_prop = round(below_ctmin_prop, 2),
    below_0_prop = round(below_0_prop, 2),
    below_ctmin_area = round(below_ctmin_area / 1e6, 4),
    below_0_area = round(below_0_area / 1e6, 4),
    total_area = round(total_area / 1e6, 4)
  ) |>
  select(
    family, genus, species,
    below_ctmin_prop, below_0_prop,
    below_ctmin_area, below_0_area, total_area
  ) |>
  arrange(genus, species) |>
  mutate(
    genus = paste0("<i>", genus, "</i>"),
    species = paste0("<i>", species, "</i>")
  ) |>
  kbl(
    col.names = c(
      "Family", "Genus", "Species",
      "Proportion < CT<sub>min</sub>",
      "Proportion < 0 °C", "Area < CT<sub>min</sub> (10<sup>6</sup> km<sup>2</sup>)",
      "Area < 0 °C (10<sup>6</sup> km<sup>2</sup>)", "Total area (10<sup>6</sup> km<sup>2</sup>)"
    ),
    escape = FALSE
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  ) |>
  scroll_box(height = "500px")
```

:::
