---
title: "Green anoles live beyond their physiological cold tolerance"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

```{r setup, warning = FALSE, message = FALSE}
# packages
library(tidyverse)
library(NicheMapR)
library(kableExtra)
library(terra)
library(sf)
library(geodata)
library(elevatr)
library(ggpubr)
library(ggh4x)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)
```

## Overview

## CT<sub>min</sub> does not scale with climate

Below we demonstrate how the critical thermal minimum (CT<sub>min</sub>) of green anoles does not scale with climatic cooling. We do this by comparing empirical CT<sub>min</sub> data with the minimum annual temperatures across a latitudinal gradient spanning the range of the green anole. 

We collected the CT<sub>min</sub> data from 7 locations across the range of *A. carolinensis*. We measured CT<sub>min</sub> by {Insert protocol details}. We also extracted the minimum temperature of the coldest month (BIO6) from WorldClim and obtained the average, minimum and maximum BIO6 values for each 0.1° latitude band across the range of the green anole.

The figure below shows that while BIO6 decreases with latitude, CT<sub>min</sub> does not follow this trend, indicating that green anoles are living in climates colder than their physiological cold tolerance.

```{r ctmin-does-not-scale-with-climate-process-plot}
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5
#| out-width: "50%"

# read CTmin data
ctmin_gradient <- readRDS("data/ctmin_gradient.rds")

# read global min temperatures dataset
global_min_temp <- readRDS("data/global_min_temp.rds")

# define a search box for the bio6 dataset
search_box <- ext(
  min(ctmin_gradient$lon) - 1, max(ctmin_gradient$lon) + 1,
  min(ctmin_gradient$lat) - 1, max(ctmin_gradient$lat) + 0.5
)

# read in BIO6 data
bio6 <- terra::rast("raw_data/bio_6.tif")

# extract all BIO6 values and coordinates within the search box
bio6_data <- terra::crop(bio6, search_box)
bio6_df <- as.data.frame(bio6_data, xy = TRUE, na.rm = TRUE)
colnames(bio6_df) <- c("lon", "lat", "bio6")
bio6_df <- bio6_df |>
  mutate(lat = round(lat, 1)) |>
  group_by(lat) |>
  summarise(
    mean_bio6 = mean(bio6, na.rm = TRUE),
    max_bio6 = max(bio6, na.rm = TRUE),
    min_bio6 = min(bio6, na.rm = TRUE),
    .groups = "drop"
  )

# plot CTmin and BIO6 data
ggplot() +
  geom_hline(yintercept = 0, color = "steelblue4", linetype = 2) +
  geom_jitter(
    data = ctmin_gradient,
    aes(x = lat, y = ctmin, color = "CTmin"),
    width = 0.1, height = 0,
    size = 3, alpha = 0.5,
    show.legend = FALSE
  ) +
  geom_smooth(
    data = ctmin_gradient,
    aes(x = lat, y = ctmin, color = "CTmin"),
    method = "lm", fill = "forestgreen", alpha = 0.2,
    show.legend = TRUE,
    fullrange = TRUE
  ) +
  geom_line(
    data = bio6_df,
    aes(x = lat, y = mean_bio6, color = "BIO6"),
    linewidth = 1
  ) +
  geom_ribbon(
    data = bio6_df,
    aes(x = lat, ymin = min_bio6, ymax = max_bio6),
    fill = "steelblue", alpha = 0.2, show.legend = FALSE
  ) +
  scale_color_manual(
    name = NULL,
    values = c("CTmin" = "forestgreen", "BIO6" = "steelblue"),
    labels = c(
      "CTmin" = expression(CT[min]),
      "BIO6" = "Min. Temp."
    )
  ) +
  scale_x_continuous(
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(-10, 20, 5),
  ) +
  theme_classic() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.key.spacing.y = unit(0.5, "lines"),
    legend.text = element_text(size = 15),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 14)
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        linetype = c(1, 1),
        linewidth = c(2.5, 2.5),
        fill = c(NA, NA)
      )
    )
  ) +
  labs(
    x = "Latitude (°N)",
    y = "Temperature (°C)"
  )

```

Below, we perform a linear regression to test whether CT<sub>min</sub> and BIO6 scale with latitude. We find that while BIO6 decreases with latitude, CT<sub>min</sub> does not, indicating that green anoles are living in climates colder than their physiological cold tolerance.

```{r ctmin-does-not-scale-with-climate-stats}
# prepare data
bio6_df_model <- bio6_df |>
  select(lat, mean_bio6) |>
  rename(temp = mean_bio6) |>
  mutate(type = "bio6")
ctmin_model <- ctmin_gradient |>
  select(lat, ctmin) |>
  rename(temp = ctmin) |>
  mutate(type = "ctmin")
combined_df <- bind_rows(bio6_df_model, ctmin_model)

# fit linear model with interaction between latitude and type
lat_temp_model <- lm(temp ~ lat * type, data = combined_df)
summary(lat_temp_model)

broom::tidy(lat_temp_model)
broom::glance(lat_temp_model) |>
  pivot_longer(
    cols = everything(),
    names_to = "stat",
    values_to = "value"
  )

```

The interaction between latitude and type is positive (`{r} lat_temp_model$coefficients["lat:typectmin"]`), and significant (`{r} summary(lat_temp_model)$coefficients["lat:typectmin", "Pr(>|t|)"]`) indicating that CT<sub>min</sub> does not decrease with latitude as BIO6 does. This suggests that green anoles are living in climates colder than their physiological cold tolerance.

```{r table-ctmin-by-latitude}

# get average temps by location
bio6_df_table <- bio6_df |>
  filter(lat %in% round(unique(ctmin_gradient$lat), 1)) |>
  mutate(
    site = case_when(
      lat == 33.7 ~ "Atlanta, GA",
      lat == 35.9 ~ "Knoxville, TN",
      lat == 29.7 ~ "Gainesville, FL",
      lat == 31.5 ~ "Tifton, GA",
      lat == 27.3 ~ "Port Saint Lucie, FL",
      lat == 28.5 ~ "Orlando, FL",
      lat == 25.8 ~ "Miami, FL"
    )
  ) |>
  select(site, mean_bio6, min_bio6, max_bio6)


ctmin_gradient |>
  group_by(site, lat) |>
  summarise(
    mean_ctmin = mean(ctmin, na.rm = TRUE),
    sd_ctmin = sd(ctmin, na.rm = TRUE),
    n = n(),
  ) |>
  mutate(
    site = case_when(
      site == "ATL" ~ "Atlanta, GA",
      site == "KNX" ~ "Knoxville, TN",
      site == "GNV" ~ "Gainesville, FL",
      site == "TIF" ~ "Tifton, GA",
      site == "PSL" ~ "Port Saint Lucie, FL",
      site == "ORL" ~ "Orlando, FL",
      site == "MIA" ~ "Miami, FL"
    )
  ) |>
  ungroup() |>
  left_join(bio6_df_table, by = c("site")) |>
  mutate(
    across(
      c(mean_ctmin, sd_ctmin, mean_bio6, min_bio6, max_bio6),
      ~ round(.x, 2)
    )
  ) |>
  select(
    site, lat, n, mean_ctmin, sd_ctmin, mean_bio6, min_bio6, max_bio6
  ) |>
  kbl(
    col.names = c(
      "Site", "Latitude (°N)", "N", "Mean CT[min] (°C)",
      "SD CT[min] (°C)", "Mean BIO6 (°C)", "Min BIO6 (°C)", "Max BIO6 (°C)"
    ),
    escape = FALSE,
    align = "c"
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

## The green anole is an exceptional case of climatic overfilling

### Data processing

We compiled a dataset of CT<sub>min</sub> and elevational limits for 181 lizard species

```{r load-ctmin-lizards}
ctmin_lizards <- readRDS("data/ctmin_lizards.rds")
ctmin_lizards
```

Below we combine data on the CT<sub>min</sub> and elevational limits of globally distributed lizards to find the total area and proportion of each species' range that is below their CT<sub>min</sub> and below 0°C. We also calculate the maximum latitude at which each species can occur above their CT<sub>min</sub> and above 0°C.

```{r extract-data, eval = FALSE}
## Note: This code chunk is not evaluated as it takes a long time to run.
## The same code can be run locally to reproduce the results and the output
## is stored in the `data/lizard_climatic_overfilling.rds` file.

# read in global BIO6 tif
bio6 <- terra::rast("raw_data/bio_6.tif")
elev <- terra::rast("raw_data/elev.tif")
ranges <- readRDS("raw_data/ranges_vector.rds")

# filter ranges to only include lizard species with CTmin data
ctmin_lizards_ranges <- ranges[ranges$TaxonID %in% ctmin_lizards$gard_id, ]

# define function to conver lat-lon degree cell to km^2
deg_to_area <- function(lat, delta_lat = 2.5 / 60, delta_lon = 2.5 / 60) {
  
  # unit conversions
  lat_rad <- lat * pi / 180 # convert to radians

  # define constants
  equat_radius <- 6378.1370 # equatorial radius
  polar_radius <- 6356.7523142 # polar radius
  flat_factor <- 1 / 298.257223563 # flattening factor

  # radius curvature in the meridian
  e_sq <- (equat_radius^2 - polar_radius^2) / equat_radius^2
  denominator <- (1 - e_sq * sin(lat_rad)^2)^(1.5)
  m <- (equat_radius * (1 - e_sq)) / denominator

  # set length of 1 degree latitude in km
  km_per_degree_lat <- m * (pi / 180)
  lat_dist <- km_per_degree_lat * delta_lat

  # Calculate distance of 1 degree longitude along the parallel
  dist_km <- (pi / 180) * equat_radius * cos(lat_rad) / sqrt(1 - (2 * flat_factor - flat_factor^2) * sin(lat_rad)^2) # nolint
  lon_dist <- delta_lon * dist_km

  # Return cell area in km^2
  lat_dist * lon_dist
}

# define results list to extract data
results_list <- vector("list", nrow(ctmin_lizards))

for (i in seq_along(ctmin_lizards$gard_id)) {

  # get species range and ctmin
  range <- ctmin_lizards_ranges[ctmin_lizards_ranges$TaxonID == ctmin_lizards$gard_id[i], ] # nolint
  ctmin <- ctmin_lizards$ctmin[i]
  min_elev <- ctmin_lizards$min_elev[i]
  max_elev <- ctmin_lizards$max_elev[i]

  # crop BIO6 and elevation rasters to species range
  bio6_crop <- terra::crop(terra::mask(bio6, range), range)
  elev_crop <- terra::crop(terra::mask(elev, range), range)

  # filter all areas within range and within elevation limits
  valid_cells <- which(
    !is.na(bio6_crop[]) &
      !is.na(elev_crop[]) &
      elev_crop[] >= min_elev &
      elev_crop[] <= max_elev
  )

  # filter valid cells in BIo6 raster
  bio6_crop_filtered <- bio6_crop[valid_cells]
  bio6_crop_filtered <- bio6_crop_filtered$bio_6

  # extract coordinates of valid cells
  coords <- terra::xyFromCell(bio6_crop, valid_cells)[, "y"]
  dist <- deg_to_area(abs(coords))

  # get metrics
  total_area <- sum(dist)
  below_ctmin_area <- sum(dist[bio6_crop_filtered < ctmin])
  below_0_area <- sum(dist[bio6_crop_filtered < 0])
  below_ctmin_prop <- below_ctmin_area / total_area
  below_0_prop <- below_0_area / total_area
  max_lat_above_ctmin <- ifelse(
    below_ctmin_area %in% c(0, 1),
    NA,
    max(abs(coords[bio6_crop_filtered >= ctmin]), na.rm = TRUE)
  )
  max_lat_above_0 <- ifelse(
    below_0_area %in% c(0, 1),
    NA,
    max(abs(coords[bio6_crop_filtered >= 0]), na.rm = TRUE)
  )

  # store results in list
  results_list[[i]] <- data.frame(
    family = ctmin_lizards$family[i],
    genus = ctmin_lizards$genus[i],
    species = ctmin_lizards$species[i],
    ctmin = ctmin,
    substrate = ctmin_lizards$substrate_type[i],
    min_lat = min(abs(terra::ext(range)[3:4])),
    max_lat = max(abs(terra::ext(range)[3:4])),
    total_area = total_area,
    below_ctmin_area = below_ctmin_area,
    below_0_area = below_0_area,
    below_ctmin_prop = below_ctmin_prop,
    below_0_prop = below_0_prop,
    max_lat_above_ctmin = max_lat_above_ctmin,
    max_lat_above_0 = max_lat_above_0
  )

  # clear up memory
  rm(
    range, ctmin, min_elev, max_elev, bio6_crop, elev_crop,
    valid_cells, bio6_crop_filtered, coords, dist,
    total_area, below_ctmin_area, below_0_area, below_ctmin_prop,
    below_0_prop, max_lat_above_ctmin, max_lat_above_0
  )
  terra::tmpFiles(remove = TRUE)
  gc()

  # print progress
  cat("Processed species", i, "of", nrow(ctmin_lizards), "\n")

}

# bind rows for results and store data
lizard_climatic_overfilling <- bind_rows(results_list)

# change any NaN or -Inf values to NA across all columns
lizard_climatic_overfilling <- lizard_climatic_overfilling |>
  mutate(across(
    where(is.numeric),
    ~ ifelse(is.nan(.x) | is.infinite(.x), NA, .x)
  ))

# calculate global bio6 mean, min and max for every absolute lat degree
bio6_global <- as.data.frame(bio6, xy = TRUE, na.rm = TRUE) |>
  mutate(lat = round(abs(y), 0)) |>
  group_by(lat) |>
  summarise(
    mean_bio6 = mean(bio_6, na.rm = TRUE),
    min_bio6 = mean_bio6 - sd(bio_6, na.rm = TRUE),
    max_bio6 = mean_bio6 + sd(bio_6, na.rm = TRUE),
    .groups = "drop"
  )

saveRDS(lizard_climatic_overfilling, "data/lizard_climatic_overfilling.rds")
saveRDS(bio6_global, "data/bio6_global.rds")
```

### Statistical analysis

Below we estimate the Z-score of *Anolis carolinensis* range proportion below CT<sub>min</sub> and below 0 °C relative to the distribution of Z-scores for all lizard species in our dataset. 

```{r z-score-analysis}

# load datasets
lizard_climatic_overfilling <- readRDS("data/lizard_climatic_overfilling.rds")

# define function to get a Z-score for an observation relative to a group
get_z_score <- function(x, group) {
  (x - mean(group, na.rm = TRUE)) / sd(group, na.rm = TRUE)
}

# get Z-scores of A. carolinensis with respect to other groups
caro_z_scores <- expand_grid(
  var = c(
    "below_ctmin_prop",
    "below_0_prop",
    "below_ctmin_area",
    "below_0_area"
  ),
  group = c("anolis", "arboreal", "all")
) |>
  mutate(
    z_score = map2_dbl(var, group, \(v, g) {
      dat <- lizard_climatic_overfilling[[v]]
      comp <- switch(
        g,
        anolis = dat[lizard_climatic_overfilling$genus == "Anolis"],
        arboreal = dat[lizard_climatic_overfilling$substrate == "arboreal"],
        all = dat
      )
      x <- dat[lizard_climatic_overfilling$species == "Anolis carolinensis"]
      get_z_score(x = x, group = comp)
    })
  )

# tidy up for display as a table
caro_z_scores |>
  pivot_wider(names_from = group, values_from = z_score) |>
  mutate(
    across(
      c(anolis, arboreal, all),
      ~ round(.x, 2)
    )
  ) |>
  mutate(
    metric = case_when(
      var == "below_ctmin_prop" ~ "Proportion of range below CT[min]",
      var == "below_0_prop" ~ "Proportion of range below 0°C",
      var == "below_ctmin_area" ~ "Area of range below CT[min] (km^2)",
      var == "below_0_area" ~ "Area of range below 0°C (km^2)"
    )
  ) |>
  select(metric, anolis, arboreal, all) |>
  kbl(
    col.names = c(
      "Metric", "Z-score (relative to Anolis genus)",
      "Z-score (relative to arboreal lizards)",
      "Z-score (relative to all lizards)"
    ),
    escape = FALSE,
    align = "c"
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )

```

We find that *A. carolinensis* has a significantly higher proportion of its range below CT<sub>min</sub> and below 0°C than other lizard species, indicating that it is an exceptional case of climatic overfilling.

### Plots

First, we plot the latitudinal distribution of CT<sub>min</sub> and BIO6 for all lizard species in our dataset. We also highlight the latitudinal distribution of CT<sub>min</sub> for *A. carolinensis* and compare it to the latitudinal distribution of BIO6 across the globe.

```{r plot-lizard-climatic-overfilling-taxonomically}
#| fig-align: "center"
#| fig-width: 9
#| fig-height: 5

# load datasets
bio6_global <- readRDS("data/bio6_global.rds")
caro_ctmin_gradient <- readRDS("data/ctmin_gradient.rds")

# create data for A. carolinensis
caro_ctmin_pred <- tibble(
  lat = seq(22.75, 36.5, by = 0.1),
  ctmin = 6.8,
  substrate = "arboreal"
)

# filter taxonomic data for plotting
lizard_climatic_overfilling <- lizard_climatic_overfilling |>
  mutate(
    group = case_when(
      genus == "Anolis" & species != "Anolis carolinensis" ~ "Anolis Genus",
      species == "Anolis carolinensis" ~ "Anolis carolinensis",
      TRUE ~ "Other Lizards"
    )
  )

# plot lat ~ temp
lat_temps <- ggplot() +
  geom_hline(
    yintercept = 0, color = "darkblue", linewidth = 0.5, alpha = 0.5,
    linetype = "dashed"
  ) +
  geom_ribbon(
    data = bio6_global,
    aes(x = lat, ymin = min_bio6, ymax = max_bio6),
    fill = "steelblue", alpha = 0.2, show.legend = FALSE
  ) +
  geom_line(
    data = bio6_global,
    aes(x = lat, y = mean_bio6, color = "Global BIO6"),
    linewidth = 1
  ) +
  geom_segment(
    data = lizard_climatic_overfilling |> filter(group != "Anolis carolinensis"), # nolint
    aes(
      x = min_lat, xend = max_lat,
      y = ctmin, yend = ctmin,
      color = group
    ),
    linewidth = 1.25, alpha = 0.5
  ) +
  geom_line(
    data = caro_ctmin_pred,
    aes(x = lat, y = ctmin, color = "Anolis carolinensis"),
    linewidth = 1.5
  ) +
  scale_color_manual(
    name = NULL,
    breaks = c(
      "Anolis carolinensis",
      "Anolis Genus",
      "Other Lizards",
      "Global BIO6"
    ),
    values = c(
      "Anolis carolinensis" = "forestgreen",
      "Anolis Genus" = "green3",
      "Other Lizards" = "darkgray",
      "Global BIO6" = "steelblue"
    ),
    labels = c(
      "Anolis carolinensis" = expression(italic("A. carolinensis")),
      "Anolis Genus" = expression("Other " * italic("Anolis")),
      "Other Lizards" = "Other Lizards",
      "Global BIO6" = "Global Min. Temp."
    )
  ) +
  scale_y_continuous(
    breaks = seq(-10, 20, 5),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    breaks = seq(0, 70, 10),
    expand = c(0, 0)
  ) +
  coord_cartesian(
    xlim = c(0, max(lizard_climatic_overfilling$max_lat, na.rm = TRUE) + 1),
    ylim = c(
      -2, max(lizard_climatic_overfilling$ctmin, na.rm = TRUE) + 2
    )
  ) +
  theme_classic() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.background = element_rect(
      fill = "white", color = "gray", linewidth = 0.5
    ),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 12)
  ) +
  labs(
    x = "Absolute Latitude (°)",
    y = "Physiological cold tolerance (°C)"
  )

# plot range proportions/areas
range_data <- lizard_climatic_overfilling |>
  dplyr::select(
    group, below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area + 1, # add 1 to avoid log10(0) issues
    below_0_area = below_0_area + 1,
    group = factor(
      group,
      levels = c("Other Lizards", "Anolis Genus", "Anolis carolinensis")
    )
  ) |>
  arrange(group) |>
  pivot_longer(
    cols = c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area),
    names_to = "metric", values_to = "value"
  ) |>
  mutate(
    area_prop = case_when(
      metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
      metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
    ),
    temp = case_when(
      metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
      metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
    )
  ) |>
  arrange(group)

proportions <- range_data |>
  filter(area_prop == "Proportion") |>
  ggplot(
    aes(x = temp, y = value, color = group, alpha = group, size = group)
  ) +
  geom_jitter(width = 0.1, height = 0, show.legend = FALSE) +
  facet_wrap(
    ~area_prop,
    scales = "free",
    strip.position = "left",
    labeller = as_labeller(
      c(Proportion = "Proportion~of~range"),
      default = label_parsed
    )
  ) +
  scale_y_continuous() +
  scale_color_manual(
    values = c(
      "Anolis carolinensis" = "darkgreen",
      "Anolis Genus" = "palegreen3",
      "Other Lizards" = "gray"
    ),
    name = "Group"
  ) +
  scale_alpha_manual(
    values = c(0.5, 0.75, 1),
    name = "Group"
  ) +
  scale_size_manual(
    values = c(2, 2.5, 4),
  ) +
  scale_x_discrete(
    labels = c("CT[min]" = expression("< " * "CT"[min]),
    "0°C" = "< 0°C")
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90, margin = margin(r = 0)),
    strip.switch.pad.grid = grid::unit(0, "pt"),
    strip.text = element_text(size = 12)
  )

areas <- range_data |>
  filter(area_prop == "Area") |>
  ggplot(
    aes(x = temp, y = value, color = group, alpha = group, size = group)
  ) +
  geom_jitter(width = 0.1, height = 0, show.legend = FALSE) +
  facet_wrap(
    ~area_prop,
    scales = "free",
    strip.position = "left",
    labeller = as_labeller(
      c(Area = "Area~of~range~(km^2)"),
      default = label_parsed
    )
  ) +
  scale_y_continuous(
    trans = "log10", 
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000)
  ) +
  scale_color_manual(
    values = c(
      "Anolis carolinensis" = "darkgreen",
      "Anolis Genus" = "palegreen3",
      "Other Lizards" = "gray"
    ),
    name = "Group"
  ) +
  scale_alpha_manual(
    values = c(0.5, 0.75, 1),
    name = "Group"
  ) +
  scale_size_manual(
    values = c(2, 2.5, 4),
  ) +
  scale_x_discrete(
    labels = c(
      "CT[min]" = expression("< " * "CT"[min]),
      "0°C" = "< 0°C"
    )
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.ticks = element_line(color = "black", linewidth = 0.25),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90, margin = margin(r = 0)),
    strip.switch.pad.grid = grid::unit(0, "pt"),
    strip.text = element_text(size = 12)
  )

combined_plot <- ggarrange(
  lat_temps, proportions, areas,
  ncol = 3, nrow = 1,
  widths = c(1.25, 0.5, 0.5),
  labels = c("A", "B", "C"),
  font.label = list(size = 15, face = "bold", color = "black")
)
combined_plot

```

Next we plot by substrate type to see if there are any differences in the latitudinal distribution of CT<sub>min</sub> and BIO6 across different ecological groups of lizards. We also highlight the latitudinal distribution of CT<sub>min</sub> for *A. carolinensis* and compare it to the latitudinal distribution of BIO6 across the globe.

```{r plot-by-substrate}
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5

# get substrate data and colors
substrate_data <- lizard_climatic_overfilling |>
  mutate(
    substrate = replace_na(as.character(substrate), "Unknown"),
    substrate = fct_infreq(substrate)
  )

substrate_levels <- levels(substrate_data$substrate)
substrate_colors <- setNames(
  c("peachpuff3", "khaki3", "springgreen3"), substrate_levels
)

lat_temps_substrate <- ggplot() +
  geom_hline(
    yintercept = 0, color = "darkblue", linewidth = 0.5, alpha = 0.5,
    linetype = "dashed"
  ) +
  geom_ribbon(
    data = bio6_global,
    aes(x = lat, ymin = min_bio6, ymax = max_bio6),
    fill = "steelblue", alpha = 0.2, show.legend = FALSE
  ) +
  geom_line(
    data = bio6_global,
    aes(x = lat, y = mean_bio6, color = "Global BIO6"), linewidth = 1
  ) +
  geom_segment(
    data = substrate_data |>
      filter(species != "Anolis carolinensis"),
    aes(
      x = min_lat, xend = max_lat, y = ctmin, yend = ctmin,
      color = substrate
    ),
    linewidth = 1.25, alpha = 0.5
  ) +
  geom_line(
    data = caro_ctmin_pred,
    aes(x = lat, y = ctmin, color = "Anolis carolinensis"),
    linewidth = 1.5
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      substrate_colors,
      "Anolis carolinensis" = "forestgreen",
      "Global BIO6" = "steelblue"
    ),
    breaks = c(substrate_levels, "Anolis carolinensis", "Global BIO6"),
    labels = c(
      as.list(setNames(substrate_levels, substrate_levels)),
      list(
        "other" = "Not Arboreal Lizards",
        "semi-arboreal" = "Semi-Arboreal Lizards",
        "arboreal" = "Arboreal Lizards",
        "Anolis carolinensis" = expression(italic("A. carolinensis")),
        "Global BIO6" = "Global Min. Temp."
      )
    )
  ) +
  scale_y_continuous(
    breaks = seq(-10, 20, 5), labels = paste0(seq(-10, 20, 5), " °C"),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    breaks = seq(0, 70, 10), labels = paste0(seq(0, 70, 10), "°"),
    expand = c(0, 0)
  ) +
  coord_cartesian(
    xlim = c(0, max(substrate_data$max_lat, na.rm = TRUE) + 1),
    ylim = c(-2, max(substrate_data$ctmin, na.rm = TRUE) + 2)
  ) +
  theme_classic() +
  theme(
    legend.position = "inside", legend.position.inside = c(1, 1),
    legend.justification = c(1, 1),
    legend.background = element_rect(
      fill = "white", color = "gray", linewidth = 0.5
    ),
    axis.title.y = element_blank(), axis.text = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_blank()
  ) +
  labs(x = "Absolute Latitude")

make_substrate_plot <- function(data, metric_type, y_label) {
  data |>
    filter(area_prop == metric_type) |>
    ggplot(aes(x = temp, y = value, color = substrate)) +
    geom_point(
      position = position_jitterdodge(
        dodge.width = 1,
        jitter.width = 0.2,
        jitter.height = 0
      ),
      show.legend = FALSE,
      alpha = 0.25
    ) +
    stat_summary(position = position_dodge(width = 1)) +
    geom_point(
      data = substrate_data |>
        filter(species == "Anolis carolinensis") |>
        mutate(
          below_ctmin_area = below_ctmin_area + 1,
          below_0_area = below_0_area + 1,
          substrate = factor(substrate, levels = substrate_levels)
        ) |>
        pivot_longer(
          cols = c(below_ctmin_prop, below_0_prop,
            below_ctmin_area, below_0_area),
          names_to = "metric",
          values_to = "value"
        ) |>
        mutate(
          area_prop = case_when(
            metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
            metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
          ),
          temp = case_when(
            metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
            metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
          )
        ) |>
        filter(area_prop == metric_type),
      aes(x = temp, y = value, group = substrate),
      position = position_nudge(x = 1 / 3),
      color = "forestgreen",
      size = 4,
      show.legend = FALSE
    ) +
    geom_vline(xintercept = 1.5, color = "gray70", linewidth = 0.2) +
    scale_color_manual(values = substrate_colors) +
    scale_x_discrete(
      labels = c(
        "CT[min]" = expression("< " * "CT"[min]),
        "0°C" = "< 0°C"
      )
    ) +
    theme_minimal() +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
      axis.ticks = element_line(color = "black", linewidth = 0.25),
      axis.text.x = element_text(size = 12, color = "black"),
      axis.text.y = element_text(color = "black"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 12, color = "black"),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      strip.text = element_blank()
    ) +
    labs(
      y = y_label,
      subtitle = NULL
    )
}

proportions_substrate <- substrate_data |>
  select(
    substrate, below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area + 1,
    below_0_area = below_0_area + 1,
    substrate = factor(substrate, levels = substrate_levels)
  ) |>
  pivot_longer(
    cols = c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    area_prop = case_when(
      metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
      metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
    ),
    temp = case_when(
      metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
      metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
    )
  ) |>
  make_substrate_plot("Proportion", "Proportion of range") +
  scale_y_continuous()

areas_substrate <- substrate_data |>
  select(
    substrate, below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area
  ) |>
  mutate(
    below_ctmin_area = below_ctmin_area + 1,
    below_0_area = below_0_area + 1,
    substrate = factor(substrate, levels = substrate_levels)
  ) |>
  pivot_longer(
    cols = c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    area_prop = case_when(
      metric %in% c("below_ctmin_prop", "below_0_prop") ~ "Proportion",
      metric %in% c("below_ctmin_area", "below_0_area") ~ "Area"
    ),
    temp = case_when(
      metric %in% c("below_ctmin_prop", "below_ctmin_area") ~ "CT[min]",
      metric %in% c("below_0_prop", "below_0_area") ~ "0°C"
    )
  ) |>
  make_substrate_plot("Area", "Area of range (km^2)") +
  scale_y_continuous(
    trans = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000)
  )

ggarrange(
  lat_temps_substrate, proportions_substrate, areas_substrate,
  ncol = 3, nrow = 1,
  widths = c(1.25, 0.5, 0.5),
  labels = c("A", "B", "C"),
  align = "hv",
  font.label = list(size = 15, face = "bold", color = "black")
)
```

## Tables

We table the average, minimum and maximum CT<sub>min</sub> and BIO6 values for each 0.1° latitude band across the range of the green anole.

Table "overfilling" results by family and genus

```{r lizard-climatic-overfilling-table-family-genus}

lizard_climatic_overfilling |>
  mutate(
    genus = gsub(" .*", "", species),
    species = gsub(".* ", "", species)
  ) |>
  mutate(
    below_ctmin_prop = round(below_ctmin_prop, 2),
    below_0_prop = round(below_0_prop, 2),
    below_ctmin_area = round(below_ctmin_area / 1e6, 4),
    below_0_area = round(below_0_area / 1e6, 4),
    total_area = round(total_area / 1e6, 4)
  ) |>
  group_by(family, genus) |>
  summarise(
    below_ctmin_prop = mean(below_ctmin_prop, na.rm = TRUE),
    below_0_prop = mean(below_0_prop, na.rm = TRUE),
    below_ctmin_area = mean(below_ctmin_area, na.rm = TRUE),
    below_0_area = mean(below_0_area, na.rm = TRUE),
    total_area = mean(total_area, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # round everything to 3 decimal digits
  mutate(
    across(
      c(below_ctmin_prop, below_0_prop, below_ctmin_area, below_0_area, total_area),
      ~ round(.x, 3)
    )
  ) |>
  arrange(family, genus) |>
  kable(
    col.names = c(
      "Family", "Genus",
      "Proportion < CTmin",
      "Proportion < 0°C", "Area < CTmin (10⁶ km²)",
      "Area < 0°C (10⁶ km²)", "Total area (10⁶ km²)"
    )
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
  


```

Last we table to "overfilling" results for all species

```{r lizard-climatic-overfilling-table}

lizard_climatic_overfilling |>
  mutate(
    genus = gsub(" .*", "", species),
    species = gsub(".* ", "", species)
  ) |>
  mutate(
    below_ctmin_prop = round(below_ctmin_prop, 2),
    below_0_prop = round(below_0_prop, 2),
    below_ctmin_area = round(below_ctmin_area / 1e6, 4),
    below_0_area = round(below_0_area / 1e6, 4),
    total_area = round(total_area / 1e6, 4)
  ) |>
  select(
    family, genus, species,
    below_ctmin_prop, below_0_prop,
    below_ctmin_area, below_0_area, total_area
  ) |>
  arrange(genus, species) |>
  kable(
    col.names = c(
      "Family", "Genus", "Species",
      "Proportion < CTmin",
      "Proportion < 0°C", "Area < CTmin (10⁶ km²)",
      "Area < 0°C (10⁶ km²)", "Total area (10⁶ km²)"
    )
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```