---
title: "Mismatch between realized and fundamental cold thermal niches in lizards"
format:
  html:
    code-fold: true
    fig-dpi: 300
---

```{r r-setup, warning = FALSE, message = FALSE}
# packages
library(tidyverse)
library(NicheMapR)
library(kableExtra)
library(terra)
library(sf)
library(geodata)
library(elevatr)
library(ggpubr)

# options
options(
  dplyr.summarise.inform = FALSE,
  knitr.table.html.attr = "quarto-disable-processing=true"
)
```

## Overview



## Acquire Data

The first step is to acquire the data needed for the analysis. This includes:

- Geographic ranges of lizards which we obtain from the GARD database.

- Global minimum annual temperature data which we obtain from WorldClim.

- Global elevation data which we obtain from the elevatr package.

```{r obtain-data, eval = FALSE}

# read in lizard range data
range_shapefile <- st_read(
  "raw_data/lizard_ranges/doi_10_5061_dryad_9cnp5hqmb__v20220427"
)
ranges_vector <- terra::vect(range_shapefile)

# get WorldClim minimum annual temperature data
global_min_temp <- geodata::worldclim_global(
  var = "tmin",
  res = 2.5,
  path = "raw_data"
)

# find minimum value across all layers of the raster for each cell
global_min_temp <- global_min_temp |> terra::app(fun = min, na.rm = TRUE)

# try WorldClim elevation first; fall back to elevatr if server is unavailable
global_elev <- elevatr::get_elev_raster(
  locations = global_min_temp,
  z = 3,
  prj = terra::crs(global_min_temp),
  clip = "locations"
)
global_elev <- terra::rast(global_elev)
global_elev <- terra::crop(global_elev, global_min_temp)
global_elev <- terra::project(global_elev, global_min_temp)
global_elev <- terra::mask(global_elev, global_min_temp)

# add elevation layer to global_min_temp raster
global_min_temp <- c(global_min_temp, global_elev)

# save datasets
saveRDS(global_min_temp, "data/global_min_temp.rds")
saveRDS(ranges_vector, "raw_data/ranges_vector.rds")

```

In addition to the above, we have also obtained data on the CT<sub>min</sub> from:

- The GlobTherm database.

- Bodensteiner et al. 2021 (https://doi.org/10.1111/1365-2435.13853)

- Salazar et al. 2022 (https://doi.org/10.1111/ele.14092)

With the latter two sources being specific to *Anolis* lizards. Further, we also obtain data on the minimum and maximum elevation at which each species occurs from the same two papers and from SquamBase (Meiri 2020). The structure of the resulting dataset is shown below:

```{r show-ctmin-data}
ctmin_data <- readRDS("data/ctmin_data.rds")
as_tibble(ctmin_data)
```

We obtained data for a total of `{r} unique(ctmin_data$species) |> length()` lizard species. Out of these, `{r} sum(!is.na(ctmin_data$min_elev) & !is.na(ctmin_data$max_elev))` species have clearly defined elevational ranges.

## Analysis

Next, for each species we will extract the minimum annual temperature data across its geographic range and determine:

- The proportion and area of the species range that is below its CT<sub>min</sub>.

- The maximum absolute latitude within the species range at which the minimum annual temperature is greater than or equal to its CT<sub>min</sub>.

```{r get-ctmin-results, eval = FALSE}

# read in CTmin data
global_min_temp <- readRDS("data/global_min_temp.rds")
ranges_vector <- readRDS("raw_data/ranges_vector.rds")

# filter CTmin data to only species within the range shapefile
ctmin_data <- ctmin_data |>
  filter(species %in% ranges_vector$binomial) |>
  filter(!is.na(min_elev) & !is.na(max_elev))

# define function to convert latitude/longitude degrees to cell area in km^2
deg_to_area <- function(lat_deg, delta_lat = 0.0416, delta_long = 0.0416) {

  # Convert latitude from degrees to radians
  lat_rad <- lat_deg * pi / 180

  # WGS84 ellipsoid constants
  a <- 6378.1370      # Semi-major axis (equatorial radius in km)
  b <- 6356.7523142   # Semi-minor axis (polar radius in km)

  # Calculate the radius of curvature in the meridian (North-South)
  e_sq <- (a^2 - b^2) / a^2
  denominator <- (1 - e_sq * sin(lat_rad)^2)^(1.5)
  m <- (a * (1 - e_sq)) / denominator

  # Length of 1 degree of latitude in km at this specific latitude
  km_per_degree_lat <- m * (pi / 180)
  lat_dist <- km_per_degree_lat * delta_lat

  # WGS84 Earth semi-major axis in km
  a_km <- 6378.137

  # Flattening factor
  f <- 1 / 298.257223563

  # Calculate distance of 1 degree longitude along the parallel
  dist_km <- (pi / 180) * a_km * cos(lat_rad) / sqrt(1 - (2 * f - f^2) * sin(lat_rad)^2) # nolint
  lon_dist <- delta_long * dist_km

  # Return cell area in km^2
  lat_dist * lon_dist
}

# loop to extract data
results_list <- vector("list", nrow(ctmin_data))

# holder list for latitudinal change in min temperature across species range
lat_temp_change_list <- list()

for (i in seq_along(ctmin_data$species)) {

  tryCatch({

    # filter data for species of interest
    species_name <- ctmin_data$species[i]
    species_range <- ranges_vector[ranges_vector$binomial == species_name, ]
    species_ctmin <- ctmin_data$ctmin[ctmin_data$species == species_name][1]

    # get minimum and maximum latitude within range
    min_lat <- min(abs(terra::ext(species_range)[3:4]))
    max_lat <- max(abs(terra::ext(species_range)[3:4]))

    # get proportion of range below CTmin
    species_min_temp <- terra::extract(
      global_min_temp, species_range, xy = TRUE
    )
    colnames(species_min_temp) <- c("id", "tmin", "elev", "x", "y")

    # filter elevations outside of range
    species_min_temp <- species_min_temp |>
      filter(elev >= ctmin_data$min_elev[i] & elev <= ctmin_data$max_elev[i])

    # calculate area of each cell based on latitude and that all cells re
    species_min_temp$area_km2 <- deg_to_area(species_min_temp$y)

    # get proportion of cells below CTmin
    props <- species_min_temp |>
      summarise(
        total_area = sum(area_km2, na.rm = TRUE),
        area_below_ctmin = sum(area_km2[tmin < species_ctmin], na.rm = TRUE),
        area_below_0 = sum(area_km2[tmin < 0], na.rm = TRUE)
      ) |>
      mutate(
        prop_below_ctmin = area_below_ctmin / total_area,
        prop_below_0 = area_below_0 / total_area
      )
    prop_below_ctmin <- props |> pull(prop_below_ctmin)
    prop_below_0 <- props |> pull(prop_below_0)
    area_below_ctmin <- props |> pull(area_below_ctmin)
    area_below_0 <- props |> pull(area_below_0)
    total_area <- props |> pull(total_area)

    # get maximum absolute latitude within range at which Tmin is >= CTmin
    above_ctmin <- species_min_temp |> filter(tmin >= species_ctmin)
    max_lat_above_ctmin <- ifelse(nrow(above_ctmin) == 0, NA, max(abs(above_ctmin$y), na.rm = TRUE)) # nolint

    # get maximum absolute latitude within range at which Tmin is >= 0
    above_0 <- species_min_temp |> filter(tmin >= 0)
    max_lat_above_0 <- if (nrow(above_0) == 0) {
      NA_real_
    } else {
      max(abs(above_0$y), na.rm = TRUE)
    }

    # store results in pre-allocated list
    results_list[[i]] <- tibble(
      species = species_name,
      ctmin = species_ctmin,
      min_lat = min_lat,
      max_lat = max_lat,
      prop_below_ctmin = prop_below_ctmin,
      prop_below_0 = prop_below_0,
      area_below_ctmin = area_below_ctmin,
      area_below_0 = area_below_0,
      total_area = total_area,
      max_lat_above_ctmin = max_lat_above_ctmin,
      max_lat_above_0 = max_lat_above_0
    )

    # get latitudinal change in min temperature across species range
    lat_temp_change_list[[i]] <- species_min_temp |>
      mutate(abs_lat = abs(round(y, 1))) |>
      group_by(abs_lat) |>
      summarise(
        mean_min_temp = mean(tmin, na.rm = TRUE),
        max_min_temp = max(tmin, na.rm = TRUE),
        min_min_temp = min(tmin, na.rm = TRUE)
      ) |>
      ungroup() |>
      mutate(species = species_name)

  }, error = function(e) {
    message(
      "Error on species ", i, " (", ctmin_data$species[i], "): ", e$message
    )
  })

  # clear up memory
  rm(species_range, species_min_temp, above_ctmin)
  terra::tmpFiles(remove = TRUE)
  gc()

  print(paste0("Processed species ", i, " of ", nrow(ctmin_data)))

}

# bind rows in results list
results <- bind_rows(results_list)

# bind all species latitudinal temperature change data into a single data frame
lat_temp_change_df <- bind_rows(lat_temp_change_list)

# save datasets
saveRDS(results, "data/ctmin_results.rds")
saveRDS(lat_temp_change_df, "data/lat_temp_change_df.rds")

```

## Figures

Below we present the results of the analysis through 3 plots:

Global minimum annual temperature as a function of absolute latitude, with the CT<sub>min</sub> of each species overlaid. The dashed line represents 0°C, the solid blue line represents the mean minimum annual temperature across all species ranges, and the shaded area represents the range of minimum annual temperatures across all species ranges.

```{r plot-min-temp-by-range-ctmin}
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 5
#| out-width: 70%

lat_temp_change_df <- readRDS("data/lat_temp_change_df.rds")
results <- readRDS("data/ctmin_results.rds")
results <- results |> distinct(species, .keep_all = TRUE)

lat_temp_change_df_sum <- lat_temp_change_df |>
  mutate(abs_lat = round(abs_lat)) |>
  group_by(abs_lat) |>
  summarise(
    mean_min_temp = mean(mean_min_temp, na.rm = TRUE),
    max_min_temp = mean(max_min_temp, na.rm = TRUE),
    min_min_temp = mean(min_min_temp, na.rm = TRUE)
  ) |>
  ungroup()

ranges <- ggplot() +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray"
  ) +
  geom_line(
    data = lat_temp_change_df_sum,
    aes(x = abs_lat, y = mean_min_temp),
    color = "darkblue"
  ) +
  geom_ribbon(
    data = lat_temp_change_df_sum,
    aes(x = abs_lat, ymin = min_min_temp, ymax = max_min_temp),
    alpha = 0.2,
    fill = "steelblue"
  ) +
  geom_segment(
    data = results |> filter(!grepl("Anolis", species)),
    aes(
      x = min_lat,
      xend = max_lat,
      y = ctmin,
      yend = ctmin,
      color = "Other species"
    ),
    alpha = 0.5,
    linewidth = 1
  ) +
  geom_segment(
    data = results |>
      filter(grepl("Anolis", species) & species != "Anolis carolinensis"),
    aes(
      x = min_lat,
      xend = max_lat,
      y = ctmin,
      yend = ctmin,
      color = "Anolis genus"
    ),
    alpha = 0.75,
    linewidth = 1.5
  ) +
  geom_segment(
    data = results |> filter(species == "Anolis carolinensis"),
    aes(
      x = min_lat,
      xend = max_lat,
      y = ctmin,
      yend = ctmin,
      color = "Anolis carolinensis"
    ),
    linewidth = 2
  ) +
  scale_color_manual(
    name = "Species",
    values = c(
      "Other species" = "gray",
      "Anolis genus" = "palegreen3",
      "Anolis carolinensis" = "darkgreen"
    ),
    labels = c(
      "Other species" = "Other species",
      "Anolis genus" = expression(italic("Anolis") ~ "genus"),
      "Anolis carolinensis" = expression(italic("Anolis carolinensis"))
    )
  ) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    labels = paste(seq(0, 60, by = 10), "°", sep = ""),
    expand = c(0, 0)
  ) +
  coord_cartesian(
    ylim = c(-5, 22),
    xlim = c(0, 60)
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(r = 10, t = 10, b = 10, l = 10),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.9),
    legend.background = element_rect(fill = "white", color = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  labs(
    x = "Absolute Latitude",
    y = expression("Minimum Annual Temperature &" ~ CT[min] ~ "(°C)")
  )

proportions_and_areas <- results |>
  mutate(
    group = case_when(
      grepl("Anolis", species) & species != "Anolis carolinensis" ~ "Anolis genus", # nolint
      species == "Anolis carolinensis" ~ "Anolis carolinensis",
      TRUE ~ "Other species"
    )
  ) |>
  mutate(
    group = factor(
      group,
      levels = c("Other species", "Anolis genus", "Anolis carolinensis")
    ),
    log_area_below_ctmin = log(area_below_ctmin + 1),
    log_area_below_0 = log(area_below_0 + 1)
  ) |>
  select(
    group,
    prop_below_ctmin, prop_below_0,
    log_area_below_ctmin, log_area_below_0
  ) |>
  pivot_longer(
    cols = c(
      prop_below_ctmin, log_area_below_ctmin, log_area_below_0, prop_below_0
    ),
    names_to = "metric",
    values_to = "value"
  ) |>
  arrange(group) |>
  ggplot(aes(x = metric, y = value, color = group, alpha = group, size = group)) +
  geom_jitter(
    width = 0.2
  ) +
  facet_wrap(
    ~metric,
    scales = "free",
    strip.position = "left",
    labeller = as_labeller(
      c(
        prop_below_ctmin = "Proportion~`<`~CT[min]",
        log_area_below_ctmin = "Area~`<`~CT[min]~(log~km^2~`+`~1)",
        log_area_below_0 = "Area~`<`~0*degree*C~(log~km^2~`+`~1)",
        prop_below_0 = "Proportion~`<`~0*degree*C"
      ),
      default = label_parsed
    )
  ) +
  scale_color_manual(
    values = c(
      "Anolis carolinensis" = "darkgreen",
      "Anolis genus" = "palegreen3",
      "Other species" = "gray"
    ),
    name = "Group"
  ) +
  scale_alpha_manual(
    values = c(0.5, 0.75, 1),
    name = "Group"
  ) +
  scale_size_manual(
    values = c(2, 2.5, 4),
    name = "Group"
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.ticks.y = element_line(color = "black", linewidth = 0.25),
    axis.text.x = element_blank(),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90, margin = margin(r = 0)),
    strip.switch.pad.grid = grid::unit(0, "pt"),
    strip.text = element_text(size = 12)
  )

combined_plot <- ggarrange(
  ranges, proportions_and_areas,
  ncol = 2, nrow = 1,
  widths = c(1.5, 1),
  labels = c("A", "B"),
  align = "hv",
  font.label = list(size = 14, face = "bold")
)

combined_plot
```

Relationship between the maximum absolute latitude of the species range and the maximum absolute latitude at which the minimum annual temperature is greater than or equal to the CT<sub>min</sub> for each species. The dashed line represents a 1:1 relationship.

```{r max-lat-vs-max-lat-above-ctmin}
#| fig-align: "center"
#| fig-height: 5
#| fig-width: 5
#| out-width: 70%
results |>
  mutate(
    group = case_when(
      grepl("Anolis", species) & species != "Anolis carolinensis" ~ "Anolis genus", # nolint
      species == "Anolis carolinensis" ~ "Anolis carolinensis",
      TRUE ~ "Other species"
    )
  ) |>
  select(species, max_lat, max_lat_above_ctmin, group) |>
  distinct() |>
  ggplot(aes(x = max_lat, y = max_lat_above_ctmin)) +
  geom_point(
    aes(color = group, alpha = group),
    size = 4,
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_manual(
    values = c("darkred", "darkorange", "gray"),
    labels = c(
      "Other species" = "Other species",
      "Anolis genus" = expression(italic("Anolis") ~ "genus"),
      "Anolis carolinensis" = expression(italic("Anolis carolinensis"))
    )
  ) +
  scale_alpha_manual(
    values = c(1, 0.75, 0.5),
    labels = c(
      "Other species" = "Other species",
      "Anolis genus" = expression(italic("Anolis") ~ "genus"),
      "Anolis carolinensis" = expression(italic("Anolis carolinensis"))
    )
  ) +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.25, 0.85)
  ) +
  scale_x_continuous(
    breaks = seq(0, 90, 10),
    labels = paste(seq(0, 90, 10), "°", sep = "")
  ) +
  scale_y_continuous(
    breaks = seq(0, 90, 10),
    labels = paste(seq(0, 90, 10), "°", sep = "")
  ) +
  labs(
    x = "Maximum latitude of range",
    y = expression("Maximum latitude where Min. Temp.">=~CT[min]),
  ) +
  guides(
    size = guide_legend(override.aes = list(size = 4))
  ) +
  coord_cartesian(xlim = c(8, 60), ylim = c(8, 60))
```

Below we plot the proportion of the range and the log of the area in km<sup>2</sup> of each species' range that falls below CT<sub>min<sub>.

```{r prop-below-ctmin-vs-species}
#| fig-align: "center"
#| fig-height: 20
#| fig-width: 12
#| out-width: 70%

# proportion of range plot
proportion_plot <- results |>
  mutate(
    group = case_when(
      grepl("Anolis", species) & species != "Anolis carolinensis" ~ "Anolis genus", # nolint
      species == "Anolis carolinensis" ~ "Anolis carolinensis",
      TRUE ~ "Other species"
    ),
  ) |>
  select(species, prop_below_ctmin, area_below_ctmin, group) |>
  filter(!is.na(prop_below_ctmin) & !is.na(area_below_ctmin)) |>
  distinct() |>
  ggplot(aes(y = reorder(species, prop_below_ctmin), x = prop_below_ctmin)) +
  geom_bar(
    aes(fill = group),
    stat = "identity",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    expand = c(0, 0)
  ) +
  labs(
    x = expression("Proportion of range below" ~ CT[min]),
    y = "Species"
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(r = 10, t = 10, b = 10, l = 10),
    axis.text.y = element_text(size = 8, face = "italic"),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 16)
  ) +
  scale_fill_manual(
    values = c(
      "Anolis carolinensis" = "forestgreen",
      "Anolis genus" = "darkorange",
      "Other species" = "gray"
    ),
    name = "Group"
  )

# area plot
area_plot <- results |>
  mutate(
    group = case_when(
      grepl("Anolis", species) & species != "Anolis carolinensis" ~ "Anolis genus", # nolint
      species == "Anolis carolinensis" ~ "Anolis carolinensis",
      TRUE ~ "Other species"
    ),
  ) |>
  select(species, prop_below_ctmin, area_below_ctmin, group) |>
  filter(!is.na(prop_below_ctmin) & !is.na(area_below_ctmin)) |>
  distinct() |>
  ggplot(
    aes(
      y = reorder(species, area_below_ctmin),
      x = log(area_below_ctmin + 1)
    )
  ) +
  geom_bar(
    aes(fill = group),
    stat = "identity",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = scales::comma_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  labs(
    x = expression("Area of range below" ~ CT[min] ~ "(" ~ log ~ km^2* ~ " + 1)"), # nolint
    y = "Species"
  ) +
  theme_classic() +
  theme(
    plot.margin = margin(r = 10, t = 10, b = 10, l = 10),
    axis.text.y = element_text(size = 8, face = "italic"),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 16)
  ) +
  scale_fill_manual(
    values = c(
      "Anolis carolinensis" = "forestgreen",
      "Anolis genus" = "darkorange",
      "Other species" = "gray"
    ),
    name = "Group"
  )

# combine plots
ggarrange(
  proportion_plot,
  area_plot,
  ncol = 2,
  nrow = 1,
  align = "v",
  labels = c("A", "B"),
  font.label = list(size = 14, face = "bold")
)

```

## Tables

:::{.callout-note collapse="true"}

## Proportion and Area of Range Below Temperature Thresholds by Genus

```{r table ctmin-results-by-genus}
results |>
  mutate(
    genus = gsub(" .*", "", species),
  ) |>
  select(
    genus, species,
    prop_below_ctmin,
    prop_below_0, area_below_ctmin, area_below_0, total_area
  ) |>
  group_by(genus) |>
  summarise(
    n_species = n(),
    mean_prop_below_ctmin = round(mean(prop_below_ctmin, na.rm = TRUE), 2),
    sd_prop_below_ctmin = round(sd(prop_below_ctmin, na.rm = TRUE), 2),
    mean_prop_below_0 = round(mean(prop_below_0, na.rm = TRUE), 2),
    sd_prop_below_0 = round(sd(prop_below_0, na.rm = TRUE), 2),
    mean_area_below_ctmin = round(mean(area_below_ctmin, na.rm = TRUE) / 1e5, 2), # nolint
    mean_area_below_0 = round(mean(area_below_0, na.rm = TRUE) / 1e5, 2),
    mean_total_area = round(mean(total_area, na.rm = TRUE) / 1e5, 2)
  ) |>
  ungroup() |>
  mutate(
    prop_below_ctmin = paste0(mean_prop_below_ctmin, " ± ", sd_prop_below_ctmin),
    prop_below_0 = paste0(mean_prop_below_0, " ± ", sd_prop_below_0)
  ) |>
  select(genus, n_species,prop_below_ctmin, prop_below_0) |>
  kable(
    col.names = c(
      "Genus", "N",
      "Proportion of range < CTmin", "Proportion of range < 0°C"
    )
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
:::

:::{.callout-note collapse="true"}

## Proportion and Area of Range Below Temperature Thresholds by Species

```{r table ctmin-results-by-species}
results |>
mutate(
  genus = gsub(" .*", "", species),
  species = gsub(".* ", "", species)
) |>
select(
  genus, species,
  ctmin, min_lat, max_lat,
  prop_below_ctmin,
  prop_below_0, area_below_ctmin, area_below_0, total_area
) |>
mutate(
  lat_range = paste0(round(min_lat, 1), "° - ", round(max_lat, 1), "°"),
  prop_below_ctmin = round(prop_below_ctmin, 2),
  prop_below_0 = round(prop_below_0, 2),
  area_below_ctmin = round(area_below_ctmin / 1e6, 4),
  area_below_0 = round(area_below_0 / 1e6, 4),
  total_area = round(total_area / 1e6, 4)
) |>
select(
  genus, species, lat_range,
  prop_below_ctmin, prop_below_0,
  area_below_ctmin, area_below_0, total_area
) |>
arrange(genus, species) |>
kable(
  col.names = c(
    "Genus", "Species", "Lat. Range",
     "Proportion < CTmin",
    "Proportion < 0°C", "Area < CTmin (10⁶ km²)",
    "Area < 0°C (10⁶ km²)", "Total area (10⁶ km²)"
  )
) |>
kable_styling(
  full_width = FALSE,
  position = "center"
)

```

:::